---
title: "Analysis of single cell neuroblastoma organoids: part 2 - test different resolutions"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    # css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---
```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)

library(here)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).
The objective of this R Markdown code is to analyse the results obtained by the single cell RNA sequencing.

```{r libraries}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

packages <- c(
  "Seurat",
  "ggplot2",
  "patchwork",
  "dplyr",
  "here",
  "clustree",
  "scCustomize",
  "ggforce",
  "concaveman",
  "openxlsx"
)

invisible(lapply(packages, library, character.only = TRUE))
```


Folder structure
```{r}
on_codeocean <- dir.exists("/data") && dir.exists("/code")
if (on_codeocean) {
  Datafolder  <- "/data/Data_organoid"
  utils_file  <- "/code/Rscripts/utils.R"
  config_file <- "/code/Rscripts/config.R"
} else {
  Datafolder  <- here("data/Data_organoid")
  utils_file  <- here("code/Rscripts/utils.R")
  config_file <- here("code/Rscripts/config.R")
}
if (file.exists(utils_file))  source(utils_file)
if (file.exists(config_file)) source(config_file)

message("run_mode = ", run_mode)
use_cached <- run_mode == "published"

Results   <- file.path(out_root, "Results")
Figures   <- file.path(out_root, "Figures", "integraton_and_markers")
Data_calc <- file.path(out_root, "Data_calculated")

dirs <- c(Results, Figures)
if (!use_cached) dirs <- c(dirs, Data_calc)   # only a rerun writes cluster objects
for (d in dirs) dir.create(d, recursive = TRUE, showWarnings = FALSE)

```


```{r helpers}
res_vector <- paste0("SCT_snn_res.", c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1, 1.1, 1.2))

# same rule as before, now in one place so both modes use it
select_dims <- function(obj, min_dims = 10) {
  stdevs <- obj[["pca"]]@stdev
  pct <- stdevs / sum(stdevs) * 100
  cum <- cumsum(pct)
  co1 <- which(cum > 90)[1]   # 90% of total variance explained
  co2 <- sort(which((pct[1:(length(pct) - 1)] - pct[2:length(pct)]) > 0.1), decreasing = TRUE)[1] + 1
  d <- min(co1, co2)
  if (is.na(d) || d < min_dims) d <- min_dims
  d
}
```

```{r input}
# input for a rerun: part 1's output of the same mode, else the shipped one
if (!use_cached) {
  seu_final_file <- file.path(Data_calc, "seu_final.rds")
  if (!file.exists(seu_final_file)) seu_final_file <- file.path(published_dir, "seu_final.rds")
  message("rerun input: ", seu_final_file)
}
```


This part clusters each cell line (EpiC, kolf2) separately and tests resolutions 0.1–1.2:
SCTransform (regressing S and G2M scores), PCA, automatic selection of the number of PCs
(minimum of the PC count explaining 90% of variance and the last PC with a >0.1% drop in
variance, at least 10), UMAP, graph-based clustering, then cluster composition per sample,
module scores of marker sets, and top markers (Wilcoxon, RNA assay) for every resolution.

**Run modes.** Clustering and UMAP are stochastic and depend on package versions, so cluster
numbers can change between runs. The manuscript figures were made from the objects in
`data/published_objects/`.
* `published` (default): loads the shipped cluster objects and regenerates all plots and
  marker tables from them. Reproduces the manuscript figures. 
* `rerun`: repeats the whole procedure. Outputs go to `results/rerun/` and may differ; 
  the cluster-to-label dictionaries in `config.R` must then be re-checked.

Mode of this report: `r run_mode`.

```{r clustering and markers}
for (ct in c("EpiC", "kolf2")){
  print(ct)
  clusters_file <- file.path(if (use_cached) published_dir else Data_calc,
                             paste0("seu_integrated_clusters_", ct, "_all_nHVG", nHVG_str, ".rds"))

  if (use_cached) {
    # ---- published: reuse the shipped object (PCA, UMAP, clusters already inside) ----
    stopifnot("shipped cluster object missing" = file.exists(clusters_file))
    seu.integrated <- LoadSeuratRds(clusters_file)
    stopifnot("resolution columns missing" = all(res_vector %in% colnames(seu.integrated@meta.data)))
    selected_dims <- select_dims(seu.integrated)

  } else {
    # ---- rerun: full stochastic pipeline ----
    seu.norm <- LoadSeuratRds(file = seu_final_file)
    message("Starting processing for: ", ct)
    seu.cell <- subset(seu.norm, subset = sample.names %in% grep(ct, unique(sample.names), value = TRUE))
    seu.cell <- JoinLayers(seu.cell)
    seu.integrated <- seu.cell
    SaveSeuratRds(seu.integrated, file = file.path(Data_calc, paste0("seu_integrated_", ct, "_all_nHVG", nHVG_str, ".rds")))

    seu.integrated <- SCTransform(seu.integrated, method = "glmGamPoi",
                                  variable.features.n = 2000,
                                  vars.to.regress = c("S.Score", "G2M.Score"),
                                  verbose = F)
    seu.integrated <- RunPCA(seu.integrated, verbose = F)
    selected_dims <- select_dims(seu.integrated)

    seu.integrated <- RunUMAP(seu.integrated, reduction = "pca", dims = 1:selected_dims, verbose = F)
    seu.integrated <- FindNeighbors(object = seu.integrated, dims = 1:selected_dims)
    seu.integrated <- FindClusters(object = seu.integrated, resolution = seq(0.1, 1.2, by = 0.1))
    # NOTE: as in the published run, this second RunUMAP has no 'dims' (uses the default 30 PCs)
    # and overwrites the first embedding. Kept on purpose to match the published object.
    seu.integrated <- RunUMAP(object = seu.integrated, dims = 1:selected_dims)

    SaveSeuratRds(seu.integrated, file = clusters_file)
  }
  print(paste(ct, "| Auto-selected Dims:", selected_dims))

  # ---- deterministic from here on, identical in both modes ----
  p <- ElbowPlot(seu.integrated, ndims = ceiling(1.5 * selected_dims)) +
    geom_vline(xintercept = selected_dims, color = "red")
  ggsave(file.path(Figures, paste0("elbow_plot_", ct, "_all_nHVG", nHVG_str, ".pdf")), p, width = 6, height = 6)

  p <- clustree::clustree(seu.integrated@meta.data[, grep("SCT_snn_res", colnames(seu.integrated@meta.data))],
                          prefix = "SCT_snn_res.")
  ggsave(file.path(Figures, paste0("clustree_", ct, "_all_nHVG", nHVG_str, ".pdf")), p, width = 20, height = 20)

  p <- DimPlot(seu.integrated, reduction = "umap", group.by = res_vector)
  ggsave(file.path(Figures, paste0("clustree_umap_", ct, "_all_nHVG", nHVG_str, ".pdf")), p, width = 20, height = 20)

    # Van-Haver markers - ensure genes are present
  marker_list_filtered <- lapply(marker_list, function(genes) {
    genes[genes %in% rownames(seu.integrated)]
  })
  
  # subtracting the average expression of "control genes" from the feature genes
  seu.integrated <- AddModuleScore(
      seu.integrated,
      features = marker_list_filtered,
      name     = names(marker_list_filtered),
      nbin     = 10
    )
  
  # add nice metadata columns (names of the marker_list_filtered list plus a "1" suffix)
  new_cols <- grep(paste0("(", paste(names(marker_list_filtered), collapse="|"), ")\\d+"), 
               colnames(seu.integrated@meta.data), value = TRUE)
  marker_names <- gsub("\\d+$", "", new_cols)
  names(marker_names) <- new_cols
  colnames(seu.integrated@meta.data)[colnames(seu.integrated@meta.data) %in% new_cols] <- (
    marker_names[match(colnames(seu.integrated@meta.data)[colnames(seu.integrated@meta.data) %in% new_cols], names(marker_names))])
  
  # selres <- "SCT_snn_res.0.9"
  top_markers_outer_list <- list()
  wb <- createWorkbook()
  for (selres in res_vector){
    seu.integrated@meta.data$seurat_clusters <- seu.integrated@meta.data[[selres]]
    Idents(seu.integrated) <- selres
    # checks:
    cluster_sample_table <- table(
      #Cluster = Idents(seu.integrated),
      Cluster = seu.integrated@meta.data$seurat_clusters,
      Sample  = seu.integrated$sample.names
    )
    # print(cluster_sample_table)
    cluster_sample_prop <- prop.table(cluster_sample_table, margin = 2)
    # print(round(cluster_sample_prop, 3))
    # tidy version for exporting
    cluster_sample_df <- as.data.frame(cluster_sample_table) %>%
      group_by(Sample) %>%
      mutate(Prop = Freq / sum(Freq))
    
    # visualize:
    p <- ggplot(cluster_sample_df,
                aes(x = Sample, y = Prop, fill = Cluster)) +
      geom_bar(stat = "identity", width = 0.8) +
      scale_y_continuous(labels = scales::percent_format()) +
      labs(
        x = NULL,
        y = "Proportion of cells",
        fill = "Cluster"
      ) +
      theme_classic() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.height = unit(0.4, "cm")
      )
    plot(p)
    outfig <- file.path(Figures, paste("cluster_proportions_", selres, "_", ct, "_all_nHVG", nHVG_str, ".pdf", sep=""))
    ggsave(outfig, plot=p, width=8, height=6)
  
    # to plot van-Haver markers:
    # add hulls
    my_hulls <- get_glasswork_layers(
          seu.integrated, 
          clusvar = "seurat_clusters",
          concavity = 3,         # Slightly higher concavity follows the dense core better
          core_quantile = 0.75   # This ignores the 25% most isolated cells
        )
    
    my_hulls_outline <- get_glasswork_layers_outline(
      seu.integrated, 
      clusvar = "seurat_clusters",
      concavity = 3,         # Slightly higher concavity follows the dense core better
      core_quantile = 0.75   # This ignores the 25% most isolated cells
    )
    
    # 2. Add them to your plot
    # I recommend setting alpha low on the points so the Hulls "pop"
    p <- DimPlot(
      seu.integrated, ncol=2,
      reduction = "umap",
      group.by  = "sample.names",
      split.by  = "sample.names",
      pt.size   = 0.1) +
      my_hulls +
      NoLegend() +
      theme_void()
    plot(p)
    
    outfig <- file.path(Figures, paste("clusters_samples_", ct, "_res_", selres, "_all_nHVG", nHVG_str, ".pdf", sep=""))
    ggsave(outfig, plot=p, width=15, height=10)


   # Van-Haver markers - ensure genes are present
   marker_list_filtered <- lapply(marker_list, function(genes) {
      genes[genes %in% rownames(seu.integrated)]
    })
    
    pp <- list()
    for (marker_name in marker_names){
      pp[[marker_name]] <- FeaturePlot(
        seu.integrated, 
        label=T, label.size = 7,
        features  = marker_name,
        reduction = "umap",
        min.cutoff = "q05",
        max.cutoff = "q95"
      ) +
        my_hulls_outline +
        scale_colour_viridis_c(option = "plasma", direction = -1) +
        NoLegend() +
        theme_void()
    }
    p_combined <- pp[[1]] + pp[[2]] + pp[[3]] + pp[[4]] + pp[[5]] + pp[[6]]
    
    # Define the layout: 
    layout <- "ABC
               DEF"
    
    p_final <- p_combined + 
      plot_layout(design = layout)
    
    plot(p_final)
    outfig <- file.path(Figures, paste("clusters_vanHaver_markers_", ct, "_res_", selres, "_all_nHVG", nHVG_str, ".pdf", sep=""))
    ggsave(outfig, plot=p_final, width=15, height=10)
    
    
    
    # marker detection - TO TEST - use "SCT"
    DefaultAssay(seu.integrated) <- "RNA"
    
    markers <- FindAllMarkers(
      seu.integrated,
      only.pos = TRUE,
      test.use = "wilcox",
      logfc.threshold = 0.25,
      min.pct = 0.1
    )
    
    top_markers <- markers %>%
      group_by(cluster) %>%
      slice_max(order_by = avg_log2FC, n = 20)
    
    # for now, filter by patterns; TO DO - filter by the gtf file!
    top_markers_clean <- markers %>%
      filter(!grepl("^ENSG|^LINC|\\-AS[0-9]$|^MIR|^SNHG", gene)) %>% 
      group_by(cluster) %>%
      slice_max(order_by = avg_log2FC, n = 50)
    
    # create a list for the dotplot
    top_markers_list <- list()
    for (cl in unique(top_markers_clean$cluster)){
      top_markers_list[[cl]] <- top_markers_clean$gene[top_markers_clean$cluster==cl]
    }
  
    gene_all <- c()
    top_markers_flat <- top_markers_list
    for (mname in names(top_markers_flat)){
      top_markers_flat[[mname]] <- setdiff(top_markers_flat[[mname]], gene_all)
      gene_all <- union(gene_all, top_markers_flat[[mname]])
    }
  
    p_dot <- DotPlot(
      seu.integrated,
      features = top_markers_flat, # Passing the list creates grouped labels
      cols = c("lightgrey", "red"),
      col.min = 0,
      dot.scale = 6,
      assay = "SCT"
    ) + RotatedAxis() + ggtitle("Top Markers")+
      theme(
    axis.text.x = element_text(size = 7))
    plot(p_dot)
    
    ggsave(file.path(Figures, paste("Top_Markers_DotPlot_", ct, "_selres_", 
              selres, "_all_nHVG", nHVG_str, ".pdf", sep="")), p_dot, 
           width = length(top_markers_list)*5.5, height = 6, limitsize = FALSE)
  
    # write markers to file
      # save top_markers_list
    max_len <- max(sapply(top_markers_list, length))
    padded_list <- lapply(top_markers_list, function(x) {
      length(x) <- max_len
      return(x)}
      ) # 2. Pad shorter lists with NA so they all match max_len
    padded_df <- as.data.frame(padded_list) # now there will be column names
    colnames(padded_df) <- gsub("X", "Cluster_", colnames(padded_df))
    top_markers_outer_list[[selres]] <- padded_df
  
    # 2. Add a new sheet for EACH resolution inside the loop
    # Clean the name (Excel sheets can't have certain characters or be too long)
    sheet_name <- substr(gsub("[^[:alnum:]]", "_", selres), 1, 31) 
    addWorksheet(wb, sheetName = sheet_name)
    writeData(wb, sheet = sheet_name, x = padded_df)
    #
  

  } # loop for selres
  # 3. Save the final workbook
  saveWorkbook(wb, file.path(Results, paste("Top_Markers_By_Sample_", ct, "_all_nHVG", nHVG_str, ".xlsx", sep="")), overwrite = TRUE)

} # loop for celltype
```


```{r session info}
sessionInfo()
```
