---
title: "Analysis of single cell neuroblastoma organoids: part 3 - selected resolutions"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    # css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---
```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.width=12, fig.height=5)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).
The objective of this R Markdown code is to analyse the results obtained by the organoid single cell RNA sequencing.

```{r libraries}

packages <- c(
  "Seurat",
  "ggplot2",
  "patchwork",    
  "Matrix",
  "dplyr",
  "here",
  "stringr",      
  "future",       
  "clustree",     
  "RColorBrewer",
  "ggforce",
  "concaveman",
  "openxlsx",
  "viridis",      
  "msigdbr"   
)
invisible(lapply(packages, library, character.only = TRUE))

```


Folder structure is defined first
```{r}

on_codeocean <- dir.exists("/data") && dir.exists("/code")

if (on_codeocean) {
  Datafolder  <- "/data/Data_organoid"
  utils_file  <- "/code/Rscripts/utils.R"
  config_file <- "/code/Rscripts/config.R"
} else {
  Datafolder  <- here("data/Data_organoid")
  utils_file  <- here("code/Rscripts/utils.R")
  config_file <- here("code/Rscripts/config.R")
}
print(utils_file)
print(config_file)
if (file.exists(config_file)) source(config_file)   # run_mode, base_dir, published_dir, out_root, nHVG_str, ...
if (file.exists(utils_file))  source(utils_file)

message("run_mode = ", run_mode)

Results          <- file.path(out_root, "Results")
Figures          <- file.path(out_root, "Figures", "selected_resolution")
Figures_dimplots <- file.path(Figures, "dimplots")

# cluster objects: shipped ones in published mode, part 2's output in rerun mode
clusters_dir <- if (run_mode == "published") published_dir else file.path(out_root, "Data_calculated")

for (d in c(Results, Figures, Figures_dimplots)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
```


# First, create MYCN - EGFP dotplot
```{r MYCN - EGFP dotplot}
outfile <- file.path(clusters_dir, paste("seu_integrated_clusters_", "EpiC", "_all_nHVG", nHVG_str, ".rds", sep=""))
seu.integrated_EpiC <- LoadSeuratRds(outfile)
outfile <- file.path(clusters_dir, paste("seu_integrated_clusters_", "kolf2", "_all_nHVG", nHVG_str, ".rds", sep=""))
seu.integrated_kolf2 <- LoadSeuratRds(outfile)

# Extract DotPlot data from each object
p1_data <- DotPlot(seu.integrated_EpiC, features = c("MYCN", "EGFP"), group.by = "sample.names", assay = "RNA")$data
p1_data <- p1_data[p1_data$pct.exp > 0, ] # remove zero percent dots!
p2_data <- DotPlot(seu.integrated_kolf2, features = c("MYCN", "EGFP"), group.by = "sample.names", assay = "RNA")$data
p2_data <- p2_data[p2_data$pct.exp > 0, ] # remove zero percent dots!

nice_sample_order <- unname(sample_names_dict)

# Combine, rename, filter, and scale
df_combined <- rbind(p1_data, p2_data) %>%
  mutate(
    # Recode raw sample IDs to nice names
    id = dplyr::recode(as.character(id), !!!sample_names_dict)
  ) %>%
  # Filter background noise
  filter(pct.exp > 2) %>%
  mutate(
    # Lock factor levels to dictionary order, reversed (so item 1 is at the top)
    id = factor(id, levels = rev(nice_sample_order))
  ) %>%
  # Calculate scaled expression per feature across all remaining samples
  group_by(features.plot) %>%
  mutate(avg.exp.scaled = as.vector(scale(avg.exp))) %>%
  ungroup()

# Plot manually with ggplot2
p <- ggplot(df_combined, aes(x = features.plot, y = id, size = pct.exp, color = avg.exp.scaled)) +
  geom_point() +
  scale_color_gradient(low = "lightgrey", high = "red") +
  scale_size_area(max_size = 6) +
  # Prevents empty samples from being removed from the Y-axis tick labels
  scale_y_discrete(drop = FALSE) +
  theme_bw() +
  RotatedAxis() +
  labs(x = NULL, y = NULL, size = "Percent Expressed", color = "Average Expression")

# Display and save 
ggsave(file.path(Figures, "MYCN_EGFP_dotplot.pdf"), p, width = 6, height = 6)


# also a nice dimplot is needed

egfp_epic <- FetchData(seu.integrated_EpiC, vars = "EGFP")[, 1]
egfp_kolf <- FetchData(seu.integrated_kolf2, vars = "EGFP")[, 1]

objects_list <- list(
  "EpiC" = seu.integrated_EpiC,
  "kolf2" = seu.integrated_kolf2
)

for (ct in names(objects_list)) {
  seu_obj <- objects_list[[ct]]
  
  p_list <- FeaturePlot(seu_obj, features = c("EGFP"), combine = FALSE, split.by = "sample.names")
  p_list_notitles <- lapply(p_list, function(p) {
    p +
      scale_y_continuous(sec.axis = waiver())
  })
  print(p_list_notitles)
  p_list_colored <- lapply(p_list_notitles, function(p) {
    p +
      scale_colour_gradient(
        low = "lightgrey",
        high = "purple",
        limits = c(min_val, max_val),
        oob = scales::squish
      )
  })
  p_list_clean <- lapply(p_list_colored, add_corner_axes)
  for (pi in 1:length(p_list_clean)){
    pname <- unique(seu_obj$sample.names)[pi]
    pf <- p_list_clean[[pi]]
    # save each
    selres <- selres_list[[ct]]
      ggsave(file.path(Figures_dimplots, paste("EGFP_dimplot_", ct, "_selres_", selres,
                                "_all_nHVG", nHVG_str, "_", pname, ".pdf", sep="")), 
       pf, width = 4, height = 4) 
  }

  p_combined <- wrap_plots(p_list_clean, ncol = 2)
  ggsave(
    file.path(Figures, paste0("EGFP_DimPlot_split_", ct, "_all_nHVG", nHVG_str, ".pdf")),
    p_combined, width = 8, height = 8
  )
  
  # repeat without split
  p_list <- FeaturePlot(seu_obj, features = c("EGFP"), combine = FALSE)
  p_list_clean <- lapply(p_list, add_corner_axes)
  for (pi in 1:length(p_list_clean)){
    pname <- names(p_list)[pi]
    pf <- p_list_clean[[pi]] + scale_colour_gradient(
          low = "lightgrey",
          high = "purple",
          limits = c(min_val, max_val),
          oob = scales::squish
        )
  }  
  ggsave(
      file.path(Figures, paste0("EGFP_DimPlot_nosplit_", ct, "_all_nHVG", nHVG_str, ".pdf")),
      p_list_clean[[1]], width = 4, height = 4
  )
}




# delete large Seurat objects
rm(seu.integrated_EpiC, seu.integrated_kolf2)
gc()
```



Normalised data for each cell type is imported.
```{r EpiC}

for (ct in c("EpiC", "kolf2")){
  outfile <- file.path(clusters_dir, paste("seu_integrated_clusters_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
  seu.integrated <- LoadSeuratRds(outfile)
  # write out cell numbers
  table(seu.integrated$sample.names)
  # use nice sample names as well:
  seu.integrated$sample_names_renamed <- dplyr::recode(
    as.character(seu.integrated$sample.names),
    !!!sample_names_dict )
  seu.integrated$sample_names_renamed <- factor(seu.integrated$sample_names_renamed, levels = unique(seu.integrated$sample_names_renamed))

  
  # add van-Haver markers
       # Van-Haver markers - ensure genes are present
  marker_list_filtered <- lapply(marker_list, function(genes) {
      genes[genes %in% rownames(seu.integrated)]
    })
    
  # subtracting the average expression of "control genes" from the feature genes
  seu.integrated <- AddModuleScore(
      seu.integrated,
      features = marker_list_filtered,
      name     = names(marker_list_filtered),
      nbin     = 10
    )
  
    # needs to be done for the marker names
  # copied from the anal2_manuscript...
  new_cols <- grep(paste0("(", paste(names(marker_list_filtered), collapse="|"), ")\\d+"), 
               colnames(seu.integrated@meta.data), value = TRUE)
  marker_names <- gsub("\\d+$", "", new_cols)
  names(marker_names) <- new_cols
  colnames(seu.integrated@meta.data)[colnames(seu.integrated@meta.data) %in% new_cols] <- (
    marker_names[match(colnames(seu.integrated@meta.data)[colnames(seu.integrated@meta.data) %in% new_cols], names(marker_names))])

  # select selected selres
  selres <- selres_list[[ct]]
  #seu.integrated@meta.data$seurat_clusters <- seu.integrated@meta.data[[selres]]
  Idents(seu.integrated) <- selres
  #print(table(Idents(seu.integrated)))
  
  # rewrite names 
  new.cluster.ids <- new.cluster.ids.list[[ct]]
  seu.integrated <- SeuratObject::RenameIdents(seu.integrated, new.cluster.ids)
  seu.integrated$cluster_type <- Idents(seu.integrated)
  
  # remove "stray" cells
  cell_counts <- table(seu.integrated$sample.names, seu.integrated$cluster_type)
  counts_df <- as.data.frame(cell_counts)
  colnames(counts_df) <- c("sample.names", "cluster_type", "Freq")
  keep_combinations <- counts_df[counts_df$Freq >= 10, ] # # 3. Identify the combinations that actually have 10 (15) or more cells
  valid_pairs <- paste(keep_combinations$sample.names, keep_combinations$cluster_type, sep = "_")
  cell_pairs <- paste(seu.integrated$sample.names, seu.integrated$cluster_type, sep = "_")
  
  # Subset the Seurat object to keep only cells in those valid pairs
  seu.integrated <- subset(seu.integrated, cells = which(cell_pairs %in% valid_pairs))
  table(seu.integrated$sample.names, seu.integrated$cluster_type)
  
  # set up the order
  bio_order <- bio_order_list[[ct]]
  seu.integrated$cluster_type <- factor(
    as.character(seu.integrated$cluster_type), 
    levels = bio_order
  )
  
  # remove levels not populated
  # seu.integrated$cluster_type <- droplevels(seu.integrated$cluster_type)
  # 
  # if (run_mode == "published") {
  #   exp <- read.csv(file.path(published_dir, paste0("expected_cell_counts_", ct, ".csv")))
  #   obs <- as.data.frame(table(sample.names = seu.integrated$sample.names,
  #                              cluster_type = seu.integrated$cluster_type))
  #   m <- merge(exp, obs, by = c("sample.names", "cluster_type"), all = TRUE,
  #              suffixes = c(".exp", ".obs"))
  #   m[is.na(m)] <- 0
  #   stopifnot("cell counts differ from published objects" = all(m$Freq.exp == m$Freq.obs))
  # }
  
  # save object with final cluster types
  outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
  SaveSeuratRds(seu.integrated, file=outfile)
  
    
}
  
```


Temporary for GEO upload - merge the two integrated seurat objects
```{r for GEO upload}
outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", "both_cells", "_all_nHVG", nHVG_str, ".rds", sep=""))

if (!file.exists(outfile)){
  outfile1 <- file.path(Results, paste("seu_integrated_clusters_newnames_", "EpiC", "_all_nHVG", nHVG_str, ".rds", sep=""))
  seu.integrated1 <-  LoadSeuratRds(file=outfile1)
  outfile2 <- file.path(Results, paste("seu_integrated_clusters_newnames_", "kolf2", "_all_nHVG", nHVG_str, ".rds", sep=""))
  seu.integrated2 <-  LoadSeuratRds(file=outfile2)
  # merge
  seu.all <- merge(x = seu.integrated1 , 
               y = seu.integrated2)
  seu.all$cell_type <- stringr::str_split_i(Cells(seu.all), "_", 2)
  
  cols <- names(seu.all@meta.data)
  new_order <- c(cols[1], "cell_type", setdiff(cols[-1], "cell_type"))
  # seu.all@meta.data <- seu.all@meta.data[, new_order]
  seu.all@meta.data <- seu.all[[new_order]]
  
  # save
  SaveSeuratRds(seu.all, file=outfile)
}


```

Session info for reproducibility
```{r session info}
sessionInfo()
```
