---
title: "Analysis of single cell neuroblastoma organoids: part 2 - obtain cell type markers"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---
```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).
The objective of this R Markdown code is to analyse the results obtained by the single cell RNA sequencing.

```{r libraries}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
# # install CRAN packages via BioConductor as well
# if (!"BiocManager" %in% installed.packages()[,"Package"]){
#   install.packages("BiocManager")
# }
# list.of.packages <- c("ggplot2", "ComplexHeatmap", "tidyHeatmap", 
#                       "circlize", "limma", "edgeR", "dplyr", "tidyr",
#                       "UpSetR", "ComplexUpset", "VennDiagram", "readr", "openxlsx", "readxl", 
#                       "ggmagnify", "readr", "ggrepel")
# new.packages <- setdiff(list.of.packages, installed.packages()[,"Package"])
# if(length(new.packages)) BiocManager::install(new.packages)

library(Seurat)
library(DoubletFinder) # for doublet detection
library(scDblFinder) # a newer doublet detection tool
library(SeuratWrappers)
library(monocle3)
library(ggplot2)
library(patchwork) # to combine plots
library(ggforce) # to add ellipse?
library(Matrix)
library(dplyr)
library(here)
library(stringr) # for the string replacements
library(future) # for parallel Seurat runs
library(clustree) # to visualize clusters
library(scCustomize) # use scVI for integration, such as in Saldana-Guerero
library(RColorBrewer)
# for concave hulls:
library(ggforce)
library(concaveman)
library(openxlsx)
library(future) # in this case to stop parallelization

library(limma)
library(edgeR) # both for pseudo-bulk analysis
# for differential expression analysis
library(DESeq2)
library(SummarizedExperiment)
library(tidyverse)
library(here) # Excellent for path construction
library(biomaRt) # for Ensembl ID harmonization
library(EnhancedVolcano)

# for TPM calculation
#library(GenomicFeatures)
#library(TxDb.Hsapiens.UCSC.hg38.knownGene) # not working!

# for functional analysis and output
library(msigdbr)
library(clusterProfiler)
library(dplyr)
library(DT)
library(htmlwidgets)
library(htmltools)

# for pathview
library(pathview)
```


<!-- # Biztosan kellene sejt -->
<!-- # populáció/cluster azonosítás marker gének alapján, pseudotime analízis -->
<!-- # és CellChat vagy CellPhoneDB analízis is -> do CellChat: https://github.com/jinworks/CellChat1 -->

Folder structure
```{r}
Datafolder     <- here("Data")
Data_calc <- here("Data_calculated") # Path to find Part 1 and 2 outputs

# Subdirectories specifically isolated for this run's outputs
Figures   <- here(file.path("Figures", "pseudo_bulk_SCP_EGFP_tumor"))
Data_calc <- here("Data_calculated")
Data_public <- here("Data_public")
Results   <- here("Results")

if (!file.exists(Figures))   dir.create(Figures, recursive = TRUE)
if (!file.exists(Data_calc)) dir.create(Data_calc, recursive = TRUE)
if (!file.exists(Results))   dir.create(Results, recursive = TRUE)

utils_file <- here("Rscripts","utils.R")
if (file.exists(utils_file)) source(utils_file)
```

This is the same chunk as in anal3_selected_resolution -> better organization is needed than hard copy!
Data is read and integrated separately for each cell type
```{r}

test_markers <- c("ASCL1", "CHGA", "HAND2", "TH")

marker_list <- list(
  proliferation = c("CDK1", "CENPF", "TOP2A", "UBE2C"), #vanHaver
  #sympathoblast = c("CDK1", "PRPH", "ELAVL4", "ELAVL3", "ELAVL2", "ISL1", "STMN2", "GAP43", "GATA3", "PHOX2B", "SOX11", "TFAP2B",  "HMX1"), # Saldana?
  sympathoblast = c("PRPH", "ELAVL2", "ELAVL3", "ELAVL4", "ISL1","STMN2", "STMN4", "GAP43", "GATA3", "PHOX2B", "SOX11", "SOX11", "TFAP2B"), #vanHaver
  chromaffin = c("ADM", "CARTPT", "CHGA", "CHGB", "DBH", "DDC", "DLK1", "EGR1", "FOS", "JUNB", "PENK", "PNMT", "TH"), #vanHaver
  #bridging = c("ASCL1", "MEIS1"), #vanHaver
  bridging  = c("ASCL1", "MEIS1", "PHOX2B", "PENK", "TH", "CHGA", "HAND2"),
  SCP = c("CDH19", "EDNRB", "ERBB3", "ERBB4", "FTS1", "FOXD3", "MPZ", "NR2F2", "PLP1", "SOX10", "SOX2", "SPARC"), #vanHaver
  #SCP_other = c("SOX10", "PLP1", "FOXD3", "FABP7", "S100B", "ERBB3", "NGFR", "MPZ", "COL2A1", "POSTN", "MOXD1", "GAS7", "ASCL1"), #Saldana?
  MES = c("COL1A1", "COL1A2", "COL12A1", "COL3A1", "VIM", "CXCL12", "TWIST1", "TWIST2") #Saldana fig2
)

vanHaver_list <- list(
  immune = c(    "FCGR1A", "CD163" ),
  liver = c(    "HNF4A",    "AHSG"  ),
  erythroid = c(    "HBA2",    "HBB"  ),
  adrenal_gland = c(    "NR5A1",    "STAR"  ),
  CPC = c(    "PENK",    "CHGA",    "STMN2"  ),
  sympathoblasts = c(    "STMN2"),
  prolif_sym = c(    "STMN2",    "TOP2A"  ),
  bridging = c(    "ASCL1"  ),
  SCP = c(    "PLP1",    "SOX10"  ),
  #subepicardial_abdominal_mesenchyme = c(    "PRRX1"  ),
  sa_mes = c(    "PRRX1"  ),
  endothelial = c(    "PECAM1",     "KDR"   ),
  melanocyte_stem = c(    "MITF",    "PMEL"  ),
  #endoneurial_fibroblasts = c(    "DCN",    "FBN1"  ),
  endoneurial_fibroblasts = c(    "DCN",    "FBN1"  ),
  myofibroblasts = c(    "ACTA2",    "TAGLN"  )
)

hox_list <- list(
  "Cervical" = c("HOXA1", "HOXB1", "HOXA3", "HOXB3", "HOXD3"),
  "Truncal" = c("HOXA5", "HOXB5", "HOXA6", "HOXA7", "HOXB7", "HOXB8", "HOXC8", "HOXD8", "HOXB9"),
  "Lumbar-caudal" = c("HOXC9", "HOXD9", "HOXD11")
)

structure_list <- list(
  "Basement Membrane" = c("LAMA2", "LAMB1", "LAMC3", "COL4A6", "COL12A1", "NID2"),
  "ECM Assembly"      = c("COL1A1", "COL3A1", "COL6A3", "COL8A1", "BGN", "LUM", "POSTN", "FMOD", "DCN"),
  "Cell Adhesion"     = c("CDH19", "CDH11", "CDH20", "MPZ", "CNTN2", "CNTNAP5", "CLDN5"),
  "Tissue Patterning" = c("WNT1", "WNT7B", "WNT10B", "RSPO3", "DKK1", "DKK2", "SLITRK6", "ROBO3"),
  "Axon/Structure"    = c("NRG1", "ERBB3", "LGI4", "RELN", "POU4F1", "UNCX")
)

# for the final manuscript:
dimplot_list <- list(
  "Cell Fate" = c("TH", "HOX2B", "GAP43", "STMN2", "ISL1", "CDKN1C", "CCND1"),
  "Neuromesodermal progenitor" = c("TBXT", "SOX2"), # TBXT: Brachyury
  "Proliferation" = c("PRPH", "SYN3", "GAP43", "NTRK1", "SOX4", "ALK", "MEIS2", "ERBB4", "VGF"),
  "GFP" = c("EGFP", "MYCN"),
  "Maturation Sympathoblasts" = c("CHRNA3", "VMAT1"), # VMAT1 missing
  "Enteric Neurons" = c("EDNRB", "RET"),
  "Sensory Neurons" = c("NGN1", "NGN2"), # missing?
  "Endoneurial Fibroblasts" = c("DCN", "FBN1"),
  "Myofibroblasts" = c("ACTA2", "TAGLN"),
  "Melanocyte stem cell" = c("PMEL", "MITF"),
  "SCP" = c("FOXD3", "SOX2", "SOX10", "MPZ", "S100B", "PLP1", "ERBB3"),
  "Briding" = c("ASCL1", "PHOX2B", "PENK", "TH", "CHGA", "HAND2"),
  "Symphatoblast" = c("GATA3", "ELAVL2", "ELAVL4", "ISL1", "STMN2", "STMN4", "PRPH")
)

violinplot_array <- c("AZIN1", "TFAP2B", "PRPH", "SYN3", "NTRK1", 
                      "SOX4", "ALK", "MEIS2", "ERBB4", "VGF", "ACTB")

# based on testing each selres differently, these selres are used for further analysis
selres_list <- list(
    "EpiC" = "SCT_snn_res.0.3",
    "kolf2" = "SCT_snn_res.0.5" # changed to 0.5!!!
)

sample_names_dict <- c(
    "1_EpiC_MIX_D26" = "H1 WT D26",
    "2_EpiC_STOP_MYCN_D26" = "H1 STOP-MYCN D26",
    "3_EpiC_STOP_MYCN_D42" = "H1 STOP-MYCN no virus D42",
    "4_EpiC_STOP_MYCN_D42_hTH" = "H1 STOP-MYCN hTH-Cre D42",
    "5_kolf2_MIX_D26" = "H2 WT D26",
    "6_kolf2_STOP_MYCN_D26" = "H2 STOP-MYCN D26",
    "7_kolf2_STOP_MYCN_D42" = "H2 STOP-MYCN no virus D42",
   "8_kolf2_STOP_MYCN_D42_hTH" = "H2 STOP-MYCN hTH-Cre D42"
  )

# MES - off target
new.cluster.ids.list <- list(
      "EpiC" =      c(
          "0" = "type_early/late_SCP", 
          "1" =  "type_chr_sym", 
          "2" =  "type_early_prog", 
          "3" =   "type_early/late_endoneural_fibroblasts", # previously "type_MES", 
          "4" = "type_early_SYM_like/mature_SYM",  # previously "type_maturing_neurons", # previously type_bridge
          "5" = "type_early/late_myofibroblast",  # previously"type_MES_like",  # previously type_early_MES
          "6" = "type_EGFP_MES",
          "7" =  "type_bridging",   # previously "type_mature_SYM",
          "8" = "type_EGFP_tumor"
        ),
      # "kolf2" =      c(
      #     "0" = "type_chr_sym", 
      #     "1" = "type_SCP", 
      #     "2" = "type_SCP",
      #     "3" = "type_MES", 
      #     "4" = "type_MES", 
      #     "5" = "type_EGFP_tumor", 
      #     "6" = "type_mature_SYM",
      #     "7" = "type_early_prog",
      #     "8" = "type_EGFP_MES"
      #   )
      "kolf2" =      c(
          "6" = "type_chromaffin", 
          "1" = "type_symphatoblast", 
          "7" = "type_bridging", 
          "0" = "type_early_SCP", #
          "3" = "type_late_SCP", #
          "8" = "type_early_endoneural_fibroblasts", #
          "9"=  "type_early_SYM_like", 
          "2" = "type_late_endoneural_fibroblasts", #
          "4" = "type_EGFP_tumor", #
          "5" = "type_mature_sym", #
          "10" = "type_early_prog", #
          "11" = "type_EGFP_MES", #
          "12" = "type_12"
        )
)

cluster_names_dict <- list(
      "EpiC" =  list(
        "D26"=  c(
          "type_early/late_SCP" = "early SCP", 
          "type_chr_sym"= "chromaffin/symphatoblast", 
          "type_early_prog"="early progenitor", 
          "type_early/late_endoneural_fibroblasts"="early endoneural fibroblast", # previously "type_MES", 
          "type_early_SYM_like/mature_SYM"= "early symphatoblast-like",  # previously "type_maturing_neurons", # previously type_bridge
          "type_early/late_myofibroblast"="early myofibroblast",  # previously"type_MES_like",  # previously type_early_MES
          "type_EGFP_MES"="EGFP MES",
          "type_bridging"="bridging",   # previously "type_mature_SYM",
          "type_EGFP_tumor"="EGFP tumor"
        ),
        "D42"=  c(
          "type_early/late_SCP" = "late SCP", 
          "type_chr_sym"= "chromaffin/symphatoblast", 
          "type_early_prog"="late progenitor", 
          "type_early/late_endoneural_fibroblasts"="late endoneural fibroblast", # previously "type_MES", 
          "type_early_SYM_like/mature_SYM"= "mature symphatoblast",  # previously "type_maturing_neurons", # previously type_bridge
          "type_early/late_myofibroblast"="late myofibroblast",  # previously"type_MES_like",  # previously type_early_MES
          "type_EGFP_MES"="EGFP MES",
          "type_bridging"="bridging",   # previously "type_mature_SYM",
          "type_EGFP_tumor"="EGFP tumor"
        ),
        "all"=  c(
          "type_early/late_SCP" = "early/late SCP", 
          "type_chr_sym"= "chromaffin/symphatoblast", 
          "type_early_prog"="early/late progenitor", 
          "type_early/late_endoneural_fibroblasts"="early/late endoneural fibroblast", # previously "type_MES", 
          "type_early_SYM_like/mature_SYM"= "mature symphatoblast / early symphatoblast-like",  # previously "type_maturing_neurons", # previously type_bridge
          "type_early/late_myofibroblast"="early/late myofibroblast",  # previously"type_MES_like",  # previously type_early_MES
          "type_EGFP_MES"="EGFP MES",
          "type_bridging"="bridging",   # previously "type_mature_SYM",
          "type_EGFP_tumor"="EGFP tumor"
        )
      ),
     "kolf2" = list(
        "D26"=  c(
          "type_chromaffin"= "chromaffin", 
          "type_symphatoblast" = "symphatoblast",
          "type_early_SCP" = "early SCP",
          "type_late_SCP" = "late SCP",
          "type_early_endoneural_fibroblasts"="early endoneural fibroblast", # previously "type_MES", 
          "type_early_SYM_like"= "early symphatoblast-like",  # previously "type_maturing_neurons", # previously type_bridge
          "type_late_endoneural_fibroblasts"="late endoneural fibroblast", # previously "type_MES", 
          "type_EGFP_tumor"="EGFP tumor",
          "type_bridging"="bridging",   # previously "type_mature_SYM",
          "type_early_prog"="early progenitor",
          "type_EGFP_MES"="EGFP MES",
          "type_12" = "cluster 12",
          "type_mature_sym" = "mature symphatoblast"
        ),
        "D42"=  c(
          "type_chromaffin"= "chromaffin", 
          "type_symphatoblast" = "symphatoblast",
          "type_early_SCP" = "early SCP",
          "type_late_SCP" = "late SCP",
          "type_early_endoneural_fibroblasts"="early endoneural fibroblast", # previously "type_MES", 
          "type_early_SYM_like"= "early symphatoblast-like",  # previously "type_maturing_neurons", # previously type_bridge
          "type_late_endoneural_fibroblasts"="late endoneural fibroblast", # previously "type_MES", 
          "type_EGFP_tumor"="EGFP tumor",
          "type_bridging"="bridging",   # previously "type_mature_SYM",
          "type_early_prog"="early progenitor",
          "type_EGFP_MES"="EGFP MES",
          "type_12" = "cluster 12",
          "type_mature_sym" = "mature symphatoblast"
        ),
        "all"=  c(
          "type_chromaffin"= "chromaffin", 
          "type_symphatoblast" = "symphatoblast",
          "type_early_SCP" = "early SCP",
          "type_late_SCP" = "late SCP",
          "type_early_endoneural_fibroblasts"="early endoneural fibroblast", # previously "type_MES", 
          "type_early_SYM_like"= "early symphatoblast-like",  # previously "type_maturing_neurons", # previously type_bridge
          "type_late_endoneural_fibroblasts"="late endoneural fibroblast", # previously "type_MES", 
          "type_EGFP_tumor"="EGFP tumor",
          "type_bridging"="bridging",   # previously "type_mature_SYM",
          "type_early_prog"="early progenitor",
          "type_EGFP_MES"="EGFP MES",
          "type_12" = "cluster 12",
          "type_mature_sym" = "mature symphatoblast"
        )
      )
)


bio_order_list <- list("EpiC" = 
                         c("type_early_prog", "type_early/late_myofibroblast", "type_early/late_endoneural_fibroblasts", "type_EGFP_MES",
                           "type_early/late_SCP", "type_chr_sym", "type_bridging", "type_early_SYM_like/mature_SYM",
                "type_EGFP_tumor"),
               "kolf2" = 
                         c("type_early_prog", "type_early_endoneural_fibroblasts", "type_late_endoneural_fibroblasts", "type_EGFP_MES",
                           "type_early_SCP",  "type_late_SCP", "type_bridging", "type_early_SYM_like", "type_chromaffin", "type_symphatoblast", "type_mature_sym",
                "type_EGFP_tumor", "type_12"))

root_list <- list(
   "EpiC"="type_SCP",
   "kolf2"="type_SCP" 
)

# for colour coding the expr levels
min_val <- 0
max_val <- 2.5

# number of genes used for integration
nHVG <- 2000 # 2000 is usually enough to capture sub-states. (5000 HVGs might be too many for a single cell type.)
nHVG_str <- as.character(nHVG)

```



Read in sample and reference data
```{r}
nHVG <- 2000 # 2000 is usually enough to capture sub-states. (5000 HVGs might be too many for a single cell type.)
nHVG_str <- as.character(nHVG)
pseudo_bulk <- list()
for (ct in c("EpiC", "kolf2")){
  
  
    
    # outfile <- file.path(Data_calc, paste("seu_integrated_clusters_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
    # use the new cluster names
    outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
    seu.integrated <- LoadSeuratRds(outfile)
    # write out cell numbers
    table(seu.integrated$sample.names)
    
    # ITT TARTOK
    seu.pseudobulk <- subset(seu.integrated, subset = sample.names %in% grep("D42", unique(sample.names), value = T) &
                               cluster_type_renamed_D42_both %in% c("late SCP", "EGFP tumor"))
    seu.pseudobulk@meta.data <- droplevels(seu.pseudobulk@meta.data)
    table(seu.pseudobulk$sample.names, seu.pseudobulk$cluster_type_renamed_D42_both)
    seu.pseudobulk$plot_id <- paste0(seu.pseudobulk$cluster_type_D42_both, "_", seu.pseudobulk$sample.names)
    Idents(seu.pseudobulk) <- "plot_id"
    table(seu.pseudobulk$plot_id, seu.pseudobulk$sample.names)
    pb_counts <- AggregateExpression(seu.pseudobulk, 
                                 assays = "RNA", 
                                 slot = "counts", 
                                 return.seurat = FALSE)$RNA
    colnames(pb_counts) <- switch(ct,
                              "EpiC"=paste("EpiC", c("SCP_novirus", "SCP_virus", "EGFP_tumor_virus"), sep="_"),
                              "kolf2"=paste("kolf2",c("SCP_novirus", "EGFP_tumor_virus", "SCP_virus"), sep="_")
    )
pseudo_bulk[[ct]] <- as_tibble(pb_counts, rownames="gene_symbol")
}
pb_all <- inner_join(pseudo_bulk[[1]], pseudo_bulk[[2]], by="gene_symbol")
```


# use both samples to do a pseudo-bulk analysis
```{r calculate normalised expression}
# 3. Create design matrix
# Assuming you have a metadata column for condition
sample_annot <- data.frame(
  sample = colnames(pb_all)[2:ncol(pb_all)],
  cell_line = factor(c("EpiC", "EpiC", "EpiC", "kolf2", "kolf2", "kolf2")),
  condition = factor(c("SCP_novirus", "SCP_virus", "EGFP_tumor_virus", 
                       "SCP_novirus", "EGFP_tumor_virus", "SCP_virus"), 
                     levels = c("SCP_novirus", "SCP_virus", "EGFP_tumor_virus"))
)
design <- model.matrix(~ 0 + cell_line + condition, data = sample_annot)
rownames(design) <- sample_annot$sample

counts_matrix <- as.matrix(pb_all[, -1]) # Remove the gene_symbol column
rownames(counts_matrix) <- pb_all$gene_symbol
# 1. Create DGEList object (edgeR)
dge <- edgeR::DGEList(counts = counts_matrix)

# 2. Filter low expression genes (optional but recommended)
keep <- filterByExpr(dge, group = colnames(pb_counts))
dge <- dge[keep, , keep.lib.sizes = FALSE]
dge <- calcNormFactors(dge) # # Apply the TMM normalization (a standard step in the limma-voom workflow)


# 4. Limma-Voom transformation
vv <- voom(dge, design, plot = TRUE)

# save & plot intermediate results
exprmx <- as_tibble(vv$E, rownames = "gene_symbol")
# #print(dim(exprmx))
# # merge with gene annotation
# exprmx <- right_join(gene_annot, exprmx)
# #print(dim(exprmx))

# # save
# tmpfile <- file.path(calcfolder, "edgeR_normcounts.txt")
# write.table(exprmx, file=tmpfile, sep="\t", row.names = F)
exprfile <- file.path(Results, "ps_bulk_logCPM_SCP_EGFP_tumor.xlsx")
write.xlsx(exprmx, file = exprfile, colNames=T, rowNames=F)

```

The distribution of normalised expressions provides a snapshot of the gene expression in each sample. Notice, that the mean gene expression is higher in the UPS samples.
```{r plot normalised expression of samples}
# for the boxplot, create a long tibble first
exprmx_long <- pivot_longer(exprmx, !c(gene_symbol),
                            names_to = "sample", 
                            values_to = "gene_expr")
# # add sample annotation as well
# exprmx_long <- full_join(exprmx_long, samples_annot, by=join_by(sample))

exprmx_long <- exprmx_long %>%
  mutate(
    # Split by '-', take the 2nd element
    cell = str_split_fixed(sample, "-", 5)[, 2],
    
    # Check if the string ends in 'hTH' to define type
    type = ifelse(str_detect(sample, "hTH"), "tumor", "normal")
  )

# Verify the result
table(exprmx_long$sample, exprmx_long$cell)
table(exprmx_long$sample, exprmx_long$type)

exprmx_long$cell <- factor(exprmx_long$cell, levels=c("EpiC", "kolf2"))
exprmx_long$type <- factor(exprmx_long$type, levels=c("normal", "tumor"))


p <- ggplot(exprmx_long, aes(x = sample, y = gene_expr, fill = cell, color = type)) + 
  # Set alpha lower for more transparency
  geom_violin(alpha = 0.4, linewidth = 2) +
  # Boxplot inside with a bit more opacity than the violin
  geom_boxplot(width = 0.1, color = "black", outlier.shape = NA, alpha = 0.6) +
  # Use manual colors for Cancer
  scale_fill_manual(values = c("EpiC" = "orange", "kolf2" = "seagreen")) +
  # Aesthetic labels
  labs(
    y = expression(Log[2] ~ "Normalized Counts (CPM)"),
    x = "Sample ID",
    title = "Gene Expression Distribution per Sample"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    legend.position = "top",
    panel.grid.major.x = element_blank() # Cleans up the background for easier reading
  )

plot(p)

# also save the plot
#figfile <- file.path(figfolder, "boxplot_normalized.pdf")
figfile <- file.path(Figures, "ps_bulk_violinplot_normalized_SCP_EGFP_tumor.pdf")
ggsave(figfile, height=7, width=10)
```


## Principal Component Analysis (PCA)
To assess global sample relationships and identify potential batch effects or outliers, we performed a Principal Component Analysis (PCA) on the log-transformed expression data. 

### Interpretation of Variance
PCA is a dimensionality reduction technique that projects samples onto axes (Principal Components) that capture the greatest sources of variation in the dataset.
* **Sample Clustering:** We observed that hybrid samples exhibit high transcriptomic similarity to their corresponding cancer-type parents, suggesting that the primary cancer lineage remains a dominant driver of the expression profile.
* **Limitations:** It is important to note that PCA captures the **largest** global variances. While this is useful for identifying major biological groups, subtle but biologically significant changes—particularly those involving smaller gene sets—may be masked by these dominant global trends and are better captured through downstream differential expression and pathway analysis.

The following plots visualize the first two principal components, which account for the majority of the total variance in the experiment. Click and drag to magnify.
```{r PCA the  data}
pca <- prcomp(t(vv$E), center = TRUE, scale. = FALSE)
# calculate explained variance
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

sumvar <- 0
for (pcind in 1:length(pca.var)){
  sumvar <- sumvar + pca.var.per[pcind]
  if (sumvar > 95){
    maxi <- pcind
    #print(maxi)
    break
  }
}
vardf <- as.data.frame(cbind('PC'=1:maxi, 'variance'=pca.var.per[1:maxi]))

# #  scree plot
# p <- ggplot(data=vardf) +
#   geom_col(mapping=aes(x=PC, y=cumsum(variance)))+
#   xlab("Principal Component") +  ylab("Cumulative percent variation") +
#   geom_hline(yintercept=95, linetype="dashed",
#              color = "red", linewidth=1)
# plot(p)

# nice labels
nicelabels <- paste0("PC", 1:maxi, " (", vardf$variance, "%)")

pca_df <- as_tibble(pca$x, rownames = "sample")
pca_annot <- inner_join(pca_df, sample_annot, by=join_by(sample))

p <- ggplot(pca_annot, aes(x = PC1, y = PC2, shape = cell_line, color = condition,
                           text = paste("Sample:", sample, "<br>Group:", condition))) +
  geom_point(size = 3) +
  theme_gray(base_size = 18) + 
  scale_color_manual(values = c("EGFP_tumor_virus"="red",
                                "SCP_virus"="purple", "SCP_novirus"="darkblue"))+
  xlab(nicelabels[[1]])+ylab(nicelabels[[2]])
  # geom_text(aes(label=Type_Exp), size = 2.5, vjust =  2.5)
print(p)

#plot(p)
outfig <- file.path(Figures, "pca_samples_SCP_EGFP_tumor.pdf")
ggsave(outfig, plot=p, height = 20, width= 30, units = "cm")

# # create interactive figure
# # tooltip = "text" ensures only your custom label shows up on hover
# ggplotly(p, tooltip = "text")
# 
# # plot B16 only
# p <- ggplot(pca_annot[pca_annot$Cancer=="B16",], aes(x = PC1, y = PC2, shape = Cancer, color = Type)) +
#   geom_point(size = 3) +
#   theme_gray(base_size = 18) + 
#   scale_color_manual(values=c("tumor"="red", "X"="darkblue"))+
#   xlab(nicelabels[[1]])+ylab(nicelabels[[2]])+
#   geom_text(aes(label=Type_Exp), size = 2.5, vjust =  2.5)
# #plot(p)
# outfig <- file.path(figfolder, "pca_samples_B16only.pdf")
# ggsave(outfig, plot=p, height = 20, width= 30, units = "cm")
```


## Differential Gene Expression Analysis

Differential Gene Expression (DGE) was performed to identify transcriptomic shifts across various biological states. For all comparisons, **upregulated** genes are defined as those with significantly higher expression in the second condition of the contrast, while **downregulated** genes are those with significantly lower expression.

### 1. Primary Experimental Contrasts
The following core comparisons were evaluated:

* **UPS vs. X ($UPS\_vs\_X$):** Quantifies expression changes transitioning from the UPS state to the X state.
* **Tumor vs. X:** Quantifies expression changes transitioning from the tumor state to the X state
* **UPS vs. Tumor:** Compares the baseline UPS state against the tumor environment.

### 2. Scope of Comparison (Prefixes)
The prefixes **MEC**, **B16**, or **both** define the samples included in the analysis:

* **MEC / B16:** The comparison is restricted to samples within that specific cancer type.
* **both:** Samples from both cancer types are aggregated. 
    * *Note:* The statistical model utilizes weighting or averaging to account for the unbalanced sample sizes between the B16 and MEC groups, ensuring neither group disproportionately influences the "both" results.

### 3. Sequential Comparisons (B16)
For the B16 samples pairwise comparisons were conducted between sub-conditions:

* **X2 vs. X1**
* **X3 vs. X2**
* **X1 vs. X3**

### Output file
Results are saved to diffexpr_all.xlsx and diffexpr_lfc1_adjp0.05_signif.xlsx files, where each tab contains the results of one of the above contrasts. In diffexpr_lfc1_adjp0.05_signif.xlsx only genes are listed for each contrast whose expression changed with at least 1 log fold change and adjusted p value <= 0.05.

```{r calculate differential gene expression}

# 5. Fit the linear model
# 1. Fit the linear model
fit <- lmFit(vv, design)

# 2. Create the pairwise contrasts for the 3 conditions
contrast.matrix <- makeContrasts(
  # Compare Virus vs No Virus (Baseline)
  SCP_VirusVsNoVirus = conditionSCP_virus,
  
  # # Compare Tumor vs No Virus (Baseline)
  # TumorVsNoVirus = conditionEGFP_tumor_virus,
  
  # Compare Tumor vs Virus (Subtracting the two coefficients)
  EGFP_TumorVsSCP_Virus   = conditionEGFP_tumor_virus - conditionSCP_virus,
  
  # Optional: Keep your cell line comparison to control for background differences
  Kolf2VsEpiC    = cell_linekolf2 - cell_lineEpiC,
  
  levels = design
)

fitc <- contrasts.fit(fit,contrast.matrix )
fite <- eBayes(fitc) ## gives moderated t-statistics
# save result for further use
outfile <- file.path(Results, "limma_fit_diffexpr_SCP_EGFP_tumor.rds")
saveRDS(fite, file=outfile)

# also write out to human readable file
lthr <- 1; pthr <- 0.05
thrchar <- paste(sprintf("lfc%01d", lthr), 
                   sprintf("adjp%.2f", pthr), sep="_")
outdata <- list()
outdata_signif <- list() # only write out genes whose diff expr is signif
for (cf in colnames(fite$coefficients)){
  toptt <- limma::topTable(fite, coef=cf, sort.by = "logFC",
          number=nrow(fite$coefficients))
  toptt <- as_tibble(toptt, rownames = "Gene_ID")
  # # add annotation
  # toptt <- left_join(toptt, gene_annot, by=c("Gene_ID"="gene_id"))
  
  # formatting
  toptt <- toptt %>%
    mutate(
      across(c(logFC, AveExpr, t, B), ~ round(., 2)),
      # P.Value and adj.P.Val formatted as strings to preserve notation in Excel
      P.Value = format.pval(P.Value),
      adj.P.Val = format.pval(adj.P.Val)
    )
  
  # excel tab names
  lname <- substr(cf,1, min(nchar(cf), 31)) # for each tab name, a maximum of 31 length is allowed
  outdata[[lname]] <- toptt

  # repeat with significant genes only
  toptt_thr <- limma::topTable(fite, coef=cf, sort.by = "logFC",
                p.value=pthr, lfc=lthr, number=nrow(fite$coefficients))
  toptt_thr <- as_tibble(toptt_thr, rownames = "Gene_ID")
  # # add annotation
  # toptt_thr <- left_join(toptt_thr, gene_annot, by=c("Gene_ID"="gene_id"))
    # formatting
  toptt_thr <- toptt_thr %>%
    mutate(
      across(c(logFC, AveExpr, t, B), ~ round(., 2)),
      # P.Value and adj.P.Val formatted as strings to preserve notation in Excel
      P.Value = format.pval(P.Value),
      adj.P.Val = format.pval(adj.P.Val)
    )
  outdata_signif[[lname]] <- toptt_thr
}
# write into a compound excel file
outfile <- file.path(Results, "diffexpr_all_SCP_EGFP_tumor.xlsx")
write.xlsx(outdata, file = outfile, colNames=T, rowNames=F)
outfile_signif <- outfile <- file.path(Results, 
                     paste("diffexpr_", thrchar, "_signif_all_SCP_EGFP_tumor.xlsx", sep=""))
write.xlsx(outdata_signif, file = outfile_signif, colNames=T, rowNames=F)

# add volcano plot
res <- topTable(fite, coef = "EGFP_TumorVsSCP_Virus", number = Inf)

# 3. Plot using EnhancedVolcano

p <- EnhancedVolcano(res,
  lab = rownames(res),
  x = 'logFC',
  y = 'adj.P.Val',
  title = 'Tumor vs Normal',
  pCutoff = 0.05,
  FCcutoff = 1.0,
  pointSize = 2.0,
  labSize = 3.0)

outfig <- file.path(Figures, "volcano_plot_SCP_EGFP_tumor.pdf")
ggsave(outfig, plot=p, height = 20, width= 30, units = "cm")

```

```{r Heatmap}
library(pheatmap)
library(dplyr)
library(tibble)

num_genes <- 50

# 1. Modify Column Annotation Metadata with your custom names
heatmap_anno <- sample_annot %>% 
  mutate(
    # ii) Convert cell line IDs: EpiC -> H1, everything else (kolf2) -> H2
    Cell_Line = ifelse(cell_line == "EpiC", "H1", "H2"),
    
    # i) Merge condition and cell state terms into the exact proper combinations
    Condition = case_when(
      condition == "SCP_novirus"        ~ "late SCP STOP-MYCN no virus D42",
      condition == "SCP_virus"          ~ "late SCP STOP-MYCN hTH-Cre D42",
      condition == "EGFP_tumor_virus"   ~ "EGFP tumor STOP-MYCN hTH-Cre D42"
    )
  ) %>%
  column_to_rownames("sample") %>% 
  select(Cell_Line, Condition)


# 2. Define custom colors for the new Condition labels
# (pheatmap requires these names to match your modified character strings exactly)
anno_colors <- list(
  Condition = c(
    "EGFP tumor STOP-MYCN hTH-Cre D42" = "red",
    "late SCP STOP-MYCN hTH-Cre D42"   = "purple",
    "late SCP STOP-MYCN no virus D42"  = "darkblue"
  )
)


# 3. Extract the underlying expression matrix
expr_mat <- if (inherits(vv, "EList")) vv$E else as.matrix(vv)


# 4. Helper function to extract top 50 up & top 50 down non-ENS genes
get_heatmap_genes <- function(limma_fit, coefficient_name) {
  tt <- limma::topTable(limma_fit, coef = coefficient_name, number = Inf)
  tt_filtered <- as_tibble(tt, rownames = "Gene_ID") %>%
    filter(!grepl("^ENS", Gene_ID)) %>% 
    filter(adj.P.Val < 0.05)
  
  top_up <- tt_filtered %>% arrange(desc(logFC)) %>% head(num_genes) %>% pull(Gene_ID)
  top_down <- tt_filtered %>% arrange(logFC) %>% head(num_genes) %>% pull(Gene_ID)
  
  return(c(top_up, top_down))
}


# 5. Gather unique target gene sets across both contrasts
genes_comp1 <- get_heatmap_genes(fite, "SCP_VirusVsNoVirus")
genes_comp2 <- get_heatmap_genes(fite, "EGFP_TumorVsSCP_Virus")
combined_genes <- unique(c(genes_comp1, genes_comp2))

# Subset the master matrix
mat_combined <- expr_mat[combined_genes, , drop = FALSE]


# 6. Generate and save the modified combined heatmap
outfig_combined <- file.path(Figures, "heatmap_combined_progression_SCP_EGFP_tumor.pdf")

# pheatmap(mat_combined,
#          scale = "row",                    
#          annotation_col = heatmap_anno,     
#          annotation_colors = anno_colors,   # 1) Injects your custom color palette
#          show_rownames = TRUE,             
#          show_colnames = F,             
#          cluster_cols = TRUE,              
#          cluster_rows = TRUE,              
#          #main = paste0("Oncogenic Progression Heatmap (", length(combined_genes), " Unique Genes)"),
#          main = "",
#          fontsize_row = 7,                 # 3) Significantly larger gene labels (Up from 4)
#          filename = outfig_combined,       
#          width = 12,                       # Expanded width to handle longer text entries gracefully
#          height = 12)                      # Expanded height so large text doesn't collapse together

pheatmap(t(mat_combined),                            # 1) Transpose the matrix (flips rows and columns)
         scale = "column",                           # 2) Changed from "row" so scaling is still done by gene
         annotation_row = heatmap_anno,              # 3) Changed from annotation_col since samples are now rows
         annotation_colors = anno_colors,   
         show_rownames = FALSE,                      # 4) Changed from TRUE (samples don't show names now)
         show_colnames = TRUE,                       # 5) Changed from FALSE (genes now show names)
         cluster_cols = TRUE,              
         cluster_rows = TRUE,              
         main = "",
         fontsize_col = 7,                           # 6) Changed from fontsize_row so gene labels stay readable
         filename = outfig_combined,       
         width = 12,                        
         height = 6)

library(ComplexHeatmap)

# 1. Generate the heatmap object with an empty string for the legend title
ht <- ComplexHeatmap::pheatmap(
        t(mat_combined),                            
        scale = "column",                           
        annotation_row = heatmap_anno,              
        annotation_colors = anno_colors,   
        show_rownames = FALSE,                      
        show_colnames = TRUE,                       
        cluster_cols = TRUE,              
        cluster_rows = TRUE,              
        main = "z-score",
        fontsize_col = 7,
        # Omit the legend name by setting title to an empty string
        heatmap_legend_param = list(title = "") 
      )

# 2. Open the graphics device
pdf(outfig_combined, width = 14, height = 4) 

# 3. Draw the heatmap with legends on top
draw(ht, 
     heatmap_legend_side = "top", 
     annotation_legend_side = "top")

# 4. Close the device
dev.off()

```



## Functional Enrichment Analysis

To interpret the biological significance of the identified differentially expressed (DGE) and equivalent gene sets, we performed functional enrichment analysis using curated gene set collections from the Molecular Signatures Database (MSigDB).

### Gene Set Collections
We utilized three distinct categories of gene sets to capture a broad range of biological insights:

* **Hallmark Gene Sets:** Represent specific, well-defined biological states or processes (e.g., hypoxia, inflammatory response) and provide a concise summary of the primary pathways involved.
* **KEGG Medicus:** A subset of the Kyoto Encyclopedia of Genes and Genomes (KEGG) that focuses on the molecular networks of diseases and drug mechanisms, offering a more clinical and mechanistic perspective.
* **GO Biological Process (GO:BP):** Comprehensive ontologies that describe the broader biological objectives (such as cell cycle or metabolic pathways) to which a gene contributes.

<!-- ### 2. Implementation and Mapping -->
<!-- The analysis was performed using "native" mouse gene sets (species: *Homo sapiens*) via the `msigdbr` package. To ensure high-confidence mapping across different data formats, gene sets were structured to include both: -->
<!-- * **Gene Symbols:** Standardized nomenclature for biological interpretation. -->
<!-- * **Ensembl IDs:** Stable identifiers for precise cross-referencing with the expression matrix. -->

This multi-faceted approach allows us to determine if the genes identified as "changing" or "stable" are non-randomly associated with specific biological functions or signaling pathways.

```{r functional analysis, message=FALSE, warning=FALSE}
# mouse native gene sets: db_species = "MM" (other option: human MSigDB with ortholog mapping to mouse)
geneset_collections <- list(
  Hallmark = msigdbr(species = "Homo sapiens", collection = "H") %>%
             dplyr::select(gs_name, gene_symbol, ensembl_gene),

  KEGG_Medicus = msigdbr(species = "Homo sapiens", collection = "C2") %>%
                 filter(gs_subcollection == "CP:KEGG_MEDICUS") %>%
                 dplyr::select(gs_name, gene_symbol, ensembl_gene),

  GO_Biological_Process = msigdbr(species = "Homo sapiens", collection = "C5") %>%
                          filter(gs_subcollection == "GO:BP") %>%
                          dplyr::select(gs_name, gene_symbol, ensembl_gene)
)
```

# Create gene annotation
```{r}
library(org.Hs.eg.db) # Or org.Mm.eg.db for mouse
library(AnnotationDbi)

# 1. Get your list of gene symbols from the object
current_symbols <- rownames(seu.integrated)

# 2. Query the database for IDs/Annotations
gene_info <- AnnotationDbi::select(org.Hs.eg.db,
                    keys = current_symbols,
                    columns = c("ENSEMBL", "GENENAME"),
                    keytype = "SYMBOL")

# 3. Clean up the query (handle potential duplicates/NAs)
# (Optional: remove duplicates if a symbol maps to multiple Ensembl IDs)
gene_info <- gene_info[!duplicated(gene_info$SYMBOL), ]
# keep non-NA only
gene_info <- gene_info[!is.na(gene_info$ENSEMBL),]
```

### Gene Set Enrichment Analysis (GSEA)

While Overrepresentation Analysis (ORA) treats gene lists as unranked sets, **Gene Set Enrichment Analysis (GSEA)** utilizes the full spectrum of gene expression changes. This allows for the detection of subtle but coordinated shifts in entire biological pathways that might be missed by simple threshold-based filtering.

#### 1. Analysis Workflow: Ranking and Running
For each experimental contrast, we performed the following steps:
* **Gene Ranking:** All genes were ranked by their $log_2(\text{Fold Change})$ in descending order. This ensures that the most upregulated genes are at the "top" of the list and the most downregulated genes are at the "bottom."
* **Enrichment Calculation:** Using the `clusterProfiler::GSEA` function, we calculated a running sum statistic to identify pathways where genes are non-randomly distributed at either end of the ranked list.

#### 2. Key Metrics and "Core Enrichment"
The results are summarized using two primary metrics:
* **Normalized Enrichment Score (NES):** A positive NES indicates the pathway is generally **activated** (enriched in upregulated genes), while a negative NES indicates the pathway is **suppressed** (enriched in downregulated genes).
* **Core Enrichment:** We extracted the specific subset of "leading-edge" genes—those that contribute most significantly to the enrichment score—and reported them along with their individual $log_2(\text{Fold Change})$ values.

#### 3. Visualization and Export
* **Summary Dotplots:** For each comparison, we visualized the top 10 activated and top 10 suppressed pathways. The size of the points represents the significance ($-\log_{10}(\text{Adjusted P-Value})$), and the x-axis indicates the NES.

#### Output file
Results are saved to GSEA_differential_expression_[geneset_name]_results.xlsx file, where each tab contains the results of one of the contrasts and only those genesets are listed for each contrast where the adjusted p value <= 0.05. If the enrichment process resulted no genesets, then the contrast is not shown in the file. If none of the genesets were significant, then the corresponding tab is empty.

```{r GSEA for differentially expressed genes, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
gsea_outer_list <- list()
p_multi_list <- list()
raw_gsea_models <- list()
 for (gs_name in names(geneset_collections)){
   current_t2g <- geneset_collections[[gs_name]]
   
    # add gene symbols
    id_map <- current_t2g %>%
    dplyr::select(ensembl_gene, gene_symbol) %>%
    distinct() %>%
      deframe() # Converts to a named vector
   
   # print("########################################")
   # print(paste("#   Current gene set:", gs_name, "    #"))
   # print("########################################")
   gsea_res_list <- list()
   # Iterate through your contrasts
   # for (contr in colnames(fite$coefficients)) {
   for (contr in c("EGFP_TumorVsSCP_Virus")) {
     # print(paste("Current contrast:", contr))

     # 1. Prepare the ranked list for GSEA
     full_stats <- topTable(fite, coef = contr, number = Inf)
     full_tbl <- as_tibble(full_stats, rownames="SYMBOL")
     # keep genes with ensembl id only
     full_tbl <- inner_join(gene_info, full_tbl, by="SYMBOL")

     # Create a named vector for GSEA ranking
     gene_list <- full_tbl$logFC
     names(gene_list) <- full_tbl$ENSEMBL
     # remove_duplicated
     keep <- which(!duplicated(names(gene_list)))
     gene_list <- gene_list[keep]
     

     # Create a lookup for the table mapping (clean Ensembl IDs)
     lfc_lookup <- full_stats$logFC
     names(lfc_lookup) <- sub("\\..*$", "", rownames(full_stats))

     # Sort for GSEA
     gene_list <- sort(gene_list, decreasing = TRUE)

     # 2. Run GSEA
     gsea_out <- clusterProfiler::GSEA(
       geneList = gene_list,
       TERM2GENE = current_t2g[, c("gs_name", "ensembl_gene")],
       pvalueCutoff = 0.05,
       verbose = FALSE
     )
     
     # store for later use
     raw_gsea_models[[gs_name]] <- gsea_out

     if (is.null(gsea_out) || nrow(as.data.frame(gsea_out)) == 0) {
       next
     }

     # 3. Process results and map Symbols + LogFC
     result_df <- as.data.frame(gsea_out)

     # 4. Update the mapping logic to include LogFC for Core Enrichment
     result_df_final <- result_df %>%
       mutate(MSigDB_Link = paste0("<a href='https://www.gsea-msigdb.org/gsea/msigdb/cards/",
                                   Description, "' target='_blank'>Link</a>")) %>%
       mutate(core_enrichment_stats = map_chr(core_enrichment, function(id_string) {
         # Split IDs (GSEA uses "/" as delimiter)
         ids <- unlist(strsplit(id_string, "/"))

         # Get Symbols from your id_map
         symbols <- id_map[ids]
         symbols[is.na(symbols)] <- ids[is.na(symbols)] # fallback to ID if symbol missing

         # Get LogFC values from our lookup
         # logfcs <- lfc_lookup[ids]
         logfcs <- lfc_lookup[symbols]

         # Format as "Symbol (LogFC)"
         formatted <- paste0(symbols, " (", round(logfcs, 2), ")")

         # Combine into a single string
         paste(formatted, collapse = ", ")
       })) %>%
       # Select columns for the final table
       dplyr::select(Description, NES, p.adjust, core_enrichment_stats, MSigDB_Link)

     # Format p-value to scientific
     result_df_final$p.adjust <- formatC(as.numeric(result_df_final$p.adjust), format = "e", digits = 2)
     result_df_final$NES <- formatC(as.numeric(result_df_final$NES), format = "e", digits = 2)

     # Add to list for Excel export
     gsea_res_list[[contr]] <- result_df_final

     # # 5. Create HTML Datatable (with the same JS logic for long text)
     # html_table <- datatable(result_df_final,
     #           escape = FALSE,
     #           filter = 'top',
     #           extensions = 'Buttons',
     #           options = list(
     #             pageLength = 10,
     #             dom = 'Bfrtip',
     #             buttons = c('copy', 'csv', 'excel'),
     #             columnDefs = list(list(
     #                 targets = 3, # Targets the core_enrichment_stats column
     #                 width = '500px',
     #                 render = JS(
     #                   "function(data, type, row, meta) {",
     #                   "return type === 'display' && data.length > 100 ?",
     #                   "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
     #                   "}"
     #                 )
     #             ))
     #           ),
     #           caption = paste("GSEA (Ranked by logFC):", contr))

     # Save individual widget
     #saveWidget(html_table, file.path(figfolder, paste0("GSEA_", contr, ".html")))

     # also add a figure
     if (!is.null(gsea_out) && nrow(as.data.frame(gsea_out)) > 0) {

     # 1. Identify the top 3 pathways by absolute NES
     # (You can change this to show the most significant or specific pathways of interest)
     top_pathways <- as.data.frame(gsea_out) %>%
       arrange(desc(abs(NES))) %>%
       slice_head(n = 3) %>%
       pull(ID)

     # # 2. Generate the plot for the top pathway
     # # p1: The Enrichment Score curve
     # # p2: The gene 'ticks' (hits)
     # # p3: The rank-ordered gene list gradient
     # p_gsea <- gseaplot2(gsea_out,
     #                     geneSetID = top_pathways[1],
     #                     title = paste("Top Pathway:", top_pathways[1]),
     #                     subplots = 1:3,
     #                     pvalue_table = TRUE)
     # 
     # # 3. Save the plot as a high-resolution PDF or PNG
     # # Since gseaplot2 returns a ggplot-compatible object, we use ggsave
     # plot_filename <- file.path(figfolder, paste0("GSEA_Plot_", contr, ".pdf"))
     # 
     # ggsave(plot_filename,
     #        plot = p_gsea,
     #        width = 10,
     #        height = 8,
     #        dpi = 300,
     #        bg = "white")

     # Optional: Plot multiple top pathways together
     p_multi <- enrichplot::gseaplot2(gsea_out,
                          geneSetID = top_pathways,
                          title = paste("Top 3 Pathways in", contr))
     p_multi_wrapped <- patchwork::wrap_plots(p_multi, ncol = 1, heights = c(2, 0.5, 0.5))
     # p_multi_list[[contr]] <- p_multi_wrapped
     subfolder <- file.path(Figures, gs_name)
     if (!file.exists(subfolder)) dir.create(subfolder)
     ggsave(file.path(subfolder, paste0("GSEA_MultiPlot_", contr, "_SCP_EGFP_tumor.pdf")),
            plot = p_multi, width = 12, height = 9)
     }
   }

   # Write out to Excel
   outfile_gsea <- file.path(Results, paste("GSEA_differential_expression_", gs_name, "_results_SCP_EGFP_tumor.xlsx", sep=""))
   write.xlsx(gsea_res_list, file = outfile_gsea, colNames = TRUE)
   gsea_outer_list[[gs_name]] <- gsea_res_list
 }
```


```{r dotplot of GSEA results, fig.width=20, fig.height=20}
for (gs_name in names(gsea_outer_list)){
  gsea_res_list <- gsea_outer_list[[gs_name]]
  plist <- list()
  for (gsi in names(gsea_res_list)){
    # create nice title:
    gsi_title <- switch(gsi,
      "TumorVsNormal"="Tumor vs Normal",
      gsi
    )
    clean_title <- paste(gs_name, gsi_title, sep = " - ")
    clean_title <- gsub("_", " ", clean_title)
    
    y <- gsea_res_list[[gsi]]
    
    # select activated and supressed genesets
    y$NES <- as.numeric(y$NES)
    y$p.adjust <- as.numeric(y$p.adjust)
    y$direction <- "activated"
    y$direction[y$NES<=0] <- "suppressed"
    numact <- length(which(y$direction=="activated"))
    numsupp <- length(which(y$direction=="suppressed"))

    # pick top and bottom genesets
    act_Nvals <- y$NES[y$direction=="activated"]
    act_thr <- sort(act_Nvals, decreasing = T)[min(10, numact)]
    act_keep <- which(y$direction=="activated" & y$NES >= act_thr)
    sup_Nvals <- y$NES[y$direction=="suppressed"]
    sup_thr <- sort(sup_Nvals, decreasing = F)[min(10, numsupp)]
    sup_keep <- which(y$direction=="suppressed" & y$NES <= sup_thr)
    y <- y[c(act_keep, sup_keep),]
    
    # sort by normalized enrichment
    y <- y[order(y$NES, decreasing = F),]
    # change the names of the gene sets
    #y$Description <- tolower(y$Description)
    #y$Description <- gsub(prefix[gsi], "", y$Description) # later if needed prefix <- c("gobp_", "kegg_", "hallmark_")
    y$Description <- gsub("_", " ", y$Description)
    y$Description <- str_to_sentence(y$Description)
    y$Description <- gsub("Gobp", "GO:BP", y$Description)
    y$Description <- gsub("dna", "DNA", y$Description)
    y$Description <- gsub("Dna", "DNA", y$Description)
    y$Description <- str_wrap(y$Description, width = 50) # width=30 is a good starting point for a 4-column layout
    #y$Description <- nice_names$name[match(y$Description, nice_names$compare)]
    y$size <- -log10(as.numeric(y$p.adjust))
    
    num_suppressed <- sum(y$direction == "suppressed")
    sep_line_pos <- num_suppressed + 0.5
    y$Description <- factor(y$Description, levels=y$Description)
      p <- ggplot(y, 
                    aes(x = NES, y = Description)) + 
               geom_point(aes(size = size, color = direction))+
               scale_color_manual(values = c("activated" = "firebrick", "suppressed" = "dodgerblue4")) +
               
              scale_size_continuous(
                 name = "-log10(p.adj)", 
                 limits = c(0, 20),      # <--- This is the fixed range
                 breaks = c(2, 4, 6, 8) # <--- Fixed circle sizes in legend
               ) +  
              
              # Only add it if both directions exist to avoid errors
             {if(num_suppressed > 0 && num_suppressed < nrow(y)) 
               geom_hline(yintercept = sep_line_pos, linetype = "solid", color = "gray80", size = 0.5)} +
              
               guides(size=guide_legend(title="-log(p.adjust)", ncol=1,byrow=TRUE,
                                title.vjust=0.9, order=1))+
               theme_bw(base_size = 16) +

               theme(
                 axis.text.y = element_text(size = 16, lineheight = 0.8),
                 panel.grid.minor = element_blank(),
                 legend.position="right",
                      legend.box = "vertical",
                     #axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
                     #strip.background =element_rect(fill=cfill),
                     strip.text = element_text(colour = 'black'),
                 legend.title = element_text(size = 16), # Legend title size
                  legend.text = element_text(size = 14)                 # Legend item text size)
               )
               ylab(NULL) + xlab("NES")+
               #scale_size(range=c(3, 8))+
               #scale_y_discrete(labels = function(x) str_wrap(x, width = 24))+
               geom_vline(xintercept = 0, linetype="dashed")+
               labs(title=clean_title)
      #plot(p)
      plist[[gsi]] <- p
    # figfile <- file.path(figfolder, paste("GSEA_", gs_name, "_", gsi, "_enrich_dotplot.pdf", sep=""))
    # ggsave(figfile, plot=p, width=14, height=6)
  }
  # plot together for the three genesets
  pp_combined <- patchwork::wrap_plots(plist, ncol = 1) + 
  patchwork::plot_layout(guides = "collect") & # Collects legends into one place
  theme(legend.position = "right")
  plot(pp_combined)
  ggsave(file.path(Figures, paste("GSEA_", gs_name, "_enrich_dotplot_combined_SCP_EGFP_tumor.pdf", sep="")), 
         plot = pp_combined, width = 10, height = 10)
}
```

