---
title: "Analysis of single cell neuroblastoma organoids: part 4 - reference mapping"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---
```{r setup, include=F}
knitr::opts_chunk$set(
	fig.height = 5,
	fig.width = 12,
	message = FALSE,
	warning = FALSE
)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).
The objective of this R Markdown code is to analyse the results obtained by the single cell RNA sequencing.
Follow steps in readme to create Python environment for scVI integration

```{r libraries}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

packages <- c(
  "Seurat",
  "ggplot2",
  "patchwork",
  "Matrix",
  "dplyr",
  "here",
  "stringr",
  "future",
  "scCustomize",
  "ggforce",
  "openxlsx",
  "reticulate",
  "FNN"
)

# TO DO: list packages in readme
# missing <- setdiff(packages, rownames(installed.packages()))
# 
# # # install CRAN packages via BioConductor as well
# if (length(missing) > 0) {
#   BiocManager::install(missing, ask = FALSE)
# }
invisible(lapply(packages, library, character.only = TRUE))
```


Folder structure
```{r}
Datafolder     <- here("Data")
# Subdirectories specifically isolated for this run's outputs
Figures   <- here(file.path("Figures", "reference_mapping"))
Data_calc <- here("Data_calculated")
Data_public <- here("Data_public")
Results   <- here("Results")

if (!file.exists(Figures))   dir.create(Figures, recursive = TRUE)
if (!file.exists(Data_calc)) dir.create(Data_calc, recursive = TRUE)
if (!file.exists(Results))   dir.create(Results, recursive = TRUE)

utils_file <- here("Rscripts","utils.R")
if (file.exists(utils_file)) source(utils_file)
```


Config parameters and marker lists
```{r}
config_file <- here("Rscripts","config.R")
source(config_file)

# for the same figures
set.seed(734)
```



Read in sample and reference data
```{r}
# creation of reference atlas
atlas_file <- file.path(Data_calc, "NB_reference_atlas.rds")
if (!file.exists(atlas_file)){
    
  # ref_info is loaded from config.R
  
  ref_names <- names(ref_info)
  
  refs <- lapply(ref_names, function(ref_name){
  
    message("Loading ", ref_name)
  
    obj <- readRDS(file.path(Data_public, ref_info[[ref_name]]$path))
    obj <- UpdateSeuratObject(obj)
  
    obj$reference_source <- ref_name
  
    obj
  })
  
  names(refs) <- ref_names 
  
  # list possibilities
  for (ri in 1:length(refs)){
    print(ref_names[ri])
    print(unique(refs[[ri]]@meta.data[[ref_info[[ri]]$label_col]]))
    print("")
  }
  
  # # create harmonized annotation
  # common_labels <- c(
  #   "tumor", "SCP", "bridge", "CPC", "sympathoblast",
  #   "chromaffin",  "prolif",  "immune", "mesenchymal",
  #   "endothelial", "cortex", "kidney",   "liver",
  #   "melanocyte",    "erythroid",   "HSC",
  #   "unknown" ) # fallback
  
  # recode references
  refs[["Jansky"]]$celltype_harmonized <- recode_labels(
    refs[["Jansky"]]$celltype,
    c(
      "Tumor cells" = "tumor",
      "Immune cells" = "other",
      "Mesenchymal cells" = "mesenchymal",
      "Endothelial cells" = "endothelial",
      "Schwann cells" = "SCP",
      "Liver cells" = "other"
    )
  )
  
  refs[["Kameneva"]]$celltype_harmonized <- recode_labels(
    refs[["Kameneva"]]$fate2,
    c(
      "SCP" = "SCP",
      "sympathoblasts" = "sympathoblast",
      "chromaffin" = "chromaffin",
      "cortex" = "other",
      "HSC_and_immune" = "other",
      "erythroid" = "other",
      "mesenchyme" = "mesenchymal",
      "endothelium" = "endothelial",
      "kidney" = "other",
      "liver" = "other",
      "melanocytes" = "other",
      "intermediate_mesoderm" = "mesenchymal"
    )
  )
  
  
  refs[["adr"]]$celltype_harmonized <- recode_labels(
    refs[["adr"]]$Annotation,
    c(
      "SCPs" = "SCP",
      "Bridge" = "bridge",
      "Sympathoblastic" = "sympathoblast",
      "Chromaffin" = "chromaffin",
      "Cortex" = "other",
      "Leukocytes" = "other",
      "Mesenchyme" = "mesenchymal",
      "Endothelium" = "endothelial",
      "Erythroblasts" = "other",
      "Other" = "other"
    )
  )
      
  refs[["GOSH"]]$celltype_harmonized <- recode_labels(
    refs[["GOSH"]]$Annotation,
    c(
      "Tumour cluster 1" = "tumor",
      "Tumour cluster 2" = "tumor",
      "Tumour cluster 3" = "tumor",
      "Leukocytes" = "other",
      "Mesenchyme" = "mesenchymal",
      "Endothelium" = "endothelial"
    )
  )    
      
  refs[["PMC"]]$celltype_harmonized <- recode_labels(
    refs[["PMC"]]$Annotation,
    c(
      "Tumour cluster 1" = "tumor",
      "Tumour cluster 2" = "tumor",
      "Tumour cluster 3" = "tumor",
      "Schwannian stroma" = "SCP",
      "Mesenchyme" = "mesenchymal",
      "Leukocytes" = "other",
      "Endothelium" = "endothelial",
      "Ambiguous 1" = "other",
      "Ambiguous 2" = "other"
    )
  )
  
  
  refs[["Saldana"]]$celltype_harmonized <- recode_labels(
    refs[["Saldana"]]$type3,
    c(
      "SCP" = "SCP",
      "sympathoblasts" = "sympathoblast",
      "chromaffin" = "chromaffin",
      "mesenchyme" = "mesenchymal",
      "other" = "other"
    )
  )
  
  refs[["vanHaver"]]$celltype_harmonized <- recode_labels(
    refs[["vanHaver"]]$seurat_clusters,
    c(
      "0" = "SCP",
      "9" = "SCP",
      "1" = "bridge",
      "10" = "bridge",
      "4" = "CPC",
      "8" = "CPC",
      "6" = "CPC",
      "2" = "CPC",
      "7" = "prolif",
      "3" = "sympathoblast",
      "5" = "sympathoblast"
    )
  )
  
  # sanity check:
  lapply(refs, function(x) {
    sort(table(x$celltype_harmonized), decreasing = TRUE)
  })
  
  # check how much gets lost
  lapply(refs, function(x) {
    mean(x$celltype_harmonized == "other")
  })
  
  # for (rname in names(refs)){
  #   # there are a huge number of uknown -> remove
  #   refs[[rname]] <- subset(
  #     refs[[rname]],
  #     subset = !celltype_harmonized %in% c("other")
  #   )
  # }
  
  # downsample huge populations
  for(i in seq_along(refs)){
  
    Idents(refs[[i]]) <- "celltype_harmonized"
    
    print(table(refs[[i]]$celltype_harmonized))
    
    refs[[i]] <- subset(
      refs[[i]],
      downsample = 500
    )
    
    # after downsampling
    print("after downsampling:")
    print(table(refs[[i]]$celltype_harmonized))
    print("===")
  }
  
  # ---------------------------------------------------------------------------
  # Create harmonized, transformed, downsampled reference files
  #
  # These are saved separately for downstream analyses.
  # They are NOT used to construct the merged reference atlas below.
  # ---------------------------------------------------------------------------
  
  message("Creating harmonized, transformed, downsampled reference files...")
  
  for (ref_name in names(refs)) {
  
    message("Transforming ", ref_name)
    
    ref_obj <- refs[[ref_name]]
    
    ref_obj <- SCTransform(
      ref_obj,
      vst.flavor = "v2"
    )
    
    ref_obj <- RunPCA(
      ref_obj
    )
    
    ref_obj <- RunUMAP(
      ref_obj,
      dims = 1:30,
      reduction = "pca",
      return.model = TRUE
    )
  
    SaveSeuratRds(
      ref_obj,
      file = file.path(
        Data_calc,
        paste0(
          "ref_",
          ref_name,
          "_harmonized_transformed_downsampled.rds"
        )
      )
    )
  }
  
  # ---------------------------------------------------------------------------
  # Create the merged reference atlas
  #
  # The atlas itself uses one consistent workflow across all references:
  # RNA normalization -> HVG selection -> scaling -> PCA -> UMAP.
  #
  # The individually SCTransformed reference objects above are not used here.
  # ---------------------------------------------------------------------------
  
  message("Merging reference datasets")
  
  merged.ref <- merge(
    refs[[1]],
    y = refs[-1]
  )
  
  # no need to join layers, since earlier seurat type data
  DefaultAssay(merged.ref) <- "RNA"
  
  print(table(merged.ref$celltype_harmonized))
  
  # normalize merged reference and identify HVGs
  message("Normalizing merged reference and identifying HVGs...")
  
  merged.ref <- NormalizeData(
    merged.ref
  )
  
  merged.ref <- FindVariableFeatures(
    merged.ref,
    nfeatures = 3000
  )
  
  hvg <- VariableFeatures(merged.ref)
  
  # This will be needed for the anchor step
  message("Scaling data and running Reference PCA...")
  
  merged.ref <- ScaleData(
    merged.ref,
    features = hvg,
    verbose = FALSE
  )
  
  merged.ref <- RunPCA(
    merged.ref,
    npcs = 30,
    features = hvg,
    verbose = FALSE
  )
  
  # ---------------------------------------------------------------------------
  # Build the reference UMAP model from the same PCA representation
  # used for the downstream anchor workflow.
  # ---------------------------------------------------------------------------
  
  message("Building reference UMAP model...")
  
  merged.ref <- RunUMAP(
    merged.ref,
    reduction = "pca",
    dims = 1:30,
    n.neighbors = 30L,
    metric = "cosine",
    min.dist = 0.3,
    return.model = TRUE,
    reduction.name = "umap",
    reduction.key = "UMAP_",
    verbose = FALSE
  )
  
  # QC plot
  p1 <- DimPlot(
    merged.ref,
    reduction = "umap",
    group.by = "reference_source"
  )
  
  p2 <- DimPlot(
    merged.ref,
    reduction = "umap",
    group.by = "celltype_harmonized",
    label = TRUE,
    cols = atlas_palette
  )
  
  combined_plot <- p1 | p2
  
  plot(combined_plot)
  
  ggsave(
    file.path(
      Figures,
      paste("Atlas", ".pdf", sep = "")
    ),
    combined_plot,
    width = 10,
    height = 6
  )
  
  # Run a quick loop across the references to print an identity matrix
  for(name in names(ref_info)) {
    
    cat("\n=============================================\n")
    cat(
      "  HARMONIZATION AUDIT TRAIL FOR:",
      toupper(name),
      "\n"
    )
    cat("=============================================\n")
    
    orig_col <- ref_info[[name]]$label_col
    
    # Filter the merged atlas metadata for just this specific source
    sub_meta <- merged.ref@meta.data[
      merged.ref$reference_source == name,
    ]
    
    # Print the cross-tabulation table
    print(
      table(
        Original_Label = sub_meta[[orig_col]],
        Harmonized_Label = sub_meta$celltype_harmonized
      )
    )
  }
  
  # finally save the atlas!
  SaveSeuratRds(
    merged.ref,
    file = atlas_file
  )
  
} else {
  
  print("Atlas already saved, load saved object")
  
  merged.ref <- LoadSeuratRds(atlas_file)
  
  # Re-attach the uwot model, since it does not survive serialization
  merged.ref[["umap"]]@misc$model <- uwot::load_uwot(
    file.path(Data_calc, "atlas_umap_model")
  )
  
} # end of reference atlas creating
```


<!-- ```{r scvi mapping} -->
<!-- # now the mapping -->
<!-- ct <- "EpiC" -->
<!-- # ct <- "kolf2" -->

<!-- # outfile <- file.path(Data_calc, paste("seu_integrated_clusters_", ct, "_all_nHVG", nHVG_str, ".rds", sep="")) -->
<!-- # use the new cluster names -->
<!-- outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep="")) -->
<!-- seu.integrated <- LoadSeuratRds(outfile) -->
<!-- # for the nice plots, remove "type_12" cluster -->
<!-- seu.integrated <- subset(seu.integrated, subset = !cluster_type %in% c("type_12")) -->
<!-- seu.integrated$cluster_type <- droplevels(seu.integrated$cluster_type) -->
<!-- # write out cell numbers -->
<!-- table(seu.integrated$sample.names, seu.integrated$cluster_type) -->


<!-- # mapping the latent space as well -->
<!-- # ---- Embed the query into the reference's scVI latent space (scArches surgery) ---- -->

<!-- reticulate::use_virtualenv(here("scvi-env"), required = TRUE) -->
<!-- scvi <- import("scvi") -->
<!-- sp   <- import("scipy.sparse") -->
<!-- ad   <- import("anndata") -->

<!-- scvi_model_dir <- file.path(Data_calc, "scvi_reference_model") -->
<!-- hvg <- readRDS(file.path(Data_calc, "scvi_reference_hvg.rds")) -->

<!-- DefaultAssay(seu.integrated) <- "RNA" -->
<!-- query_counts <- GetAssayData(seu.integrated, assay = "RNA", layer = "counts") -->

<!-- # scVI needs exactly the same gene set/order used for training the reference -->
<!-- missing_genes <- setdiff(hvg, rownames(query_counts)) -->
<!-- if (length(missing_genes) > 0) { -->
<!--   message(length(missing_genes), " reference HVGs missing in query; padding with zeros") -->
<!--   pad <- Matrix::Matrix(0, nrow = length(missing_genes), ncol = ncol(query_counts), sparse = TRUE) -->
<!--   rownames(pad) <- missing_genes -->
<!--   query_counts <- rbind(query_counts, pad) -->
<!-- } -->
<!-- query_counts <- query_counts[hvg, ] -->
<!-- query_counts@x <- round(query_counts@x)   # raw counts required -->

<!-- query_meta <- seu.integrated@meta.data -->
<!-- query_meta$reference_source <- "query"    # unseen batch label — required by scVI's setup -->

<!-- adata_query <- ad$AnnData( -->
<!--   X   = sp$csr_matrix(t(query_counts)), -->
<!--   obs = query_meta, -->
<!--   var = data.frame(gene_names = hvg, row.names = hvg) -->
<!-- ) -->

<!-- # Load reference model, register the query, fine-tune only the query-specific -->
<!-- # weights while the reference latent space stays fixed — this is what makes -->
<!-- # query and reference land in the SAME coordinate system -->
<!-- scvi$model$SCVI$prepare_query_anndata(adata_query, scvi_model_dir) -->
<!-- query_model <- scvi$model$SCVI$load_query_data(adata_query, scvi_model_dir) -->
<!-- query_model$train(max_epochs = as.integer(200), plan_kwargs = list(weight_decay = 0.0)) -->

<!-- latent_query <- as.matrix(query_model$get_latent_representation()) -->
<!-- colnames(latent_query) <- paste0("scviref_", seq_len(ncol(latent_query))) -->
<!-- rownames(latent_query) <- colnames(seu.integrated) -->

<!-- seu.integrated[["scvi.ref"]] <- CreateDimReducObject( -->
<!--   embeddings = latent_query, key = "scviref_", assay = "RNA" -->
<!-- ) -->

<!-- # ---- Transfer labels by k-NN in the shared scVI latent space ---- -->
<!-- # (replaces FindTransferAnchors/MapQuery, which assumes a linear reduction) -->

<!-- ref_latent <- Embeddings(merged.ref, "scvi") -->
<!-- ref_labels <- factor(merged.ref$celltype_harmonized) -->

<!-- k <- 30 -->
<!-- knn <- FNN::get.knnx(data = ref_latent, query = latent_query, k = k) -->

<!-- vote_frac <- t(apply(knn$nn.index, 1, function(idx) { -->
<!--   prop.table(table(ref_labels[idx])) -->
<!-- })) -->
<!-- colnames(vote_frac) <- paste0("vote_", colnames(vote_frac)) -->

<!-- seu.integrated$atlas_pred_celltype <- gsub( -->
<!--   "^vote_", "", colnames(vote_frac)[max.col(vote_frac, ties.method = "first")] -->
<!-- ) -->
<!-- seu.integrated$atlas_pred_score <- apply(vote_frac, 1, max) -->
<!-- seu.integrated <- AddMetaData(seu.integrated, as.data.frame(vote_frac)) -->


<!-- # ---- Optional: project query onto the reference's existing scVI-UMAP for plotting ---- -->
<!-- # # Requires the reference UMAP (built on "scvi") to have been created with return.model = TRUE -->
<!-- seu.integrated <- RunUMAP( -->
<!--   seu.integrated, -->
<!--   reduction       = "scvi.ref",           # the query embedding already stored in the object -->
<!--   dims            = 1:ncol(latent_query), # 1:30, matching the scVI latent dimensionality -->
<!--   reduction.model = merged.ref[["umap"]], # the scVI-based reference UMAP model -->
<!--   reduction.name  = "ref.umap", -->
<!--   reduction.key   = "refUMAP_" -->
<!-- ) -->

<!-- # end of mapping -->

<!-- # ---- Audit of prediction performance ---- -->
<!-- # (atlas_pred_celltype, atlas_pred_score, and vote_* columns were already -->
<!-- # added to seu.integrated@meta.data by the k-NN voting step above) -->
<!-- vote_cols <- grep("^vote_", colnames(seu.integrated@meta.data), value = TRUE) -->
<!-- voting_matrix <- seu.integrated@meta.data[, c("atlas_pred_celltype", "atlas_pred_score", vote_cols)] -->
<!-- head(voting_matrix) -->

<!-- # ---- Plot confidence ---- -->
<!-- p_vln <- VlnPlot( -->
<!--   seu.integrated, features = "atlas_pred_score", -->
<!--   group.by = "atlas_pred_celltype", pt.size = 0.1, -->
<!--   cols = atlas_palette -->
<!-- ) + -->
<!--   geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") + -->
<!--   theme(axis.text.x = element_text(angle = 45, hjust = 1)) + -->
<!--   ggtitle("Label Transfer Confidence by Cell Type") -->

<!-- ggsave( -->
<!--   file.path(Figures, paste("Label_transfer_confidence_scvi_", ct, ".pdf", sep="")), -->
<!--   plot = p_vln, width = 10, height = 6 -->
<!-- ) -->

<!-- # ---- PLOT A: atlas labels on the query's own UMAP layout ---- -->
<!-- # NOTE: "umap" here is the per-cell-type scVI UMAP from Part 2, not the reference mapping -->
<!-- p_query_umap <- DimPlot( -->
<!--   seu.integrated, -->
<!--   reduction = "umap", -->
<!--   group.by = "atlas_pred_celltype", -->
<!--   cols = atlas_palette, -->
<!--   label = FALSE, label.size = 5, repel = TRUE -->
<!-- ) + -->
<!--   ggtitle("Query UMAP colored by Atlas Predictions") + -->
<!--   theme(plot.title = element_text(hjust = 0.5, face = "bold")) -->
<!-- print(p_query_umap) -->

<!-- # ---- PLOT B: query cells projected onto the reference atlas UMAP ---- -->
<!-- p_projected_umap <- DimPlot( -->
<!--   seu.integrated, -->
<!--   reduction = "ref.umap", -->
<!--   group.by = "atlas_pred_celltype", -->
<!--   cols = atlas_palette, -->
<!--   label = FALSE, label.size = 5, repel = TRUE -->
<!-- ) + -->
<!--   ggtitle("Query Cells Projected onto Reference Atlas UMAP") + -->
<!--   theme(plot.title = element_text(hjust = 0.5, face = "bold")) -->
<!-- print(p_projected_umap) -->

<!-- # ---- Save both plots ---- -->
<!-- ggsave( -->
<!--   file.path(Figures, paste("Query_UMAP_with_Atlas_Labels_scvi_", ct, ".pdf", sep="")), -->
<!--   plot = p_query_umap, width = 10, height = 6 -->
<!-- ) -->
<!-- ggsave( -->
<!--   file.path(Figures, paste("Query_Projected_onto_Atlas_UMAP_scvi_", ct, ".pdf", sep="")), -->
<!--   plot = p_projected_umap, width = 10, height = 6 -->
<!-- ) -->


<!-- ``` -->



```{r seurat mapping}
# now the mapping
ct <- "EpiC"
# ct <- "kolf2"
  
# outfile <- file.path(Data_calc, paste("seu_integrated_clusters_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
# use the new cluster names
outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
seu.integrated <- LoadSeuratRds(outfile)
# for the nice plots, remove "type_12" cluster
seu.integrated <- subset(seu.integrated, subset = !cluster_type %in% c("type_12"))
seu.integrated$cluster_type <- droplevels(seu.integrated$cluster_type)
# write out cell numbers
table(seu.integrated$sample.names, seu.integrated$cluster_type)

# Switch query to RNA assay and ensure it is log-normalized
# (This is safe, fast, and does not alter your query's SCT integration embeddings)
DefaultAssay(seu.integrated) <- "RNA"
seu.integrated <- NormalizeData(seu.integrated, verbose = FALSE)

# 2. Find Transfer Anchors using LogNormalize matching your upstream data
message("Finding transfer anchors...")
anchors <- FindTransferAnchors(
  reference = merged.ref,
  query = seu.integrated,
  normalization.method = "LogNormalize", # <-- Use LogNormalize instead of SCT
  reduction = "pcaproject",               # Projects query onto ref PCA space (ultra-fast)
  reference.reduction = "pca",
  dims = 1:30
)

# 4. Map Query to Reference
message("Mapping query to reference...")
seu.integrated <- MapQuery(
  anchorset = anchors,
  reference = merged.ref,
  query = seu.integrated,
  refdata = list(predicted_celltype = "celltype_harmonized"),
  reference.reduction = "pca",
  reduction.model = "umap"
)

# 1. Clean up and store the transferred metadata under permanent names
seu.integrated$atlas_pred_celltype <- seu.integrated$predicted.predicted_celltype
seu.integrated$atlas_pred_score    <- seu.integrated$predicted.predicted_celltype.score

# audit of prediction performance
score_columns <- grep("predicted_celltype.score.", colnames(seu.integrated@meta.data), value = TRUE)
voting_matrix <- seu.integrated@meta.data[, c("atlas_pred_celltype", "atlas_pred_score", score_columns)]
colnames(voting_matrix) <- gsub("predicted_celltype.score.", "vote_", colnames(voting_matrix))
head(voting_matrix)

# plot confidence
p_vln <- VlnPlot(seu.integrated,  features = "atlas_pred_score",
  group.by = "atlas_pred_celltype", pt.size = 0.1,
  cols = atlas_palette
) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Label Transfer Confidence by Cell Type")

ggsave(
  file.path(Figures, paste("Label_transfer_confidence_", ct, ".pdf", sep="")),
  plot = p_vln,
  width = 10, height = 6
)

# 2. PLOT A: Visualizing atlas labels on your QUERY's native UMAP layout
p_query_umap <- DimPlot(
  seu.integrated,
  reduction = "umap", # Uses your query's own UMAP coordinates
  group.by = "atlas_pred_celltype",
  cols = atlas_palette,
  #label = TRUE,
  label=F,
  label.size = 5,
  repel = TRUE
) +
  ggtitle("Query UMAP colored by Atlas Predictions") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(p_query_umap)

# 3. PLOT B: Visualizing query cells projected onto the REFERENCE UMAP space
p_projected_umap <- DimPlot(
  seu.integrated,
  reduction = "ref.umap", # Uses the reference atlas UMAP coordinates calculated by MapQuery
  group.by = "atlas_pred_celltype",
  cols = atlas_palette,
  #label = TRUE,
  label=F,
  label.size = 5,
  repel = TRUE
) +
  ggtitle("Query Cells Projected onto Reference Atlas UMAP") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(p_projected_umap)


# 4. Save both plots cleanly
ggsave(
  file.path(Figures, paste("Query_UMAP_with_Atlas_Labels_", ct, ".pdf", sep="")),
  plot = p_query_umap,
  width = 10, height = 6
)

ggsave(
  file.path(Figures, paste("Query_Projected_onto_Atlas_UMAP_", ct, ".pdf", sep="")),
  plot = p_projected_umap,
  width = 10, height = 6
)

```


### Annotate clusters based on reference majority vote
```{r annotation cluster-wide}
message("Aggregating cell-level predictions to cluster-level annotations...")

# 1. Generate a cross-tabulation table (Clusters vs. Atlas Predictions)
cluster_vs_pred <- table(seu.integrated$cluster_type, seu.integrated$atlas_pred_celltype)

cat("\n--- Cross-tabulation Matrix ---\n")
print(cluster_vs_pred)

# 2. Extract the majority vote for each cluster using dplyr (already in your libraries)
majority_votes <- as.data.frame(cluster_vs_pred) %>%
  rename(cluster = Var1, predicted_celltype = Var2, cell_count = Freq) %>%
  group_by(cluster) %>%
  # Find the cell type with the highest count in each cluster
  slice_max(order_by = cell_count, n = 1, with_ties = FALSE) %>% 
  ungroup()

cat("\n--- Assigned Majority Votes per Cluster ---\n")
print(majority_votes)

# 3. Map these consensus labels back into the Seurat object metadata
# Create a named vector for quick mapping
annotation_vector <- majority_votes$predicted_celltype
names(annotation_vector) <- majority_votes$cluster

# Assign the new labels to a distinct metadata column so you don't lose original clusters
seu.integrated$cluster_annotated <- unname(annotation_vector[as.character(seu.integrated$cluster_type)])

# Also set them as the active identities for easy plotting
Idents(seu.integrated) <- "cluster_annotated"

# 4. Plot the newly annotated clusters on your query layout
p_cluster_annotated <- DimPlot(
  seu.integrated, 
  reduction = "umap", 
  #label = TRUE, 
  label = F, 
  label.size = 5, 
  repel = TRUE,
  cols = atlas_palette # Reuses your defined palette
) + 
  ggtitle("Query Clusters Annotated by Atlas Majority Vote") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(p_cluster_annotated)

# 5. Save the clean consensus plot
ggsave(
  file.path(Figures, paste("Query_UMAP_Cluster_Annotated_", ct, ".pdf", sep="")), 
  plot = p_cluster_annotated, 
  width = 10, height = 6
)

# Optional: Save the newly annotated object back to your results folder
# SaveSeuratRds(seu.integrated, file = outfile)
```



Annotate by each reference separately
```{r}

  outfile_new <- file.path(Data_calc, paste("seu_integrated_clusters_newnames_refscores_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
  if (!file.exists(outfile_new)){
        # 2. Set memory options (as you did before)
  options(future.globals.maxSize = 50000 * 1024^2)
  plan("sequential")
  
  # 3. Start the loop
  ref_objects <- list()
  for (ref_name in names(ref_info)){
    
    message(paste("--- Processing Reference:", ref_name, "---"))
    # check if the anchoring was already done:
    if (paste0("pred.", ref_name) %in% names(seu.integrated@meta.data)){
      print("reference already processed and added")
    } else {
      ref_file <-  file.path(Data_calc, 
                      paste0("ref_", ref_name, "_harmonized_transformed_downsampled.rds"))
      if (file.exists(ref_file)){
        ref_obj <- LoadSeuratRds(file=ref_file)
      } else {
        stop("Reference file does not exist: ", ref_file)
      }
      
      DefaultAssay(ref_obj) <- "RNA"
      ref_obj <- NormalizeData(ref_obj, verbose = FALSE)
      ref_obj <- FindVariableFeatures(ref_obj, nfeatures = 2000, verbose = FALSE)
      ref_obj <- ScaleData(ref_obj, verbose = FALSE)
      ref_obj <- RunPCA(ref_obj, npcs = 30, verbose = FALSE)
      
      # B. Find Anchors
      anchors <- FindTransferAnchors(
        reference = ref_obj,
        query = seu.integrated,
        #normalization.method = "SCT",
        normalization.method = "LogNormalize",
        reference.reduction = "pca",
        dims = 1:30
      )
      
      # C. Map Query and Transfer Labels
      # We use the ref_name to create unique metadata columns in your query object
      seu.integrated <- MapQuery(
        anchorset = anchors,
        reference = ref_obj,
        query = seu.integrated,
        refdata = list(prediction = "celltype_harmonized"),
        reference.reduction = "pca", 
        reduction.model = "umap"
      )
      
      # D. Rename the transferred metadata columns so they don't get overwritten in the next loop
      # This stores them as "pred.adr", "pred.GOSH", etc.
      seu.integrated[[paste0("pred.", ref_name)]] <- seu.integrated$predicted.prediction
      seu.integrated[[paste0("score.", ref_name)]] <- seu.integrated$predicted.prediction.score
      
      ref_objects[[ref_name]] <- ref_obj
      
      # E. Clean up memory before next dataset
      rm(ref_obj, anchors)
      gc()
    } # end of if/else of processing reference
    
  } # enf of loop ref_name in names(ref_info)
  
  # new save the new seurat object
  SaveSeuratRds(seu.integrated, file = outfile_new)
  } else {
    seu.integrated <- LoadSeuratRds(outfile_new)
  }
  
  
  # plot all of them - TO DO - do it automatically
  cols_to_plot <- paste("pred.", names(ref_info), sep="")
  
  # calculate confidence of each score:
  summary_list <- list()
  for (ref in cols_to_plot) {
    rscore <- gsub("pred.", "score.", ref)
    rqual <- gsub("score.", "qual.", rscore)
    seu.integrated[[rqual]] <- ifelse(
    seu.integrated[[rscore]] >= 0.5, 
    "High (>0.5)", 
    "Low (<0.5)"
       )
    
    # p_scvi <- DimPlot(seu.integrated,
    #                   reduction = "umap",
    #                   group.by = ref,
    #                   split.by = rqual,
    #                   label = TRUE,
    #                   repel = TRUE) +
    #   ggtitle(paste("SCVI UMAP: ",ref, " Labels",sep=""))
    # print(p_scvi)
    
  counts <- table(seu.integrated[[rqual]])
  percent <- round(100 * prop.table(counts), 2)

  summary_list[[ref]] <- data.frame(
    Classification = names(counts),
    Count = as.numeric(counts),
    Percentage = as.numeric(percent)
  )  
}
  
  # 2. Create a list of plots
  plot_list <- list()
  for (ref in cols_to_plot) {
    print(ref)
    print(table(seu.integrated[[ref]]))
    rscore <- gsub("pred.", "score.", ref)
    rqual <- gsub("score.", "qual.", rscore)
    # replace label of low quality plots
    ref_vector    <- seu.integrated[[ref]][, 1]
    rscore_vector <- seu.integrated[[rscore]][, 1]
    seu.integrated[[ref]] <- ifelse(rscore_vector >= 0.5, ref_vector, "low (<0.5)")
    print(table(seu.integrated[[ref]]))
    # Check if the column exists to avoid errors
    if (ref %in% colnames(seu.integrated@meta.data)) {
      
      p <- DimPlot(seu.integrated, 
                   reduction = "umap", 
                   group.by = ref, 
                  cols = atlas_palette,
                   # split.by = rqual,
                   label = F, 
                   label.size = 6, 
                   repel = TRUE) + 
        ggtitle(paste("Reference:", gsub("pred.", "", ref))) +
        # NoLegend() + # Removing legends to keep it clean; labels are on the clusters
        theme(plot.title = element_text(hjust = 0.5, face = "bold"))
      
      plot_list[[ref]] <- p
      # also save
      ggsave(file.path(Figures, paste("Dimplot_ref_", ref, "_", ct,
                    "_all_nHVG", nHVG_str,".pdf", sep="")), p, width = 10,   height = 6)
    } # end of if
  } # loop for different reference data
  
  # repeat loop for reference data with different days
  day_options <- c("D26_both", "D42_both", "D26_WT", "D26_STOP_MYCN",
                   "D42_novirus", "D42_hTH", "all")

  for (day in day_options){
      seu.day <- subset_seu_by_day(seu.integrated, day)
    
      # drop unused levels - do it everywhere!!!
      cluster_sizes <- table(seu.day$cluster_type)
      keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
      seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
      if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
        seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
      }
    
      # set up the order
      bio_order <- bio_order_list[[ct]]
      seu.day$cluster_type <- droplevels(seu.day$cluster_type)
   
      # use new names
      actual_day <- limma::strsplit2(day, "_")[[1]]
      
      # Rename clusters from previous annotation
      if (actual_day %in% names(cluster_names_dict[[ct]])){
            new_cluster_ids <- dplyr::recode(
          as.character(seu.day$cluster_type),
          !!!cluster_names_dict[[ct]][[actual_day]]
        )
        # Preserve cluster order
        renamed_cluster_order <- unique(dplyr::recode(
          bio_order,
          !!!cluster_names_dict[[ct]][[actual_day]]
        ))
        
        new_cluster_ids <- factor(
          new_cluster_ids,
          levels = renamed_cluster_order
        )
        
        # Store back into Seurat object and construct final legend text
        seu.day$cluster_type_renamed <- new_cluster_ids
        Idents(seu.day) <- "cluster_type_renamed"
      }
      
      # drop unused levels
      seu.day$cluster_type_renamed <- droplevels(seu.day$cluster_type_renamed)
      print(table(seu.day$cluster_type_renamed))
        
      # Print out
      cat("\nProcessing Day:", day, "\n")
      print(table(seu.day$sample.names))
      
      # now the loop for different reference types
      for (ref in cols_to_plot) {
        print(ref)
        print(table(seu.day[[ref]]))
        rscore <- gsub("pred.", "score.", ref)
        rqual <- gsub("score.", "qual.", rscore)
        # replace label of low quality plots
        ref_vector    <- seu.day[[ref]][, 1]
        rscore_vector <- seu.day[[rscore]][, 1]
        seu.day[[ref]] <- ifelse(rscore_vector >= 0.5, ref_vector, "low (<0.5)")
        print(table(seu.day[[ref]]))
        # Check if the column exists to avoid errors
        if (ref %in% colnames(seu.day@meta.data)) {
          
          p <- DimPlot(seu.day, 
                       reduction = "umap", 
                       group.by = ref, 
                      cols = atlas_palette,
                       # split.by = rqual,
                       label = F, 
                       label.size = 6, 
                       repel = TRUE) + 
            ggtitle(paste("Reference:", gsub("pred.", "", ref))) +
            # NoLegend() + # Removing legends to keep it clean; labels are on the clusters
            theme(plot.title = element_text(hjust = 0.5, face = "bold"))
          
          plot_list[[ref]] <- p
          # also save
          ggsave(file.path(Figures, paste("Dimplot_ref_", ref, "_", ct,
                "_day_", day, "_nHVG", nHVG_str,".pdf", sep="")), p, width = 10,   height = 6)
        } # end of if
      } # loop for different reference data
      
  }
  

  
  
#} # loop for celltype

```



Session info for reproducibility - run once after all rmd files had been run
```{r session info}
sessionInfo()
```


