---
title: "Analysis of single cell neuroblastoma organoids: part 6b - functional analysis for specific clusters"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---
```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).
The objective of this R Markdown code is to analyse the results obtained by the single cell RNA sequencing.

```{r libraries}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
packages <- c(
  "knitr",
  "Seurat",
  "ggplot2",
  "patchwork",
  "dplyr",
  "tidyr",
  "tibble",
  "here",
  "stringr",
  "openxlsx",
  "limma",
  "edgeR",
  "msigdbr",
  "clusterProfiler",
  "enrichplot",
  "EnhancedVolcano",
  "pathview",
  "AnnotationDbi",
  "org.Hs.eg.db",
  "enrichplot"
)
invisible(lapply(packages, library, character.only = TRUE))
```


Folder structure
```{r}
Datafolder     <- here("Data")
Data_calc <- here("Data_calculated") # Path to find Part 1 and 2 outputs

# Subdirectories specifically isolated for this run's outputs
Figures   <- here(file.path("Figures", "pseudo_bulk_SCP_EGFP_tumor"))
Data_calc <- here("Data_calculated")
Data_public <- here("Data_public")
Results   <- here("Results")

if (!file.exists(Figures))   dir.create(Figures, recursive = TRUE)
if (!file.exists(Data_calc)) dir.create(Data_calc, recursive = TRUE)
if (!file.exists(Results))   dir.create(Results, recursive = TRUE)

utils_file <- here("Rscripts","utils.R")
if (file.exists(utils_file)) source(utils_file)
```

Config parameters and marker lists
```{r}
config_file <- here("Rscripts","config.R")
source(config_file)

# for the same figures
set.seed(734)
```


Read in sample and reference data
Select all clusters that will be uesd for the DE analysis
```{r selection of specific clusters}
pseudo_bulk <- list()
for (ct in c("EpiC", "kolf2")) {
  # 1. Load Data
  outfile <- file.path(Results, paste0("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds"))
  seu.integrated <- LoadSeuratRds(outfile)
  day <- "D42_both"
  seu.day <- subset_seu_by_day(seu.integrated, day)
  
  cluster_sizes <- table(seu.day$cluster_type)
  keep_clusters <- names(cluster_sizes[cluster_sizes >= 10])
  seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
  seu.day$cluster_type <- droplevels(seu.day$cluster_type)
  
  # 3. Rename Clusters based on Dictionary
  actual_day <- limma::strsplit2(day, "_")[[1]]
  bio_order <- bio_order_list[[ct]]
  
  if (actual_day %in% names(cluster_names_dict[[ct]])) {
    new_cluster_ids <- dplyr::recode(
      as.character(seu.day$cluster_type),
      !!!cluster_names_dict[[ct]][[actual_day]]
    )
    
    renamed_cluster_order <- unique(dplyr::recode(
      bio_order,
      !!!cluster_names_dict[[ct]][[actual_day]]
    ))
    
    seu.day$cluster_type_renamed <- factor(new_cluster_ids, levels = renamed_cluster_order)
    Idents(seu.day) <- "cluster_type_renamed"
  }
  
  seu.day$cluster_type_renamed <- droplevels(seu.day$cluster_type_renamed)
  table(seu.day$sample_names_renamed, seu.day$cluster_type_renamed)
  
  target_cell <- ifelse(ct == "EpiC", "H1", "H2")
  target_list <- list(
    "SCP_in_virus_vs_novirus" = list(
      "target_samples"  = paste(target_cell, c("STOP-MYCN no virus D42", "STOP-MYCN hTH-Cre D42"), sep = " "),
      "target_clusters" = c("late SCP", "late SCP")
    ),
    "EGFP_tumor_vs_SCP_in_virus" = list(
      "target_samples"  = paste(target_cell, c("STOP-MYCN hTH-Cre D42", "STOP-MYCN hTH-Cre D42"), sep = " "),
      "target_clusters" = c("late SCP", "EGFP ADRN-like")
    ),
    "ENF_in_virus_vs_novirus" = list(
      "target_samples"  = paste(target_cell, c("STOP-MYCN no virus D42", "STOP-MYCN hTH-Cre D42"), sep = " "),
      "target_clusters" = c("late endoneural fibroblast", "late endoneural fibroblast")
    ),
    "MES_tumor_vs_ENF_in_virus" = list(
      "target_samples"  = paste(target_cell, c("STOP-MYCN hTH-Cre D42", "STOP-MYCN hTH-Cre D42"), sep = " "),
      "target_clusters" = c("late endoneural fibroblast", "EGFP MES-like")
    )
  )
  
  # Collect unique combinations requested across all comparisons
  req_pairs <- map_dfr(target_list, function(x) {
    tibble(
      sample = x$target_samples,
      cluster = x$target_clusters
    )
  }) %>% distinct()
  
  # Filter Seurat object to keep cells matching valid (sample, cluster) pairs
  seu.day$group_check <- paste0(seu.day$cluster_type_renamed, "___", seu.day$sample_names_renamed)
  valid_groups <- paste0(req_pairs$cluster, "___", req_pairs$sample)

  seu.pb.sub <- subset(seu.day, subset = group_check %in% valid_groups)
  # ITT TARTOK
  seu.pb.sub$plot_id <- paste0(ct, "_", seu.pb.sub$cluster_type_renamed, "_", seu.pb.sub$sample.names)
  Idents(seu.pb.sub) <- "plot_id"
  
  # Generate Pseudobulk Counts Matrix
  pb_counts <- AggregateExpression(
    seu.pb.sub, 
    assays = "RNA", 
    slot = "counts", 
    return.seurat = FALSE
  )$RNA
  
  # Clean up column names for clarity
  colnames(pb_counts) <- gsub("[ -]", "_", colnames(pb_counts))
  
  pseudo_bulk[[ct]] <- as_tibble(pb_counts, rownames = "gene_symbol")
}

# 6. Merge both cell lines into final pb_all
pb_all <- reduce(pseudo_bulk, inner_join, by = "gene_symbol")
```




# calculate gene expression for each condition
```{r calculate normalised expression}
sample_annot <- data.frame(sample = colnames(pb_all)[-1]) %>%
  mutate(
    cell_line = factor(
      if_else(startsWith(sample, "EpiC"), "EpiC", "kolf2"), 
      levels = c("EpiC", "kolf2")
    ),
    condition = case_when(
      str_detect(sample, "late_SCP.*STOP_MYCN_D42$")                ~ "SCP_novirus",
      str_detect(sample, "late_SCP.*hTH")                           ~ "SCP_virus",
      str_detect(sample, "EGFP_ADRN")                              ~ "EGFP_ADRN_like_virus",
      str_detect(sample, "late_endoneural_fibroblast.*STOP_MYCN_D42$") ~ "ENF_novirus",
      str_detect(sample, "late_endoneural_fibroblast.*hTH")         ~ "ENF_virus",
      str_detect(sample, "EGFP_MES")                                ~ "EGFP_MES_like_virus"
    ),
    condition = factor(condition, levels = c(
      "SCP_novirus", "SCP_virus", "EGFP_ADRN_like_virus",
      "ENF_novirus", "ENF_virus", "EGFP_MES_like_virus"
    ))
  )

# Construct design matrix
design <- model.matrix(~ 0 + condition + cell_line, data = sample_annot)
rownames(design) <- sample_annot$sample

counts_matrix <- as.matrix(pb_all[, -1]) # Remove the gene_symbol column
rownames(counts_matrix) <- pb_all$gene_symbol

# 1. Create DGEList object (edgeR)
dge <- edgeR::DGEList(counts = counts_matrix)

# 2. Filter low expression genes (optional but recommended)
keep <- filterByExpr(dge, group = colnames(pb_counts))
dge <- dge[keep, , keep.lib.sizes = FALSE]
dge <- calcNormFactors(dge) # # Apply the TMM normalization (a standard step in the limma-voom workflow)


# 4. Limma-Voom transformation
vv <- voom(dge, design, plot = F)

# save & plot intermediate results
exprmx <- as_tibble(vv$E, rownames = "gene_symbol")
# #print(dim(exprmx))
# # merge with gene annotation
# exprmx <- right_join(gene_annot, exprmx)
# #print(dim(exprmx))

# # save
# tmpfile <- file.path(calcfolder, "edgeR_normcounts.txt")
# write.table(exprmx, file=tmpfile, sep="\t", row.names = F)
exprfile <- file.path(Results, "ps_bulk_logCPM_SCP_EGFP_tumor.xlsx")
write.xlsx(exprmx, file = exprfile, colNames=T, rowNames=F)

```


The distribution of normalised expressions provides a snapshot of the gene expression in each sample. 
```{r plot normalised expression of samples}
# for the boxplot, create a long tibble first
exprmx_long <- pivot_longer(exprmx, !c(gene_symbol),
                            names_to = "sample", 
                            values_to = "gene_expr")
# # add sample annotation as well
# exprmx_long <- full_join(exprmx_long, samples_annot, by=join_by(sample))

exprmx_long <- exprmx_long %>%
  mutate(
    # Extract everything before the first '_'
    cell = str_extract(sample, "^[^_]+"),
    
    # Check if the string cluster is tumor cluster
    cluster = ifelse(str_detect(sample, "EGFP"), "tumor", "normal"),
    
    # Check if the sample is virus or novirus
    type = ifelse(str_detect(sample, "hTH"), "virus", "novirus")
  )

# Verify the result
table(exprmx_long$sample, exprmx_long$cell)
table(exprmx_long$sample, exprmx_long$cluster)
table(exprmx_long$sample, exprmx_long$type)

exprmx_long$cell <- factor(exprmx_long$cell, levels=c("EpiC", "kolf2"))
exprmx_long$cluster <- factor(exprmx_long$cell, levels=c("normal", "tumor"))
exprmx_long$type <- factor(exprmx_long$type, levels=c("novirus", "virus"))


p <- ggplot(exprmx_long, aes(x = sample, y = gene_expr, fill = cell, color = type)) + 
  # Set alpha lower for more transparency
  geom_violin(alpha = 0.4, linewidth = 2) +
  # Boxplot inside with a bit more opacity than the violin
  geom_boxplot(width = 0.1, color = "black", outlier.shape = NA, alpha = 0.6) +
  # Use manual colors for Cancer
  scale_fill_manual(values = c("EpiC" = "orange", "kolf2" = "seagreen")) +
  # Aesthetic labels
  labs(
    y = expression(Log[2] ~ "Normalized Counts (CPM)"),
    x = "Sample ID",
    title = "Gene Expression Distribution per Sample"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    legend.position = "top",
    panel.grid.major.x = element_blank() # Cleans up the background for easier reading
  )

plot(p)

# also save the plot
#figfile <- file.path(figfolder, "boxplot_normalized.pdf")
figfile <- file.path(Figures, "ps_bulk_violinplot_normalized_SCP_EGFP_tumor.pdf")
ggsave(figfile, height=7, width=10)
```


## Principal Component Analysis (PCA)
To assess global sample relationships and identify potential batch effects or outliers, we performed a Principal Component Analysis (PCA) on the log-transformed expression data. 

### Interpretation of Variance
PCA is a dimensionality reduction technique that projects samples onto axes (Principal Components) that capture the greatest sources of variation in the dataset.
* **Sample Clustering:**

* **Limitations:** It is important to note that PCA captures the **largest** global variances. While this is useful for identifying major biological groups, subtle but biologically significant changes—particularly those involving smaller gene sets—may be masked by these dominant global trends and are better captured through downstream differential expression and pathway analysis.

The following plots visualize the first two principal components, which account for the majority of the total variance in the experiment. Click and drag to magnify.
```{r PCA the  data}
pca <- prcomp(t(vv$E), center = TRUE, scale. = FALSE)
# calculate explained variance
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

sumvar <- 0
for (pcind in 1:length(pca.var)){
  sumvar <- sumvar + pca.var.per[pcind]
  if (sumvar > 95){
    maxi <- pcind
    #print(maxi)
    break
  }
}
vardf <- as.data.frame(cbind('PC'=1:maxi, 'variance'=pca.var.per[1:maxi]))

# #  scree plot
# p <- ggplot(data=vardf) +
#   geom_col(mapping=aes(x=PC, y=cumsum(variance)))+
#   xlab("Principal Component") +  ylab("Cumulative percent variation") +
#   geom_hline(yintercept=95, linetype="dashed",
#              color = "red", linewidth=1)
# plot(p)

# nice labels
nicelabels <- paste0("PC", 1:maxi, " (", vardf$variance, "%)")

pca_df <- as_tibble(pca$x, rownames = "sample")
pca_annot <- inner_join(pca_df, sample_annot, by=join_by(sample))
pca_annot$cell_nice <- ifelse(pca_annot$cell_line=="EpiC", "H1", "H2")

p <- ggplot(pca_annot, aes(x = PC1, y = PC2, shape = cell_nice, color = condition,
                           text = paste("Sample:", sample, "<br>Group:", condition))) +
  geom_point(size = 3) +
  theme_gray(base_size = 18) + 
  scale_shape(name = "Cell Line")+
  scale_color_manual(name="Condition",
        values = c("EGFP_ADRN_like_virus"="#111111",
                                "SCP_virus"="#08519C", "SCP_novirus"="#4292C6",
                   "EGFP_MES_like_virus"="#00441B",
                                "ENF_virus"="#00796B", "ENF_novirus"="#41AE76"), 
       labels = c("EGFP_ADRN_like_virus"="EGFP ADRN-like virus",
                                "SCP_virus"="SCP virus", "SCP_novirus"="SCP novirus",
                   "EGFP_MES_like_virus"="EGFP MES-like",
                                "ENF_virus"="ENF virus", "ENF_novirus"="ENF novirus"))+
  xlab(nicelabels[[1]])+ylab(nicelabels[[2]])
  # geom_text(aes(label=Type_Exp), size = 2.5, vjust =  2.5)
print(p)

#plot(p)
outfig <- file.path(Figures, "pca_samples_SCP_EGFP_tumor.pdf")
ggsave(outfig, plot=p, height = 20, width= 30, units = "cm")

p <- ggplot(pca_annot, aes(x = PC3, y = PC2, shape = cell_nice, color = condition,
                           text = paste("Sample:", sample, "<br>Group:", condition))) +
  geom_point(size = 3) +
  theme_gray(base_size = 18) + 
  scale_shape(name = "Cell Line")+
  scale_color_manual(name="Condition",
        values = c("EGFP_tumor_virus"="#111111",
                                "SCP_virus"="#08519C", "SCP_novirus"="#4292C6",
                   "MES_tumor_virus"="#00441B",
                                "ENF_virus"="#00796B", "ENF_novirus"="#41AE76"), 
       labels = c("EGFP_tumor_virus"="EGFP ADRN-like virus",
                                "SCP_virus"="SCP virus", "SCP_novirus"="SCP novirus",
                   "MES_tumor_virus"="EGFP MES-like",
                                "ENF_virus"="ENF virus", "ENF_novirus"="ENF novirus"))+
  xlab(nicelabels[[3]])+ylab(nicelabels[[2]])
  # geom_text(aes(label=Type_Exp), size = 2.5, vjust =  2.5)
print(p)

#plot(p)
outfig <- file.path(Figures, "pca_samples_SCP_EGFP_tumor_PC3_vs_PC2.pdf")
ggsave(outfig, plot=p, height = 20, width= 30, units = "cm")

#---------------------- nicely formatted plot_ly 3D plot--------------------------#

# 1. Prepare data with one-liner strings
pca_annot <- pca_annot %>%
  mutate(
    condition_clean = str_replace_all(as.character(condition), "_", " "),
    legend_group = paste0(cell_nice, ": ", condition_clean)
  )

# 2. Map colors to the full "CellLine: Condition" string
legend_colors <- c(
  "H1: EGFP ADRN like virus" = "#111111", "H2: EGFP ADRN like virus" = "#111111",
  "H1: SCP virus"        = "#08519C", "H2: SCP virus"        = "#08519C",
  "H1: SCP novirus"      = "#4292C6", "H2: SCP novirus"      = "#4292C6",
  "H1: EGFP MES like virus"  = "#00441B", "H2: EGFP MES like virus"  = "#00441B",
  "H1: ENF virus"        = "#00796B", "H2: ENF virus"        = "#00796B",
  "H1: ENF novirus"      = "#41AE76", "H2: ENF novirus"      = "#41AE76"
)

# 3. Map symbols to the full "CellLine: Condition" string
# (H1 entries get 'circle', H2 entries get 'diamond')
legend_symbols <- c(
  "H1: EGFP ADRN like virus" = "circle",  "H2: EGFP ADRN like virus" = "diamond",
  "H1: SCP virus"        = "circle",  "H2: SCP virus"        = "diamond",
  "H1: SCP novirus"      = "circle",  "H2: SCP novirus"      = "diamond",
  "H1: EGFP MES like virus"  = "circle",  "H2: EGFP MES like virus"  = "diamond",
  "H1: ENF virus"        = "circle",  "H2: ENF virus"        = "diamond",
  "H1: ENF novirus"      = "circle",  "H2: ENF novirus"      = "diamond"
)

# 4. Build 3D Plot
p3d <- plot_ly(
  pca_annot, 
  x = ~PC1, 
  y = ~PC2, 
  z = ~PC3,
  color = ~legend_group,                 # Single variable controls legend entries
  colors = legend_colors,
  symbol = ~legend_group,                # Passed here to prevent secondary legend traces
  symbols = legend_symbols,              # Explicit mapping per combined key
  type = "scatter3d", 
  mode = "markers",
  marker = list(size = 7), 
  text = ~paste("Sample:", sample, "<br>", legend_group),
  hoverinfo = "text"
) %>%
  layout(
    font = list(size = 14),              
    legend = list(
      font = list(size = 18),            # Legend font size
      itemsizing = "constant"            
    ),
    scene = list(
      xaxis = list(
        title = nicelabels[[1]],
        backgroundcolor = "#E5E5E5",
        gridcolor = "#FFFFFF",
        showbackground = TRUE,
        zerolinecolor = "#FFFFFF"
      ),
      yaxis = list(
        title = nicelabels[[2]],
        backgroundcolor = "#E5E5E5",
        gridcolor = "#FFFFFF",
        showbackground = TRUE,
        zerolinecolor = "#FFFFFF"
      ),
      zaxis = list(
        title = nicelabels[[3]],
        backgroundcolor = "#E5E5E5",
        gridcolor = "#FFFFFF",
        showbackground = TRUE,
        zerolinecolor = "#FFFFFF"
      )
    )
  )

# 2. Save the plot as an interactive HTML file
htmlwidgets::saveWidget(as_widget(p3d), 
  file.path(Figures, "pca_samples_SCP_EGFP_tumor_3D.html"))
```


## Differential Gene Expression Analysis

Differential Gene Expression (DGE) was performed to identify transcriptomic shifts across various biological states. For all comparisons, **upregulated** genes are defined as those with significantly higher expression in the second condition of the contrast, while **downregulated** genes are those with significantly lower expression.


### Output file
Results are saved to diffexpr_all.xlsx and diffexpr_lfc1_adjp0.05_signif.xlsx files, where each tab contains the results of one of the above contrasts. In diffexpr_lfc1_adjp0.05_signif.xlsx only genes are listed for each contrast whose expression changed with at least 1 log fold change and adjusted p value <= 0.05.

```{r calculate differential gene expression}

# Fit the linear model
fit <- lmFit(vv, design)


# Note: Formatted as (Target - Baseline) so positive logFC = higher expression in Target
contrast.matrix <- makeContrasts(
  # 1. SCP: Virus vs No Virus baseline
  SCP_virus_vs_novirus     = conditionSCP_virus - conditionSCP_novirus,
  
  # 2. EGFP ADRN-like vs SCP baseline (in virus condition)
  EGFP_tumor_vs_SCP_virus  = conditionEGFP_ADRN_like_virus - conditionSCP_virus,
  
  # 3. ENF: Virus vs No Virus baseline
  ENF_virus_vs_novirus     = conditionENF_virus - conditionENF_novirus,
  
  # 4. MES Tumor vs ENF baseline (in virus condition)
  MES_tumor_vs_ENF_virus   = conditionEGFP_MES_like_virus - conditionENF_virus,
  
  # 4. MES Tumor vs ENF baseline (in virus condition)
  EGFP_tumor_vs_MES_tumor   = conditionEGFP_ADRN_like_virus - conditionEGFP_MES_like_virus,
  
  levels = design
)

# # 2. Create the pairwise contrasts for the 3 conditions
# contrast.matrix <- makeContrasts(
#   # Compare Virus vs No Virus (Baseline)
#   SCP_VirusVsNoVirus = conditionSCP_virus,
#   
#   # # Compare Tumor vs No Virus (Baseline)
#   # TumorVsNoVirus = conditionEGFP_tumor_virus,
#   
#   # Compare Tumor vs Virus (Subtracting the two coefficients)
#   EGFP_TumorVsSCP_Virus   = conditionEGFP_tumor_virus - conditionSCP_virus,
#   
#   # Optional: Keep your cell line comparison to control for background differences
#   Kolf2VsEpiC    = cell_linekolf2 - cell_lineEpiC,
#   
#   levels = design
# )

fitc <- contrasts.fit(fit,contrast.matrix )
fite <- eBayes(fitc) ## gives moderated t-statistics
# save result for further use
outfile <- file.path(Results, "limma_fit_diffexpr_SCP_EGFP_tumor_all_conditions.rds")
saveRDS(fite, file=outfile)

# also write out to human readable file
lthr <- 1; pthr <- 0.05
thrchar <- paste(sprintf("lfc%01d", lthr), 
                   sprintf("adjp%.2f", pthr), sep="_")
outdata <- list()
outdata_signif <- list() # only write out genes whose diff expr is signif
for (cf in colnames(fite$coefficients)){
  toptt <- limma::topTable(fite, coef=cf, sort.by = "logFC",
          number=nrow(fite$coefficients))
  toptt <- as_tibble(toptt, rownames = "Gene_ID")
  # # add annotation
  # toptt <- left_join(toptt, gene_annot, by=c("Gene_ID"="gene_id"))
  
  # formatting
  toptt <- toptt %>%
    mutate(
      across(c(logFC, AveExpr, t, B), ~ round(., 2)),
      # P.Value and adj.P.Val formatted as strings to preserve notation in Excel
      P.Value = format.pval(P.Value),
      adj.P.Val = format.pval(adj.P.Val)
    )
  
  # excel tab names
  lname <- substr(cf,1, min(nchar(cf), 31)) # for each tab name, a maximum of 31 length is allowed
  outdata[[lname]] <- toptt

  # repeat with significant genes only
  toptt_thr <- limma::topTable(fite, coef=cf, sort.by = "logFC",
                p.value=pthr, lfc=lthr, number=nrow(fite$coefficients))
  toptt_thr <- as_tibble(toptt_thr, rownames = "Gene_ID")
  # # add annotation
  # toptt_thr <- left_join(toptt_thr, gene_annot, by=c("Gene_ID"="gene_id"))
    # formatting
  toptt_thr <- toptt_thr %>%
    mutate(
      across(c(logFC, AveExpr, t, B), ~ round(., 2)),
      # P.Value and adj.P.Val formatted as strings to preserve notation in Excel
      P.Value = format.pval(P.Value),
      adj.P.Val = format.pval(adj.P.Val)
    )
  outdata_signif[[lname]] <- toptt_thr
}
# write into a compound excel file
outfile <- file.path(Results, "diffexpr_pseudobulk_all_SCP_EGFP_tumor_all_conditions.xlsx")
write.xlsx(outdata, file = outfile, colNames=T, rowNames=F)
outfile_signif <- outfile <- file.path(Results, 
                     paste("diffexpr_pseudobulk_", thrchar, "_signif_all_SCP_EGFP_tumor_all_conditions.xlsx", sep=""))
write.xlsx(outdata_signif, file = outfile_signif, colNames=T, rowNames=F)

# # add volcano plot using EnhancedVolcano
# res <- topTable(fite, coef = "EGFP_TumorVsSCP_Virus", number = Inf)
#
# p <- EnhancedVolcano(res,
#   lab = rownames(res),
#   x = 'logFC',
#   y = 'adj.P.Val',
#   title = 'Tumor vs Normal',
#   pCutoff = 0.05,
#   FCcutoff = 1.0,
#   pointSize = 2.0,
#   labSize = 3.0)
# 
# outfig <- file.path(Figures, "volcano_plot_SCP_EGFP_tumor.pdf")
# ggsave(outfig, plot=p, height = 20, width= 30, units = "cm")

# Generate and save Volcano Plots for ALL contrasts
for (cf in colnames(fite$coefficients)) {
  # Pull unformatted numeric table for EnhancedVolcano
  res_numeric <- topTable(fite, coef = cf, number = Inf)
  p <- EnhancedVolcano(
    res_numeric,
    lab = rownames(res_numeric),
    x = 'logFC',
    y = 'adj.P.Val',
    title = gsub("_", " ", cf),
    subtitle = paste("Pseudobulk DE (adj.P <", pthr, ", |logFC| >", lthr, ")"),
    pCutoff = pthr,
    FCcutoff = lthr,
    pointSize = 2.0,
    labSize = 3.0
  )
  fig_path <- file.path(Figures, paste0("volcano_plot_", cf, ".pdf"))
  ggsave(fig_path, plot = p, height = 20, width = 30, units = "cm")
}

```


# Heatmap

```{r heatmap}
num_genes <- 30

# 1. Modify Column Annotation Metadata with standardized names
heatmap_anno <- sample_annot %>% 
  mutate(
    # Convert cell line IDs: EpiC -> H1, everything else (kolf2) -> H2
    Cell_Line = ifelse(cell_line %in% c("EpiC", "H1"), "H1", "H2"),
    
    # Replace underscores with spaces to match PCA legend naming
    Condition = as.character(condition) %>%
      str_replace_all("_", " ") %>%
      str_replace_all("novirus", "no AAV1") %>%   # do this BEFORE the plain "virus" replace
      str_replace_all("virus", "AAV1")
  ) %>%
  column_to_rownames("sample") %>% 
  dplyr::select(Cell_Line, Condition)


# 2. Define custom hex colors for Condition AND Cell_Line matching PCA palette
anno_colors <- list(
  Condition = c(
    "EGFP ADRN like AAV1" = "#111111",
    "SCP AAV1"        = "#08519C", 
    "SCP no AAV1"      = "#4292C6",
    "EGFP MES like AAV1"  = "#00441B",
    "ENF AAV1"        = "#00796B", 
    "ENF no AAV1"      = "#41AE76"
  ),
  Cell_Line = c(
    "H1" = "#4A4A4A",
    "H2" = "#D1D1D1"
  )
)


# 3. Extract the underlying expression matrix
expr_mat <- if (inherits(vv, "EList")) vv$E else as.matrix(vv)


# 4. Helper function to extract top 50 up & top 50 down non-ENS genes
get_heatmap_genes <- function(limma_fit, coefficient_name) {
  tt <- limma::topTable(limma_fit, coef = coefficient_name, number = Inf)
  tt_filtered <- as_tibble(tt, rownames = "Gene_ID") %>%
    filter(!grepl("^ENS", Gene_ID)) %>% 
    filter(adj.P.Val < 0.05)
  
  top_up   <- tt_filtered %>% arrange(desc(logFC)) %>% head(num_genes) %>% pull(Gene_ID)
  top_down <- tt_filtered %>% arrange(logFC)        %>% head(num_genes) %>% pull(Gene_ID)
  
  return(c(top_up, top_down))
}


# 5. Gather unique target gene sets across contrasts
genes_comp1 <- get_heatmap_genes(fite, "EGFP_tumor_vs_SCP_virus")
genes_comp2 <- get_heatmap_genes(fite, "MES_tumor_vs_ENF_virus")
combined_genes <- unique(c(genes_comp1, genes_comp2))

# Subset the master matrix
mat_combined <- expr_mat[combined_genes, , drop = FALSE]


# 6. Generate and save the modified combined heatmap using ComplexHeatmap
outfig_combined <- file.path(Figures, "heatmap_combined_progression_SCP_EGFP_tumor.pdf")

# Generate heatmap object with transposed matrix (samples as rows, genes as columns)
ht <- ComplexHeatmap::pheatmap(
  t(mat_combined),                            
  scale = "column",                            # Z-score scaling across genes (now columns)
  annotation_row = heatmap_anno,              # Sample annotations on rows
  annotation_colors = anno_colors,            # Exact hex palette for Condition & Cell_Line
  show_rownames = FALSE,                      # Hide individual sample IDs
  show_colnames = TRUE,                       # Display gene names
  cluster_cols = TRUE,              
  cluster_rows = TRUE,              
  main = "z-score",
  fontsize_col = 7,
  heatmap_legend_param = list(title = "")     # Remove z-score title text
)

# Render plot and save to PDF
pdf(outfig_combined, width = 16, height = 5) 

draw(
  ht, 
  heatmap_legend_side = "top", 
  annotation_legend_side = "top"
)

dev.off()
```


## Functional Enrichment Analysis

To interpret the biological significance of the identified differentially expressed (DGE) and equivalent gene sets, we performed functional enrichment analysis using curated gene set collections from the Molecular Signatures Database (MSigDB).

### Gene Set Collections
We utilized three distinct categories of gene sets to capture a broad range of biological insights:

* **Hallmark Gene Sets:** Represent specific, well-defined biological states or processes (e.g., hypoxia, inflammatory response) and provide a concise summary of the primary pathways involved.
* **KEGG Medicus:** A subset of the Kyoto Encyclopedia of Genes and Genomes (KEGG) that focuses on the molecular networks of diseases and drug mechanisms, offering a more clinical and mechanistic perspective.
* **GO Biological Process (GO:BP):** Comprehensive ontologies that describe the broader biological objectives (such as cell cycle or metabolic pathways) to which a gene contributes.

This multi-faceted approach allows us to determine if the genes identified as "changing" or "stable" are non-randomly associated with specific biological functions or signaling pathways.

```{r functional analysis, message=FALSE, warning=FALSE}
geneset_collections <- list(
  Hallmark = msigdbr(species = "Homo sapiens", collection = "H") %>%
             dplyr::select(gs_name, gene_symbol, ensembl_gene),

  KEGG_Medicus = msigdbr(species = "Homo sapiens", collection = "C2") %>%
                 filter(gs_subcollection == "CP:KEGG_MEDICUS") %>%
                 dplyr::select(gs_name, gene_symbol, ensembl_gene),

  GO_Biological_Process = msigdbr(species = "Homo sapiens", collection = "C5") %>%
                          filter(gs_subcollection == "GO:BP") %>%
                          dplyr::select(gs_name, gene_symbol, ensembl_gene)
)
```

# Create gene annotation
```{r}

# 1. Get your list of gene symbols from the object
current_symbols <- rownames(seu.integrated)

# 2. Query the database for IDs/Annotations
gene_info <- AnnotationDbi::select(org.Hs.eg.db,
                    keys = current_symbols,
                    columns = c("ENSEMBL", "GENENAME"),
                    keytype = "SYMBOL")

# 3. Clean up the query (handle potential duplicates/NAs)
# (Optional: remove duplicates if a symbol maps to multiple Ensembl IDs)
gene_info <- gene_info[!duplicated(gene_info$SYMBOL), ]
# keep non-NA only
gene_info <- gene_info[!is.na(gene_info$ENSEMBL),]
```

### Gene Set Enrichment Analysis (GSEA)

While Overrepresentation Analysis (ORA) treats gene lists as unranked sets, **Gene Set Enrichment Analysis (GSEA)** utilizes the full spectrum of gene expression changes. This allows for the detection of subtle but coordinated shifts in entire biological pathways that might be missed by simple threshold-based filtering.

#### 1. Analysis Workflow: Ranking and Running
For each experimental contrast, we performed the following steps:
* **Gene Ranking:** All genes were ranked by their $log_2(\text{Fold Change})$ in descending order. This ensures that the most upregulated genes are at the "top" of the list and the most downregulated genes are at the "bottom."
* **Enrichment Calculation:** Using the `clusterProfiler::GSEA` function, we calculated a running sum statistic to identify pathways where genes are non-randomly distributed at either end of the ranked list.

#### 2. Key Metrics and "Core Enrichment"
The results are summarized using two primary metrics:
* **Normalized Enrichment Score (NES):** A positive NES indicates the pathway is generally **activated** (enriched in upregulated genes), while a negative NES indicates the pathway is **suppressed** (enriched in downregulated genes).
* **Core Enrichment:** We extracted the specific subset of "leading-edge" genes—those that contribute most significantly to the enrichment score—and reported them along with their individual $log_2(\text{Fold Change})$ values.

#### 3. Visualization and Export
* **Summary Dotplots:** For each comparison, we visualized the top 10 activated and top 10 suppressed pathways. The size of the points represents the significance ($-\log_{10}(\text{Adjusted P-Value})$), and the x-axis indicates the NES.

#### Output file
Results are saved to GSEA_differential_expression_[geneset_name]_results.xlsx file, where each tab contains the results of one of the contrasts and only those genesets are listed for each contrast where the adjusted p value <= 0.05. If the enrichment process resulted no genesets, then the contrast is not shown in the file. If none of the genesets were significant, then the corresponding tab is empty.

```{r GSEA for differentially expressed genes, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
gsea_outer_list <- list()
p_multi_list <- list()
raw_gsea_models <- list()
 for (gs_name in names(geneset_collections)){
   current_t2g <- geneset_collections[[gs_name]]
   
    # add gene symbols
    id_map <- current_t2g %>%
    dplyr::select(ensembl_gene, gene_symbol) %>%
    distinct() %>%
      deframe() # Converts to a named vector
   
   # print("########################################")
   # print(paste("#   Current gene set:", gs_name, "    #"))
   # print("########################################")
   gsea_res_list <- list()
   # Iterate through your contrasts
   # for (contr in colnames(fite$coefficients)) {
   for (contr in c("EGFP_tumor_vs_SCP_virus", "MES_tumor_vs_ENF_virus", "EGFP_tumor_vs_MES_tumor")) {
     # print(paste("Current contrast:", contr))

     # 1. Prepare the ranked list for GSEA
     full_stats <- topTable(fite, coef = contr, number = Inf)
     full_tbl <- as_tibble(full_stats, rownames="SYMBOL")
     # keep genes with ensembl id only
     full_tbl <- inner_join(gene_info, full_tbl, by="SYMBOL")

     # Create a named vector for GSEA ranking
     gene_list <- full_tbl$logFC
     names(gene_list) <- full_tbl$ENSEMBL
     # remove_duplicated
     keep <- which(!duplicated(names(gene_list)))
     gene_list <- gene_list[keep]
     

     # Create a lookup for the table mapping (clean Ensembl IDs)
     lfc_lookup <- full_stats$logFC
     names(lfc_lookup) <- sub("\\..*$", "", rownames(full_stats))

     # Sort for GSEA
     gene_list <- sort(gene_list, decreasing = TRUE)

     # 2. Run GSEA
     gsea_out <- clusterProfiler::GSEA(
       geneList = gene_list,
       TERM2GENE = current_t2g[, c("gs_name", "ensembl_gene")],
       pvalueCutoff = 0.05,
       verbose = FALSE
     )
     
     # store for later use
     raw_gsea_models[[gs_name]] <- gsea_out

     if (is.null(gsea_out) || nrow(as.data.frame(gsea_out)) == 0) {
       next
     }

     # 3. Process results and map Symbols + LogFC
     result_df <- as.data.frame(gsea_out)

     # 4. Update the mapping logic to include LogFC for Core Enrichment
     result_df_final <- result_df %>%
       mutate(MSigDB_Link = paste0("<a href='https://www.gsea-msigdb.org/gsea/msigdb/cards/",
                                   Description, "' target='_blank'>Link</a>")) %>%
       mutate(core_enrichment_stats = map_chr(core_enrichment, function(id_string) {
         # Split IDs (GSEA uses "/" as delimiter)
         ids <- unlist(strsplit(id_string, "/"))

         # Get Symbols from your id_map
         symbols <- id_map[ids]
         symbols[is.na(symbols)] <- ids[is.na(symbols)] # fallback to ID if symbol missing

         # Get LogFC values from our lookup
         # logfcs <- lfc_lookup[ids]
         logfcs <- lfc_lookup[symbols]

         # Format as "Symbol (LogFC)"
         formatted <- paste0(symbols, " (", round(logfcs, 2), ")")

         # Combine into a single string
         paste(formatted, collapse = ", ")
       })) %>%
       # Select columns for the final table
       dplyr::select(Description, NES, p.adjust, core_enrichment_stats, MSigDB_Link)

     # Format p-value to scientific
     result_df_final$p.adjust <- formatC(as.numeric(result_df_final$p.adjust), format = "e", digits = 2)
     result_df_final$NES <- formatC(as.numeric(result_df_final$NES), format = "e", digits = 2)

     # Add to list for Excel export
     gsea_res_list[[contr]] <- result_df_final

     # # 5. Create HTML Datatable (with the same JS logic for long text)
     # html_table <- datatable(result_df_final,
     #           escape = FALSE,
     #           filter = 'top',
     #           extensions = 'Buttons',
     #           options = list(
     #             pageLength = 10,
     #             dom = 'Bfrtip',
     #             buttons = c('copy', 'csv', 'excel'),
     #             columnDefs = list(list(
     #                 targets = 3, # Targets the core_enrichment_stats column
     #                 width = '500px',
     #                 render = JS(
     #                   "function(data, type, row, meta) {",
     #                   "return type === 'display' && data.length > 100 ?",
     #                   "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
     #                   "}"
     #                 )
     #             ))
     #           ),
     #           caption = paste("GSEA (Ranked by logFC):", contr))

     # Save individual widget
     #saveWidget(html_table, file.path(figfolder, paste0("GSEA_", contr, ".html")))

     # also add a figure
     if (!is.null(gsea_out) && nrow(as.data.frame(gsea_out)) > 0) {

     # 1. Identify the top 3 pathways by absolute NES
     # (You can change this to show the most significant or specific pathways of interest)
     top_pathways <- as.data.frame(gsea_out) %>%
       arrange(desc(abs(NES))) %>%
       slice_head(n = 3) %>%
       pull(ID)

     # # 2. Generate the plot for the top pathway
     # # p1: The Enrichment Score curve
     # # p2: The gene 'ticks' (hits)
     # # p3: The rank-ordered gene list gradient
     # p_gsea <- gseaplot2(gsea_out,
     #                     geneSetID = top_pathways[1],
     #                     title = paste("Top Pathway:", top_pathways[1]),
     #                     subplots = 1:3,
     #                     pvalue_table = TRUE)
     # 
     # # 3. Save the plot as a high-resolution PDF or PNG
     # # Since gseaplot2 returns a ggplot-compatible object, we use ggsave
     # plot_filename <- file.path(figfolder, paste0("GSEA_Plot_", contr, ".pdf"))
     # 
     # ggsave(plot_filename,
     #        plot = p_gsea,
     #        width = 10,
     #        height = 8,
     #        dpi = 300,
     #        bg = "white")

     # Optional: Plot multiple top pathways together
     p_multi <- enrichplot::gseaplot2(gsea_out,
                          geneSetID = top_pathways,
                          title = paste("Top 3 Pathways in", contr))
     p_multi_wrapped <- patchwork::wrap_plots(p_multi, ncol = 1, heights = c(2, 0.5, 0.5))
     # p_multi_list[[contr]] <- p_multi_wrapped
     subfolder <- file.path(Figures, gs_name)
     if (!file.exists(subfolder)) dir.create(subfolder)
     ggsave(file.path(subfolder, paste0("GSEA_MultiPlot_", contr, "_SCP_EGFP_tumor_all_conditions.pdf")),
            plot = p_multi, width = 12, height = 9)
     }
   }

   # Write out to Excel
   outfile_gsea <- file.path(Results, paste("GSEA_differential_expression_", gs_name, "_results_SCP_EGFP_tumor_all_conditions.xlsx", sep=""))
   write.xlsx(gsea_res_list, file = outfile_gsea, colNames = TRUE)
   gsea_outer_list[[gs_name]] <- gsea_res_list
 }
```


```{r dotplot of GSEA results, fig.width=20, fig.height=20}
for (gs_name in names(gsea_outer_list)){
  gsea_res_list <- gsea_outer_list[[gs_name]]
  plist <- list()
  for (gsi in names(gsea_res_list)){
    # create nice title:
    gsi_title <- switch(gsi,
      "EGFP_tumor_vs_SCP_virus"="EGFP ADRN-like vs SCP virus",
      "MES_tumor_vs_ENF_virus"="MES tumor vs ENF virus",
      gsi
    )
    clean_title <- paste(gs_name, gsi_title, sep = " - ")
    clean_title <- gsub("_", " ", clean_title)
    
    y <- gsea_res_list[[gsi]]
    
    # select activated and supressed genesets
    y$NES <- as.numeric(y$NES)
    y$p.adjust <- as.numeric(y$p.adjust)
    y$direction <- "activated"
    y$direction[y$NES<=0] <- "suppressed"
    numact <- length(which(y$direction=="activated"))
    numsupp <- length(which(y$direction=="suppressed"))

    # pick top and bottom genesets
    act_Nvals <- y$NES[y$direction=="activated"]
    act_thr <- sort(act_Nvals, decreasing = T)[min(10, numact)]
    act_keep <- which(y$direction=="activated" & y$NES >= act_thr)
    sup_Nvals <- y$NES[y$direction=="suppressed"]
    sup_thr <- sort(sup_Nvals, decreasing = F)[min(15, numsupp)]
    sup_keep <- which(y$direction=="suppressed" & y$NES <= sup_thr)
    y <- y[c(act_keep, sup_keep),]
    
    # sort by normalized enrichment
    y <- y[order(y$NES, decreasing = F),]
    # change the names of the gene sets
    #y$Description <- tolower(y$Description)
    #y$Description <- gsub(prefix[gsi], "", y$Description) # later if needed prefix <- c("gobp_", "kegg_", "hallmark_")
    y$Description <- gsub("_", " ", y$Description)
    y$Description <- str_to_sentence(y$Description)
    y$Description <- gsub("Gobp", "GO:BP", y$Description)
    y$Description <- gsub("dna", "DNA", y$Description)
    y$Description <- gsub("Dna", "DNA", y$Description)
    y$Description <- str_wrap(y$Description, width = 50) # width=30 is a good starting point for a 4-column layout
    #y$Description <- nice_names$name[match(y$Description, nice_names$compare)]
    y$size <- -log10(as.numeric(y$p.adjust))
    
    num_suppressed <- sum(y$direction == "suppressed")
    sep_line_pos <- num_suppressed + 0.5
    y$Description <- factor(y$Description, levels=y$Description)
      p <- ggplot(y, 
                    aes(x = NES, y = Description)) + 
               geom_point(aes(size = size, color = direction))+
               scale_color_manual(values = c("activated" = "firebrick", "suppressed" = "dodgerblue4")) +
               
              scale_size_continuous(
                 name = "-log10(p.adj)", 
                 limits = c(0, 20),      # <--- This is the fixed range
                 breaks = c(2, 4, 6, 8) # <--- Fixed circle sizes in legend
               ) +  
              
              # Only add it if both directions exist to avoid errors
             {if(num_suppressed > 0 && num_suppressed < nrow(y)) 
               geom_hline(yintercept = sep_line_pos, linetype = "solid", color = "gray80", size = 0.5)} +
              
               guides(size=guide_legend(title="-log(p.adjust)", ncol=1,byrow=TRUE,
                                title.vjust=0.9, order=1))+
               theme_bw(base_size = 16) +

               theme(
                 axis.text.y = element_text(size = 16, lineheight = 0.8),
                 panel.grid.minor = element_blank(),
                 legend.position="right",
                      legend.box = "vertical",
                     #axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
                     #strip.background =element_rect(fill=cfill),
                     strip.text = element_text(colour = 'black'),
                 legend.title = element_text(size = 16), # Legend title size
                  legend.text = element_text(size = 14)                 # Legend item text size)
               )
               ylab(NULL) + xlab("NES")+
               #scale_size(range=c(3, 8))+
               #scale_y_discrete(labels = function(x) str_wrap(x, width = 24))+
               geom_vline(xintercept = 0, linetype="dashed")+
               labs(title=clean_title)
      #plot(p)
      plist[[gsi]] <- p
    # figfile <- file.path(figfolder, paste("GSEA_", gs_name, "_", gsi, "_enrich_dotplot.pdf", sep=""))
    # ggsave(figfile, plot=p, width=14, height=6)
      figfile <- file.path(Figures, paste0("GSEA_enrich_dotplot_combined_SCP_EGFP_tumor_all_conditions_", gs_name, "_", gsi, "_dotplot.pdf"))
      ggsave(figfile, plot = p, width = 10, height = 7)
  }
  # plot together for the three genesets
  # pp_combined <- patchwork::wrap_plots(plist, ncol = 1) + 
  # patchwork::plot_layout(guides = "collect") & # Collects legends into one place
  # theme(legend.position = "right")
  # plot(pp_combined)
  # ggsave(file.path(Figures, paste("GSEA_", gs_name, "_enrich_dotplot_combined_SCP_EGFP_tumor_all_conditions.pdf", sep="")), 
  #        plot = pp_combined, width = 10, height = 10)
}
```
