---
title: "Demultiplexing and Index Hopping Analysis"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width=12, fig.height=12)
library(openxlsx)
library(tidyverse)
library(scales)
library(here)
```

# Overview

This report summarizes demultiplexing performance, index mismatch profiles, and index hopping for a single-cell RNA-seq experiment using kinetic confinement–based cell separation.
All samples were sequenced on Lane 8 of a single flowcell.

<!-- The analysis addresses: -->

<!-- ## Read assignment efficiency -->

<!-- ## Index mismatch rates -->

<!-- ## Undetermined reads and unknown barcode composition -->

<!-- ## Index hopping magnitude and directionality -->

<!-- ## Consequences for downstream single-cell analysis -->

## Demultiplexing statistics 
```{r}
demux <- read.xlsx(
  here("Data", "BS_stats_anonymous.xlsx"),
  sheet = "Demultiplex_Stats",
  rows = 1:14
)

colnames(demux) <- gsub("#", "Num", colnames(demux))
colnames(demux) <- gsub("%", "Perc", colnames(demux))

# sequencing depth and undetermined fraction
total_reads <- sum(demux$Num.Reads, na.rm = TRUE)

undet_reads <- demux %>%
  filter(SampleID == "Undetermined") %>%
  pull(Num.Reads)

undet_percent <- (undet_reads / total_reads) * 100

flowcell_summary <- tibble(
  Total_Reads = total_reads,
  Undetermined_Reads = undet_reads,
  Undetermined_Percent = round(undet_percent, 2)
)

knitr::kable(flowcell_summary, caption = "Flowcell-level sequencing summary")

```

---

## Per-sample demultiplexing 
```{r demux-health}
demux_report <- demux %>%
  filter(SampleID != "Undetermined") %>%
  mutate(Percent_of_Flowcell = Num.Reads / total_reads * 100) %>%
  select(
    SampleID,
    Num.Reads,
    Percent_of_Flowcell,
    Perc.Perfect.Index.Reads,
    Perc.One.Mismatch.Index.Reads
  ) %>%
  arrange(desc(Num.Reads))

knitr::kable(
  demux_report,
  digits = 3,
  caption = "Per-sample demultiplexing performance"
)
```



## Visualization of read balance across samples 
```{r demux-barplot, fig.height=4}
ggplot(demux_report, aes(x = reorder(SampleID, Num.Reads), y = Num.Reads)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Sample",
    y = "Number of reads",
    title = "Read distribution across samples"
  ) +
  theme_minimal()
```



## Composition of undetermined reads (top unknown barcodes) - Problem: polyG barcodes (undetermined in 2-color chemistry)
```{r unknown-barcodes}
# Read in "Top_Unknown_Barcodes" 
top_unknown <- openxlsx::read.xlsx(
  here("Data", "BS_stats_anonymous.xlsx"),  
  sheet = "Top_Unknown_Barcodes"
) 

colnames(top_unknown) <- gsub("#", "Num", colnames(top_unknown))
colnames(top_unknown) <- gsub("%", "Perc", colnames(top_unknown))

top_unknown_summary <- top_unknown %>%
  arrange(desc(Num.Reads)) %>%
  head(20)

knitr::kable(
  top_unknown_summary,
  digits = 4,
  caption = "Top 20 unknown index combinations"
)
```


## Index hopping relationships - Some index hopping is available
```{r hopping-reconstruction}
# The index hopping counts
index_hopping <- openxlsx::read.xlsx(
  here("Data", "BS_stats_anonymous.xlsx"), 
  sheet = "Index_Hopping_Counts"
)

# CRITICAL FIX: Replaced brittle hardcoded row subtraction with explicit NA filtering
index_hopping <- index_hopping %>% 
  filter(!is.na(index) & index != "")

colnames(index_hopping) <- gsub("#", "Num", colnames(index_hopping))
colnames(index_hopping) <- gsub("%", "Perc", colnames(index_hopping))

# 2. Separate valid assignments from hopped reads
valid_samples <- index_hopping %>% 
  filter(SampleID != "" & !is.na(SampleID)) %>%
  select(SampleID, Index1 = index, Index2 = index2, Valid_Reads = Num.Reads)

hopped_reads <- index_hopping %>% 
  filter(SampleID == "" | is.na(SampleID)) %>%
  select(Index1 = index, Index2 = index2, Hopped_Reads = Num.Reads)

analysis <- hopped_reads %>%
  left_join(
    valid_samples %>% select(SampleID_1 = SampleID, Index1),
    by = "Index1"
  ) %>%
  left_join(
    valid_samples %>% select(SampleID_2 = SampleID, Index2),
    by = "Index2"
  ) %>%
  filter(!is.na(SampleID_1) & !is.na(SampleID_2))

sample_hopping_summary <- analysis %>%
  group_by(SampleID_1) %>%
  summarise(
    Total_Hopped_Reads = sum(Hopped_Reads),
    Top_Donor = SampleID_2[which.max(Hopped_Reads)],
    .groups = "drop"
  ) %>%
  left_join(
    valid_samples %>% select(SampleID, Valid_Reads),
    by = c("SampleID_1" = "SampleID")
  ) %>%
  mutate(Hopping_Rate_Percent = Total_Hopped_Reads / Valid_Reads * 100) %>%
  arrange(desc(Hopping_Rate_Percent))

knitr::kable(
  sample_hopping_summary,
  digits = 3,
  caption = "Per-sample index hopping burden"
)
```


## Visualization: index hopping rate per sample
```{r hopping-barplot, fig.height=4}
ggplot(sample_hopping_summary,
       aes(x = reorder(SampleID_1, Hopping_Rate_Percent),
           y = Hopping_Rate_Percent)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Sample",
    y = "Index hopping rate (%)",
    title = "Index hopping burden per sample"
  ) +
  theme_minimal()

analysis %>%
  group_by(SampleID_1, SampleID_2) %>%
  summarise(Hopped_Reads = sum(Hopped_Reads), .groups = "drop") %>%
  ggplot(aes(SampleID_2, SampleID_1, fill = Hopped_Reads)) +
  geom_tile() +
  scale_fill_viridis_c(trans = "log10") +
  labs(
    x = "Donor sample (Index2)",
    y = "Recipient sample (Index1)",
    fill = "Hopped reads",
    title = "Index hopping source–recipient heatmap"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
