---
title: "Analysis of single cell neuroblastoma organoids: part 5 - pseudotime analysis"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---
```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).
The objective of this R Markdown code is to analyse the results obtained by the single cell RNA sequencing.

```{r libraries}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
packages <- c(
  "Seurat",       
  "scDblFinder",  
  "ggplot2",
  "patchwork",           
  "plotly",              
  "ggforce",             
  "Matrix",
  "dplyr",
  "here",
  "stringr",             
  "future",              
  "scCustomize",    
  "openxlsx",
  "destiny",
  "SingleCellExperiment",
  "slingshot",
  "viridis"
)
invisible(lapply(packages, library, character.only = TRUE))

```

<!-- # Biztosan kellene sejt -->
<!-- # populáció/cluster azonosítás marker gének alapján, pseudotime analízis -->
<!-- # és CellChat vagy CellPhoneDB analízis is -> do CellChat: https://github.com/jinworks/CellChat1 -->

Folder structure
```{r}
Datafolder     <- here("Data")
Data_calc <- here("Data_calculated") # Path to find Part 1 and 2 outputs

# Subdirectories specifically isolated for this run's outputs
Figures   <- here(file.path("Figures", "diffusion_plot"))
Data_calc <- here("Data_calculated")
Data_public <- here("Data_public")
Results   <- here("Results")

if (!file.exists(Figures))   dir.create(Figures, recursive = TRUE)
if (!file.exists(Data_calc)) dir.create(Data_calc, recursive = TRUE)
if (!file.exists(Results))   dir.create(Results, recursive = TRUE)

utils_file <- here("Rscripts","utils.R")
if (file.exists(utils_file)) source(utils_file)
```

Config parameters and marker lists
```{r}
config_file <- here("Rscripts","config.R")
source(config_file)
```


# Parameters 
```{r}
# to always have the same parameters 
set.seed(2)

# Specific for HPC to make plots
options(bitmapType = "cairo")

# ct <- "kolf2"
ct <- "EpiC"
day <- "D42_hTH"
```

# Diffusion plot 1
TO DO: use same colors in 3D plot as in other plots
```{r}
start_cluster_lookup <- list(
  "kolf2" = "late SCP",      
  "EpiC"  = "late SCP"
)

# possible days: D26, D42, D42_hTH
#day_options <- c("D26", "D42", "D42_hTH", "all", "healthy")
day <- "D42_hTH"
plot2d_list <- list("kolf2"=list(), "EpiC"=list())
slingshot_plot_list  <- list("kolf2"=list(), "EpiC"=list())

outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
seu.integrated <- LoadSeuratRds(outfile)
# for the nice plots, remove "type_12" cluster
seu.integrated <- subset(seu.integrated, subset = !cluster_type %in% c("type_12"))
seu.integrated$cluster_type <- droplevels(seu.integrated$cluster_type)
# write out cell numbers
table(seu.integrated$sample.names, seu.integrated$cluster_type)
    
  # now select the sample
  #for (day in day_options){
    # initialize lists
    plot2d_list[[ct]][[day]] <- list()
    slingshot_plot_list[[ct]][[day]] <- list()
    seu.day <- subset_seu_by_day(seu.integrated, day)
    
    # drop unused levels - do it everywhere!!!
    cluster_sizes <- table(seu.day$cluster_type)
    keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
    seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
    if (day=="D42_hTH"){
      seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
    }
    
    # set up the order
    bio_order <- bio_order_list[[ct]]
    seu.day$cluster_type <- droplevels(seu.day$cluster_type)
   
    # use new names
    actual_day <- limma::strsplit2(day, "_")[[1]]
    
    # Rename clusters from previous annotation
    if (actual_day %in% names(cluster_names_dict[[ct]])){
          new_cluster_ids <- dplyr::recode(
        as.character(seu.day$cluster_type),
        !!!cluster_names_dict[[ct]][[actual_day]]
      )
      # Preserve cluster order
      renamed_cluster_order <- unique(dplyr::recode(
        bio_order,
        !!!cluster_names_dict[[ct]][[actual_day]]
      ))
      
      new_cluster_ids <- factor(
        new_cluster_ids,
        levels = renamed_cluster_order
      )
      
      # Store back into Seurat object and construct final legend text
      seu.day$cluster_type_renamed <- new_cluster_ids
      Idents(seu.day) <- "cluster_type_renamed"
    }
    
    # drop unused levels
    seu.day$cluster_type_renamed <- droplevels(seu.day$cluster_type_renamed)
    print(table(seu.day$cluster_type_renamed))
      
      # Process the results
      cat("\nProcessing Day:", day, "\n")
      print(table(seu.day$sample.names))
      
      # select desired clusters
      # seu.day <- switch(ct,
      #                   "kolf2" = subset(seu.day, subset= num_id_D42_hTH %in% c("3", "4", "5", "7") ),
      #                   "EpiC"= subset(seu.day, subset= num_id_D42_hTH %in% c("0", "1", "7", "8"))
      # )
      
      seu.day <- switch(ct,
                        "kolf2" = subset(seu.day, subset= cluster_type_renamed %in% c("early SCP", "late SCP", "chromaffin", "sympathoblast", "bridging", "EGFP ADRN-like") ),
                        "EpiC"= subset(seu.day, subset= cluster_type_renamed%in% c("late SCP", "chromaffin/symphatoblast", "sympathoblast", "bridging", "EGFP ADRN-like"))
      )
      #table(seu.day$cluster_type_renamed_D42_hTH)
      
      
      # drop unused levels
      seu.day@meta.data <- droplevels(seu.day@meta.data)
      table(seu.day$cluster_type)
      
      # need to be a matrix
      sce <- as.SingleCellExperiment(seu.day)
      
      # warning is not a problem - diffusion map will be run on the pca component
      # other warnings -> check!
      all(rownames(seu.day@reductions$pca@cell.embeddings) == colnames(seu.day))
      dim(Embeddings(seu.day, "pca")[,1:20])
      
      # Dim 18
      dm_S <- DiffusionMap(Embeddings(seu.day, "pca")[,1:18])
      
      #cellLabels <- sce$cluster_type
      cellLabels <- sce$cluster_type_renamed
      tmp <- data.frame(DC1 = eigenvectors(dm_S)[, 1],
                        DC2 = eigenvectors(dm_S)[, 2],
                        DC3 = eigenvectors(dm_S)[, 3],
                        DC4 = eigenvectors(dm_S)[, 4],
                        Clusters = cellLabels)
      
      p2d <- ggplot(tmp, aes(x = DC1, y = DC2, color = Clusters)) +
        geom_point()  + 
        xlab("Diffusion component 1") + 
        ylab("Diffusion component 2") +
        theme_classic() 
      plot2d_list[[ct]][[day]] <- p2d
      ggsave(file.path(Figures, paste("Diffusion_plot_", ct, "_day_", day,
                            "_nHVG", nHVG_str,".pdf", sep="")), p2d, width = 6,   height = 6)
      
      # 1. Create the 3D plot
      p3d <- plot_ly(tmp, x = ~DC1, y = ~DC2, z = ~DC3, 
                     color = ~Clusters, 
                     colors = "Set1", # You can change this to any color palette
                     type = "scatter3d", 
                     mode = "markers",
                     marker = list(size = 7)) %>%
        layout(scene = list(xaxis = list(title = 'Diffusion Component 1'),
                            yaxis = list(title = 'Diffusion Component 2'),
                            zaxis = list(title = 'Diffusion Component 3')),
                             font = list(size = 14),              
                                  legend = list(
                                    font = list(size = 18),            # Legend font size
                                    itemsizing = "constant"))
      # 2. Save the plot as an interactive HTML file
      htmlwidgets::saveWidget(as_widget(p3d), 
                             file.path(Figures, paste0("Diffusion_plot_3D_", ct, "_day_", day, ".html")))
      
      # 3. If you want to view it immediately in RStudio:
      # p3d
      
      
      # add pseudotime analysis
      # Add DC components in the seurat object - I have worked with your diffusion map components because of the two nice lineages 
    dimentions_DC <- cbind(dm_S$DC1, dm_S$DC2, dm_S$DC3, dm_S$DC4)
    #head(dimentions_DC)
    colnames(dimentions_DC) <- c("Diffusion Component 1", 
                                 "Diffusion Component 2", 
                                 "Diffusion Component 3",
                                 "Diffusion Component 4")
    
    seu.day[["DC"]] <- CreateDimReducObject(embeddings = dimentions_DC, key = "DC_", assay = DefaultAssay(seu.day))
    DimPlot(seu.day, reduction = "DC")

    # transform again to add the new metadata
    sce <- as.SingleCellExperiment(seu.day)
    current_start <- start_cluster_lookup[[ct]]
    # slingshot function with DC as reduction 
    # sce_slingshot <- slingshot(sce, clusterLabels = "cluster_type", reducedDim = "DC", 
    #                            start.clus = "type_early_prog", omega = TRUE)
    sce_slingshot <- slingshot(sce, 
                               #clusterLabels = "cluster_type", 
                               clusterLabels = "cluster_type_renamed",
                               reducedDim = "DC", 
                               start.clus = current_start, # Fixed here!
                               omega = TRUE)
    num_lineages <- length(SlingshotDataSet(sce_slingshot)@lineages)
    print(num_lineages)
    # print summary
    SlingshotDataSet(sce_slingshot)
    
    #summary(sce_slingshot$slingPseudotime_1)
    
    # Make this avaible for a ggplot! 
    df_sce <- as.data.frame(reducedDims(sce_slingshot)$DC)
    
    # plot in a loop
    for (i in 1:num_lineages) {
  
      # Dynamically pull the pseudotime vector for lineage i
      pseudotime_col <- paste0("slingPseudotime_", i)
      df_sce$current_pseudotime <- sce_slingshot[[pseudotime_col]]
      
      # add clusters as well
      #df_sce$Cluster <- sce_slingshot$cluster_type
      #df_sce$Cluster <- sce_slingshot$cluster_type
      df_sce$Cluster <- sce_slingshot$cluster_type_renamed
      
            # --- NEW: Calculate and print summary statistics ---
      stats <- df_sce %>%
        filter(!is.na(current_pseudotime)) %>% # Exclude cells not on this lineage
        group_by(Cluster) %>%
        summarise(
          Mean_Pseudotime   = mean(current_pseudotime),
          Median_Pseudotime = median(current_pseudotime),
          SD_Pseudotime     = sd(current_pseudotime),
          Cell_Count        = n()
        )
      
      cat("\n=========================================\n")
      cat("  Summary Stats for Lineage", i, "(", ct, "-", day, ")\n")
      cat("=========================================\n")
      print(stats)
      
      # Generate the plot
      p <- ggplot(df_sce, aes(x = DC_1, y = DC_2, color = current_pseudotime)) + 
        geom_point() + 
        theme_classic() + 
        scale_color_viridis(na.value = "lightgray") + # Cells not on this lineage turn gray
        ggtitle(paste("Pseudotime Lineage", i)) +
        xlab("Diffusion component 1") +
        ylab("Diffusion component 2") +
        labs(color = "Pseudotime")
      
      # # Save the plot dynamically
      # file_name <- paste0("Slingshot_plot_Lineage", i, "_", ct, "_day_", day, "_nHVG", nHVG_str, ".pdf")
      # ggsave(file.path(Figures, file_name), p, width = 6, height = 6)
      
      # Store it in your nested tracking list dynamically
      slingshot_plot_list[[ct]][[day]][[paste0("lineage_", i)]] <- p
      
      # p_box <- ggplot(subset(df_sce, !is.na(current_pseudotime)), 
      #             aes(x = Cluster, y = current_pseudotime)) +
      #   geom_boxplot(aes(fill = Cluster), alpha = 0.7, show.legend = FALSE) +
      #   theme_bw() +
      #   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      #   labs(y = "Pseudotime", x = "Cluster", title = paste("Lineage", i, "Distribution"))
      # 
      # # Print it to the screen as it loops through
      # plot(p_box)
      # slingshot_plot_list[[ct]][[day]][["boxplot"]] <- p_box
      
    }
    # add names of clusters
    slingshot_plot_list[[ct]][[day]][["clusters"]] <- plot2d_list[[ct]][[day]]
    # plot the lineage plots together as well
    combined_plot <-  wrap_plots(slingshot_plot_list[[ct]][[day]], nrow = 1) + 
    plot_annotation(
      #title = paste("Slingshot Pseudotime Lineages -", ct, day),
      #subtitle = paste("Total lineages found:", num_lineages),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
    # Save the master combined plot
    combined_file_name <- paste0("Slingshot_combined_", ct, "_day_", day, "_nHVG", nHVG_str, ".pdf")
    ggsave(
      filename = file.path(Figures, combined_file_name), 
      plot = combined_plot, 
      # width = 10, 
      # height = ceiling((num_lineages+1) / 2) * 4.5
      width = 5*(num_lineages+1), 
      height = 4.5
    )
  #} # loop for day
#} # loop for cell type

```



# Diffusion plot 2
```{r}
start_cluster_lookup <- list(
  "kolf2" = "early endoneural fibroblast",      
  "EpiC"  = "late endoneural fibroblast"
)

# possible days: D26, D42, D42_hTH
#day_options <- c("D26", "D42", "D42_hTH", "all", "healthy")
day <- "D42_hTH"
plot2d_list <- list("kolf2"=list(), "EpiC"=list())
slingshot_plot_list  <- list("kolf2"=list(), "EpiC"=list())

outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
seu.integrated <- LoadSeuratRds(outfile)
# for the nice plots, remove "type_12" cluster
seu.integrated <- subset(seu.integrated, subset = !cluster_type %in% c("type_12"))
seu.integrated$cluster_type <- droplevels(seu.integrated$cluster_type)
# write out cell numbers
table(seu.integrated$sample.names, seu.integrated$cluster_type)
    
  # now select the sample
  #for (day in day_options){
    # initialize lists
    plot2d_list[[ct]][[day]] <- list()
    slingshot_plot_list[[ct]][[day]] <- list()
    seu.day <- subset_seu_by_day(seu.integrated, day)
    
        # drop unused levels - do it everywhere!!!
    cluster_sizes <- table(seu.day$cluster_type)
    keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
    seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
    if (day=="D42_hTH"){
      seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
    }
    
    # set up the order
    bio_order <- bio_order_list[[ct]]
    seu.day$cluster_type <- droplevels(seu.day$cluster_type)
   
    # use new names
    actual_day <- limma::strsplit2(day, "_")[[1]]
    
    # Rename clusters from previous annotation
    if (actual_day %in% names(cluster_names_dict[[ct]])){
          new_cluster_ids <- dplyr::recode(
        as.character(seu.day$cluster_type),
        !!!cluster_names_dict[[ct]][[actual_day]]
      )
      # Preserve cluster order
      renamed_cluster_order <- unique(dplyr::recode(
        bio_order,
        !!!cluster_names_dict[[ct]][[actual_day]]
      ))
      
      new_cluster_ids <- factor(
        new_cluster_ids,
        levels = renamed_cluster_order
      )
      
      # Store back into Seurat object and construct final legend text
      seu.day$cluster_type_renamed <- new_cluster_ids
      Idents(seu.day) <- "cluster_type_renamed"
    }
    
    # drop unused levels
    seu.day$cluster_type_renamed <- droplevels(seu.day$cluster_type_renamed)
    print(table(seu.day$cluster_type_renamed))
      
      # Process the results
      cat("\nProcessing Day:", day, "\n")
      print(table(seu.day$sample.names))
      
      # select desired clusters
      # seu.day <- switch(ct,
      #                   "kolf2" = subset(seu.day, subset= num_id_D42_hTH %in% c("3", "4", "5", "7") ),
      #                   "EpiC"= subset(seu.day, subset= num_id_D42_hTH %in% c("0", "1", "7", "8"))
      # )
      
      seu.day <- switch(ct,
                        "kolf2" = subset(seu.day, subset= cluster_type_renamed %in% c("early endoneural fibroblast", "late endoneural fibroblast", "EGFP MES-like") ),
                        "EpiC"= subset(seu.day, subset= cluster_type_renamed%in% c("late myofibroblast", "late endoneural fibroblast", "EGFP MES-like"))
      )
      #table(seu.day$cluster_type_renamed_D42_hTH)
      
      
      # drop unused levels
      seu.day@meta.data <- droplevels(seu.day@meta.data)
      table(seu.day$cluster_type)
      
      # need to be a matrix
      sce <- as.SingleCellExperiment(seu.day)
      
      # warning is not a problem - diffusion map will be run on the pca component
      # other warnings -> check!
      all(rownames(seu.day@reductions$pca@cell.embeddings) == colnames(seu.day))
      dim(Embeddings(seu.day, "pca")[,1:20])
      
      # Dim 18
      dm_S <- DiffusionMap(Embeddings(seu.day, "pca")[,1:18])
      
      #cellLabels <- sce$cluster_type
      cellLabels <- sce$cluster_type_renamed
      tmp <- data.frame(DC1 = eigenvectors(dm_S)[, 1],
                        DC2 = eigenvectors(dm_S)[, 2],
                        DC3 = eigenvectors(dm_S)[, 3],
                        DC4 = eigenvectors(dm_S)[, 4],
                        Clusters = cellLabels)
      
      p2d <- ggplot(tmp, aes(x = DC1, y = DC2, color = Clusters)) +
        geom_point()  + 
        xlab("Diffusion component 1") + 
        ylab("Diffusion component 2") +
        theme_classic() 
      plot2d_list[[ct]][[day]] <- p2d
      ggsave(file.path(Figures, paste("Diffusion_plot_MES_", ct, "_day_", day,
                            "_nHVG", nHVG_str,".pdf", sep="")), p2d, width = 6,   height = 6)
      
      # 1. Create the 3D plot
      p3d <- plot_ly(tmp, x = ~DC1, y = ~DC2, z = ~DC3, 
                     color = ~Clusters, 
                     colors = "Set1", # You can change this to any color palette
                     type = "scatter3d", 
                     mode = "markers",
                     marker = list(size = 2)) %>%
        layout(scene = list(xaxis = list(title = 'Diffusion Component 1'),
                            yaxis = list(title = 'Diffusion Component 2'),
                            zaxis = list(title = 'Diffusion Component 3')))
      
      # 2. Save the plot as an interactive HTML file
      htmlwidgets::saveWidget(as_widget(p3d), 
                             file.path(Figures, paste0("Diffusion_plot_MES_3D_", ct, "_day_", day, ".html")))
      
      # 3. If you want to view it immediately in RStudio:
      # p3d
      
      
      # add pseudotime analysis
      # Add DC components in the seurat object - I have worked with your diffusion map components because of the two nice lineages 
    dimentions_DC <- cbind(dm_S$DC1, dm_S$DC2, dm_S$DC3, dm_S$DC4)
    #head(dimentions_DC)
    colnames(dimentions_DC) <- c("Diffusion Component 1", 
                                 "Diffusion Component 2", 
                                 "Diffusion Component 3",
                                 "Diffusion Component 4")
    
    seu.day[["DC"]] <- CreateDimReducObject(embeddings = dimentions_DC, key = "DC_", assay = DefaultAssay(seu.day))
    DimPlot(seu.day, reduction = "DC")

    # transform again to add the new metadata
    sce <- as.SingleCellExperiment(seu.day)
    current_start <- start_cluster_lookup[[ct]]
    # slingshot function with DC as reduction 
    # sce_slingshot <- slingshot(sce, clusterLabels = "cluster_type", reducedDim = "DC", 
    #                            start.clus = "type_early_prog", omega = TRUE)
    sce_slingshot <- slingshot(sce, 
                               #clusterLabels = "cluster_type", 
                               clusterLabels = "cluster_type_renamed",
                               reducedDim = "DC", 
                               start.clus = current_start, # Fixed here!
                               omega = TRUE)
    num_lineages <- length(SlingshotDataSet(sce_slingshot)@lineages)
    print(num_lineages)
    # print summary
    SlingshotDataSet(sce_slingshot)
    
    #summary(sce_slingshot$slingPseudotime_1)
    
    # Make this avaible for a ggplot! 
    df_sce <- as.data.frame(reducedDims(sce_slingshot)$DC)
    
    # plot in a loop
    for (i in 1:num_lineages) {
  
      # Dynamically pull the pseudotime vector for lineage i
      pseudotime_col <- paste0("slingPseudotime_", i)
      df_sce$current_pseudotime <- sce_slingshot[[pseudotime_col]]
      
      # add clusters as well
      #df_sce$Cluster <- sce_slingshot$cluster_type
      #df_sce$Cluster <- sce_slingshot$cluster_type
      df_sce$Cluster <- sce_slingshot$cluster_type_renamed
      
            # --- NEW: Calculate and print summary statistics ---
      stats <- df_sce %>%
        filter(!is.na(current_pseudotime)) %>% # Exclude cells not on this lineage
        group_by(Cluster) %>%
        summarise(
          Mean_Pseudotime   = mean(current_pseudotime),
          Median_Pseudotime = median(current_pseudotime),
          SD_Pseudotime     = sd(current_pseudotime),
          Cell_Count        = n()
        )
      
      cat("\n=========================================\n")
      cat("  Summary Stats for Lineage", i, "(", ct, "-", day, ")\n")
      cat("=========================================\n")
      print(stats)
      
      # Generate the plot
      p <- ggplot(df_sce, aes(x = DC_1, y = DC_2, color = current_pseudotime)) + 
        geom_point() + 
        theme_classic() + 
        scale_color_viridis(na.value = "lightgray") + # Cells not on this lineage turn gray
        ggtitle(paste("Pseudotime Lineage", i)) +
        xlab("Diffusion component 1") +
        ylab("Diffusion component 2") +
        labs(color = "Pseudotime")
      
      # # Save the plot dynamically
      # file_name <- paste0("Slingshot_plot_Lineage", i, "_", ct, "_day_", day, "_nHVG", nHVG_str, ".pdf")
      # ggsave(file.path(Figures, file_name), p, width = 6, height = 6)
      
      # Store it in your nested tracking list dynamically
      slingshot_plot_list[[ct]][[day]][[paste0("lineage_", i)]] <- p
      
      # p_box <- ggplot(subset(df_sce, !is.na(current_pseudotime)), 
      #             aes(x = Cluster, y = current_pseudotime)) +
      #   geom_boxplot(aes(fill = Cluster), alpha = 0.7, show.legend = FALSE) +
      #   theme_bw() +
      #   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      #   labs(y = "Pseudotime", x = "Cluster", title = paste("Lineage", i, "Distribution"))
      # 
      # # Print it to the screen as it loops through
      # plot(p_box)
      # slingshot_plot_list[[ct]][[day]][["boxplot"]] <- p_box
      
    }
    # add names of clusters
    slingshot_plot_list[[ct]][[day]][["clusters"]] <- plot2d_list[[ct]][[day]]
    # plot the lineage plots together as well
    combined_plot <-  wrap_plots(slingshot_plot_list[[ct]][[day]], nrow = 1) + 
    plot_annotation(
      #title = paste("Slingshot Pseudotime Lineages -", ct, day),
      #subtitle = paste("Total lineages found:", num_lineages),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
    # Save the master combined plot
    combined_file_name <- paste0("Slingshot_combined_MES_", ct, "_day_", day, "_nHVG", nHVG_str, ".pdf")
    ggsave(
      filename = file.path(Figures, combined_file_name), 
      plot = combined_plot, 
      # width = 10, 
      # height = ceiling((num_lineages+1) / 2) * 4.5
      width = 5*(num_lineages+1), 
      height = 4.5
    )
  #} # loop for day
#} # loop for cell type

```


Repeat with 
H1 & H2 DIV42  hTH all clusters
H1 & H2 DIV42 both all clusters
H1 & H2 all
Here we will mix up the parameters


# Diffusion plot 3
```{r}
start_cluster_lookup <- list(
  "kolf2" =  list( "D42_hTH"  ="early SCP",    
                   "D42_both"  ="early SCP", 
                   "all"  ="early SCP"),
  "EpiC"  =  list( "D42_hTH"  ="late SCP",    
                   "D42_both"  ="late SCP", 
                   "all"  ="early/late SCP")
)

for (ct in c("kolf2", "EpiC")){
  for (day in c("D42_hTH", "D42_both", "all")){
    # possible days: D26, D42, D42_hTH
  plot2d_list <- list("kolf2"=list(), "EpiC"=list())
  slingshot_plot_list  <- list("kolf2"=list(), "EpiC"=list())
  
  outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
  seu.integrated <- LoadSeuratRds(outfile)
  # for the nice plots, remove "type_12" cluster
  seu.integrated <- subset(seu.integrated, subset = !cluster_type %in% c("type_12"))
  seu.integrated$cluster_type <- droplevels(seu.integrated$cluster_type)
  # write out cell numbers
  table(seu.integrated$sample.names, seu.integrated$cluster_type)
    
  # now select the sample
  #for (day in day_options){
    # initialize lists
    plot2d_list[[ct]][[day]] <- list()
    slingshot_plot_list[[ct]][[day]] <- list()
    seu.day <- subset_seu_by_day(seu.integrated, day)
    
        # drop unused levels - do it everywhere!!!
    cluster_sizes <- table(seu.day$cluster_type)
    keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
    seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
    if (day=="D42_hTH"){
      seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
    }
    
    # set up the order
    bio_order <- bio_order_list[[ct]]
    seu.day$cluster_type <- droplevels(seu.day$cluster_type)
   
    # use new names
    actual_day <- limma::strsplit2(day, "_")[[1]]
    
    # Rename clusters from previous annotation
    if (actual_day %in% names(cluster_names_dict[[ct]])){
          new_cluster_ids <- dplyr::recode(
        as.character(seu.day$cluster_type),
        !!!cluster_names_dict[[ct]][[actual_day]]
      )
      # Preserve cluster order
      renamed_cluster_order <- unique(dplyr::recode(
        bio_order,
        !!!cluster_names_dict[[ct]][[actual_day]]
      ))
      
      new_cluster_ids <- factor(
        new_cluster_ids,
        levels = renamed_cluster_order
      )
      
      # Store back into Seurat object and construct final legend text
      seu.day$cluster_type_renamed <- new_cluster_ids
      Idents(seu.day) <- "cluster_type_renamed"
    }
    
    # drop unused levels
    seu.day$cluster_type_renamed <- droplevels(seu.day$cluster_type_renamed)
    print(table(seu.day$cluster_type_renamed))
      
      # Process the results
      cat("\nProcessing Day:", day, "\n")
      print(table(seu.day$sample.names))

      
      # need to be a matrix
      sce <- as.SingleCellExperiment(seu.day)
      
      # warning is not a problem - diffusion map will be run on the pca component
      # other warnings -> check!
      all(rownames(seu.day@reductions$pca@cell.embeddings) == colnames(seu.day))
      dim(Embeddings(seu.day, "pca")[,1:20])
      
      # Dim 18
      dm_S <- DiffusionMap(Embeddings(seu.day, "pca")[,1:18])
      
      #cellLabels <- sce$cluster_type
      cellLabels <- sce$cluster_type_renamed
      tmp <- data.frame(DC1 = eigenvectors(dm_S)[, 1],
                        DC2 = eigenvectors(dm_S)[, 2],
                        DC3 = eigenvectors(dm_S)[, 3],
                        DC4 = eigenvectors(dm_S)[, 4],
                        Clusters = cellLabels)
      
      p2d <- ggplot(tmp, aes(x = DC1, y = DC2, color = Clusters)) +
        geom_point()  + 
        xlab("Diffusion component 1") + 
        ylab("Diffusion component 2") +
        theme_classic() 
      plot2d_list[[ct]][[day]] <- p2d
      ggsave(file.path(Figures, paste("Diffusion_plot_allclusters_", ct, "_day_", day,
                            "_nHVG", nHVG_str,".pdf", sep="")), p2d, width = 6,   height = 6)
      
      # 1. Create the 3D plot
      p3d <- plot_ly(tmp, x = ~DC1, y = ~DC2, z = ~DC3, 
                     color = ~Clusters, 
                     colors = "Set1", # You can change this to any color palette
                     type = "scatter3d", 
                     mode = "markers",
                     marker = list(size = 2)) %>%
        layout(scene = list(xaxis = list(title = 'Diffusion Component 1'),
                            yaxis = list(title = 'Diffusion Component 2'),
                            zaxis = list(title = 'Diffusion Component 3')))
      
      # 2. Save the plot as an interactive HTML file
      htmlwidgets::saveWidget(as_widget(p3d), 
                             file.path(Figures, paste0("Diffusion_plot_allclusters_3D_", ct, "_day_", day, ".html")))
      
      # 3. If you want to view it immediately in RStudio:
      # p3d
      
      
      # add pseudotime analysis
      # Add DC components in the seurat object - I have worked with your diffusion map components because of the two nice lineages 
    dimentions_DC <- cbind(dm_S$DC1, dm_S$DC2, dm_S$DC3, dm_S$DC4)
    #head(dimentions_DC)
    colnames(dimentions_DC) <- c("Diffusion Component 1", 
                                 "Diffusion Component 2", 
                                 "Diffusion Component 3",
                                 "Diffusion Component 4")
    
    seu.day[["DC"]] <- CreateDimReducObject(embeddings = dimentions_DC, key = "DC_", assay = DefaultAssay(seu.day))
    DimPlot(seu.day, reduction = "DC")

    # transform again to add the new metadata
    sce <- as.SingleCellExperiment(seu.day)
    current_start <- start_cluster_lookup[[ct]][[day]]
    # slingshot function with DC as reduction 
    # sce_slingshot <- slingshot(sce, clusterLabels = "cluster_type", reducedDim = "DC", 
    #                            start.clus = "type_early_prog", omega = TRUE)
    sce_slingshot <- slingshot(sce, 
                               #clusterLabels = "cluster_type", 
                               clusterLabels = "cluster_type_renamed",
                               reducedDim = "DC", 
                               start.clus = current_start, # Fixed here!
                               #omega = TRUE
                               omega = 1.1)
    
    # # drop any lineage where an intermediate step relies on fewer than 10 cells:
    # # Check cell counts per cluster across each lineage
    # sds <- SlingshotDataSet(sce_slingshot)
    # lineages <- sds@lineages
    # 
    # # Keep lineages where all intermediate/terminal clusters have >= 10 cells
    # valid_lineages <- c()
    # for (i in seq_along(lineages)) {
    #   lin_clusters <- lineages[[i]]
    #   # Calculate cell count for each cluster in this lineage
    #   counts <- table(sce_slingshot$cluster_type_renamed[sce_slingshot$cluster_type_renamed %in% lin_clusters])
    #   
    #   if (all(counts >= 10)) {
    #     valid_lineages <- c(valid_lineages, i)
    #   }
    # }
    # cat("Valid Lineages with strong transition support:", valid_lineages, "\n")

    
    num_lineages <- length(SlingshotDataSet(sce_slingshot)@lineages)
    print(num_lineages)
    # print summary
    SlingshotDataSet(sce_slingshot)
    
    #summary(sce_slingshot$slingPseudotime_1)
    
    # Make this avaible for a ggplot! 
    df_sce <- as.data.frame(reducedDims(sce_slingshot)$DC)
    
    # plot in a loop
    for (i in 1:num_lineages) {
  
      # Dynamically pull the pseudotime vector for lineage i
      pseudotime_col <- paste0("slingPseudotime_", i)
      df_sce$current_pseudotime <- sce_slingshot[[pseudotime_col]]
      
      # add clusters as well
      #df_sce$Cluster <- sce_slingshot$cluster_type
      #df_sce$Cluster <- sce_slingshot$cluster_type
      df_sce$Cluster <- sce_slingshot$cluster_type_renamed
      
            # --- NEW: Calculate and print summary statistics ---
      stats <- df_sce %>%
        filter(!is.na(current_pseudotime)) %>% # Exclude cells not on this lineage
        group_by(Cluster) %>%
        summarise(
          Mean_Pseudotime   = mean(current_pseudotime),
          Median_Pseudotime = median(current_pseudotime),
          SD_Pseudotime     = sd(current_pseudotime),
          Cell_Count        = n()
        )
      
      cat("\n=========================================\n")
      cat("  Summary Stats for Lineage", i, "(", ct, "-", day, ")\n")
      cat("=========================================\n")
      print(stats)
      
      # Generate the plot
      p <- ggplot(df_sce, aes(x = DC_1, y = DC_2, color = current_pseudotime)) + 
        geom_point() + 
        theme_classic() + 
        scale_color_viridis(na.value = "lightgray") + # Cells not on this lineage turn gray
        ggtitle(paste("Pseudotime Lineage", i)) +
        xlab("Diffusion component 1") +
        ylab("Diffusion component 2") +
        labs(color = "Pseudotime")
      
      # # Save the plot dynamically
      # file_name <- paste0("Slingshot_plot_Lineage", i, "_", ct, "_day_", day, "_nHVG", nHVG_str, ".pdf")
      # ggsave(file.path(Figures, file_name), p, width = 6, height = 6)
      
      # Store it in your nested tracking list dynamically
      slingshot_plot_list[[ct]][[day]][[paste0("lineage_", i)]] <- p
      
      # p_box <- ggplot(subset(df_sce, !is.na(current_pseudotime)), 
      #             aes(x = Cluster, y = current_pseudotime)) +
      #   geom_boxplot(aes(fill = Cluster), alpha = 0.7, show.legend = FALSE) +
      #   theme_bw() +
      #   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      #   labs(y = "Pseudotime", x = "Cluster", title = paste("Lineage", i, "Distribution"))
      # 
      # # Print it to the screen as it loops through
      # plot(p_box)
      # slingshot_plot_list[[ct]][[day]][["boxplot"]] <- p_box
      
    }
    # add names of clusters
    slingshot_plot_list[[ct]][[day]][["clusters"]] <- plot2d_list[[ct]][[day]]
    # plot the lineage plots together as well
    combined_plot <-  wrap_plots(slingshot_plot_list[[ct]][[day]], nrow = 1) + 
    plot_annotation(
      #title = paste("Slingshot Pseudotime Lineages -", ct, day),
      #subtitle = paste("Total lineages found:", num_lineages),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
    # Save the master combined plot
    combined_file_name <- paste0("Slingshot_combined_allclusters_", ct, "_day_", day, "_nHVG", nHVG_str, ".pdf")
    ggsave(
      filename = file.path(Figures, combined_file_name), 
      plot = combined_plot, 
      # width = 10, 
      # height = ceiling((num_lineages+1) / 2) * 4.5
      width = 5*(num_lineages+1), 
      height = 4.5
    )
    
    
    # remove artificial lineages
    weights <- slingCurveWeights(sce_slingshot)
    min_cell_threshold <- 15
    valid_lineages <- which(colSums(weights > 0.5) >= min_cell_threshold)
    
  }  # loop for day
} # loop for cell type



```



