---
title: "Analysis of single cell neuroblastoma organoids: part 3 - selected resolutions"
author: "Anna Lovrics"
date: "`r Sys.Date()`"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
---
```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.width=12, fig.height=5)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).
The objective of this R Markdown code is to analyse the results obtained by the organoid single cell RNA sequencing.

```{r libraries}

packages <- c(
  "Seurat",
  "ggplot2",
  "patchwork",    
  "Matrix",
  "dplyr",
  "here",
  "stringr",      
  "future",       
  "clustree",     
  "RColorBrewer",
  "ggforce",
  "concaveman",
  "openxlsx",
  "viridis",      
  "msigdbr"   
)
invisible(lapply(packages, library, character.only = TRUE))

```


<!-- # Biztosan kellene sejt -->
<!-- # populáció/cluster azonosítás marker gének alapján, pseudotime analízis -->
<!-- # és CellChat vagy CellPhoneDB analízis is -> do CellChat: https://github.com/jinworks/CellChat1 -->

Folder structure is defined first
```{r}
Datafolder     <- here("Data")
# Subdirectories specifically isolated for this run's outputs
Figures   <- here(file.path("Figures", "selected_resolution"))
Figures_dimplots <- file.path(Figures, "dimplots")
Data_calc <- here("Data_calculated")
Results   <- here("Results")

if (!file.exists(Figures))   dir.create(Figures, recursive = TRUE)
if (!file.exists(Figures_dimplots))   dir.create(Figures_dimplots, recursive = TRUE)
if (!file.exists(Data_calc)) dir.create(Data_calc, recursive = TRUE)
if (!file.exists(Results))   dir.create(Results, recursive = TRUE)

utils_file <- here("Rscripts","utils.R")
if (file.exists(utils_file)) source(utils_file)
```


Config parameters and marker lists
```{r}
config_file <- here("Rscripts","config.R")
source(config_file)

# Select cell line
ct <- "EpiC"
# ct <- "kolf2"
```


# Integrated UMAP Visualizations
Below are the UMAP plots for the integrated cell line specific data, split by sample name. Convex hulls are included to better delineate the clusters. 
```{r EpiC-UMAP-Visualizations, fig.width=15, fig.height=10}

outfile <- file.path(Results, paste("seu_integrated_clusters_newnames_", ct, "_all_nHVG", nHVG_str, ".rds", sep=""))
seu.integrated <- LoadSeuratRds(file=outfile)

# for the nice plots, remove "type_12" cluster
seu.integrated <- subset(seu.integrated, subset = !cluster_type %in% c("type_12"))
seu.integrated$cluster_type <- droplevels(seu.integrated$cluster_type)

#DimPlot(seu.integrated)
table(seu.integrated$sample.names, seu.integrated$cluster_type)

# to plot van-Haver markers:
# add hulls
my_hulls <- get_glasswork_layers(
    seu.integrated,
    clusvar = "cluster_type",
    concavity = 4,         # Slightly higher concavity follows the dense core better
    core_quantile = 0.80   # This ignores the 20% most isolated cells
  )
names(my_hulls) <- sort(unique(seu.integrated$cluster_type))

my_hulls_outline <- get_glasswork_layers_outline(
    seu.integrated,
    clusvar = "cluster_type",
    concavity = 3,         # Slightly higher concavity follows the dense core better
    core_quantile = 0.75   # This ignores the 25% most isolated cells
  )

# Calculate percentages for each sample
cluster_num <- table(seu.integrated$cluster_type, seu.integrated$sample.names)
cluster_perc <- round(prop.table(cluster_num, margin = 2) * 100, digits=2)

# Calculate coordinates for labels
umap_coords <- seu.integrated@reductions$umap@cell.embeddings
label_coords <- umap_coords %>% as.data.frame() %>%
  mutate(
    cluster_type = seu.integrated$cluster_type,
  ) %>%
  group_by(cluster_type) %>%
  summarize(umap_1 = median(umap_1), umap_2 = median(umap_2), .groups = "drop")

label_coords_num <- umap_coords %>% as.data.frame() %>%
  mutate(
    cluster_num = unlist(seu.integrated[[selres]])
  ) %>%
  group_by(cluster_num) %>%
  summarize(umap_1 = median(umap_1), umap_2 = median(umap_2), .groups = "drop")

```

# Integrated UMAP for each day

```{r UMAP for each day}

  day_options <- c("D26_WT", "D26_STOP_MYCN", "D42_novirus", "D42_hTH")

  for (day in day_options){
    seu.day <- subset_seu_by_day(seu.integrated, day)
        
    # drop unused levels - do it everywhere!!!
      cluster_sizes <- table(seu.day$cluster_type)
      keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
      seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
      if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
        seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
      }
      seu.day$cluster_type <- droplevels(seu.day$cluster_type)
      
      seu.day$num_id <- seu.day[[selres]]
      seu.day$num_id <- droplevels(seu.day$num_id)
      
      # add num_id order; TO DO later: num_id_final - take into account that some clusters appear in earlier days first
      actual_numbers <- as.numeric(as.character(unique(seu.day$num_id)))
      sorted_numeric_levels <- as.character(sort(actual_numbers))
      seu.day$num_id <- factor(seu.day$num_id, levels = sorted_numeric_levels)
        
      # print(day)
      # print(table(seu.day$sample.names))
          
      # use new names
      actual_day <- limma::strsplit2(day, "_")[[1]]
      
      # Rename clusters from previous annotation
      if (actual_day %in% names(cluster_names_dict[[ct]])){
          new_cluster_ids <- dplyr::recode(
          as.character(seu.day$cluster_type),
          !!!cluster_names_dict[[ct]][[actual_day]]
        )
        # Preserve cluster order
        renamed_cluster_order <- unique(dplyr::recode(
          bio_order,
          !!!cluster_names_dict[[ct]][[actual_day]]
        ))
        
        new_cluster_ids <- factor(
          new_cluster_ids,
          levels = unique(renamed_cluster_order)
        )
        
        # Store back into Seurat object and construct final legend text
        seu.day$cluster_type_renamed <- new_cluster_ids
        Idents(seu.day) <- "cluster_type_renamed"
        seu.day$legend_label <- paste0(seu.day$num_id, ": ", seu.day$cluster_type_renamed)
      } else {
          new_cluster_ids <- seu.day$cluster_type
          renamed_cluster_order <- bio_order
          # Store back into Seurat object and construct final legend text
          seu.day$cluster_type_renamed <- new_cluster_ids
          Idents(seu.day) <- "cluster_type_renamed"
          seu.day$legend_label <- paste0(seu.day$num_id, ": ", seu.day$cluster_type_renamed)
      }
  
      # add back to original seurat object (for missing cells, it will be NA) - this is not actually used after all
      meta_names <- c("cluster_type", "legend_label", "num_id", "cluster_type_renamed")
      meta_colnames <- paste(meta_names, 
                         day, sep="_")
      # this is only used in this script and not later - remove?
      meta_toadd <- data.frame(
        seu.day@meta.data[, meta_names],
        row.names = colnames(seu.day)
      )
      colnames(meta_toadd) <- meta_colnames
      seu.integrated <- AddMetaData(
        seu.integrated,
        metadata = meta_toadd
      )
  
      
      # ==========================================
      # 1. CREATE A DUAL-KEY PALETTE FOR TODAY
      # ==========================================
      day_mapping <- unique(as.data.frame(seu.day@meta.data[, c("cluster_type", "legend_label")]))
      
      existing_bio_order <- bio_order[bio_order %in% day_mapping$cluster_type]
      day_mapping <- day_mapping[match(existing_bio_order, day_mapping$cluster_type), ]
      # use same order as in previous figures
      numeric_levels <- stringr::str_sort(unique(day_mapping$legend_label), numeric = TRUE)
      seu.day$legend_label <- factor(seu.day$legend_label, levels = numeric_levels)
      #seu.day$legend_label <- factor(seu.day$legend_label, levels = day_mapping$legend_label)
      
      # Colours for the biological cluster names
      palette_old <- cluster_palette[as.character(day_mapping$cluster_type)]
      names(palette_old) <- as.character(day_mapping$cluster_type)
      
      # Same colours, but named by the legend labels
      palette_new <- cluster_palette[as.character(day_mapping$cluster_type)]
      names(palette_new) <- as.character(day_mapping$legend_label)
      
      day_universal_palette <- c(palette_old, palette_new)
      day_universal_palette <- day_universal_palette[!duplicated(names(day_universal_palette))]
      
      # remove unused hulls
      my_hulls_day <- my_hulls[unique(as.character(seu.day$cluster_type))]
      # drop unused levels
      my_hulls_day <- my_hulls_day[!sapply(my_hulls_day, is.null)]
      
      # ==========================================
      # 1.6 FILTER THE NUMBER LABELS FOR TODAY
      # ==========================================
      # Filter the coordinate dataframe so it only keeps clusters present in seu.day
      label_coords_day <- label_coords_num[label_coords_num$cluster_num %in% unique(seu.day$num_id), ]
      
      # ==========================================
      # 2. PLOT WITH SYNCHRONIZED COLORSETS
      # ==========================================
      p_orig <- DimPlot(
          seu.day,
          cols = day_universal_palette,      # Maps dots using legend_label keys
          ncol = 2, 
          label = FALSE, 
          reduction = "umap",
          group.by = "legend_label",
          split.by  = "sample_names_renamed"
          #pt.size   = 0.1
          ) +
          my_hulls_day +                         # OMITTED the forced aes() line here!
          scale_fill_manual(values = day_universal_palette) +  # Maps hulls using cluster_type keys
          scale_color_manual(values = day_universal_palette) + # Maps hull borders if needed
          theme_void() + 
          theme(
            #legend.position = "top",
            legend.position = "right",
            legend.text = element_text(size = 12),
            strip.text = element_text(size = 14, face = "bold", margin = margin(b = 10)),
            plot.title = element_blank()
          ) +
          guides(
            fill = "none", 
            color = guide_legend(override.aes = list(alpha = 1, size = 4), ncol = 1)
          ) + 
          geom_text(
            data = label_coords_day, 
            mapping = aes(x = umap_1, y = umap_2, label = cluster_num),
            size = 8,
            inherit.aes = FALSE
          )
    
      print(p_orig)
      
      outfig <- file.path(Figures, paste("clusters_samples_newnames_", ct, "_res_", 
              selres, "_day_", day, "_nHVG", nHVG_str, ".pdf", sep=""))
      ggsave(outfig, plot=p_orig, width=8*length(unique(seu.day$sample.names)), height=8)
      ggsave(outfig, plot=p_orig, width=10*length(unique(seu.day$sample.names)), height=6)
      
      # plot with smaller hulls
      # 2 options to draw the hulls
      my_hulls_day <- get_glasswork_layers(
        seu.day,
        clusvar = "cluster_type",
        concavity = 4,         # Slightly higher concavity follows the dense core better
        core_quantile = 0.80   # This ignores the 20% most isolated cells
      )

      p_orig <- DimPlot(
          seu.day,
          cols = day_universal_palette,      # Maps dots using legend_label keys
          ncol = 2,
          label = FALSE,
          reduction = "umap",
          group.by = "legend_label",
          split.by  = "sample_names_renamed",
          pt.size   = 0.1) +
          my_hulls_day +                         # OMITTED the forced aes() line here!
          scale_fill_manual(values = day_universal_palette) +  # Maps hulls using cluster_type keys
          scale_color_manual(values = day_universal_palette) + # Maps hull borders if needed
          theme_void() +
          theme(
            legend.position = "right",
            legend.text = element_text(size = 12),
            strip.text = element_text(size = 14, face = "bold", margin = margin(b = 10)),
            plot.title = element_blank()
          ) +
          guides(
            fill = "none",
            color = guide_legend(override.aes = list(alpha = 1, size = 4), ncol = 1)
          ) +
          geom_text(
            data = label_coords_day,
            mapping = aes(x = umap_1, y = umap_2, label = cluster_num),
            size = 8,
            inherit.aes = FALSE
          )

      print(p_orig)

      outfig <- file.path(Figures, paste("clusters_samples_newnames_smallhulls_", ct, "_res_",
              selres, "_day_", day, "_nHVG", nHVG_str, ".pdf", sep=""))
      #ggsave(outfig, plot=p_orig, width=12, height=12)
      ggsave(outfig, plot=p_orig, width=10*length(unique(seu.day$sample.names)), height=6)
      
  }

```

# Cluster Ordering and Proportions
This section defines the biological ordering of clusters for all subsequent plots and visualizes the cluster distribution across samples using barplots.
Also plot cell cycle scores, mito.percent, stress score and violin plots

```{r EpiC-Ordering-Proportions}

day_options <- c("D26_both", "D42_both")
for (day in day_options){
      seu.day <- subset_seu_by_day(seu.integrated, day)
      
  # drop unused levels - do it everywhere!!!
    cluster_sizes <- table(seu.day$cluster_type)
    keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
    seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
    if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
        seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
    }
    seu.day$cluster_type <- droplevels(seu.day$cluster_type)
  
  # Calculate percentages for each sample
  cluster_num <- table(seu.day$cluster_type, seu.day$sample.names)
  cluster_perc <- round(prop.table(cluster_num, margin = 2) * 100, digits=2)
    # stacked bars plot:
  # Make sure Sample is a factor (for ordering) and Cluster is a factor
  prop_long <- as.data.frame(cluster_perc)
  colnames(prop_long) <- c("Cluster", "Sample", "Percentage")
  prop_long$Cluster <- factor(prop_long$Cluster, levels = bio_order)
  prop_long$Cluster <- droplevels(prop_long$Cluster)
  prop_long$Sample <- factor(prop_long$Sample)
  
  actual_day <- limma::strsplit2(day, "_")[[1]]
  
  # use cluster_names_dict & sample_names_dict
  prop_long$Cluster <- dplyr::recode(
    as.character(prop_long$Cluster),
    !!!cluster_names_dict[[ct]][[actual_day]]
  )
  # also keep order
  prop_long$Cluster <- factor(prop_long$Cluster, levels = unique(prop_long$Cluster))
  
  # rename palette as well
  # my_palette <- my_palette_orig
  my_palette <- cluster_palette
  names(my_palette) <- dplyr::recode(
    names(my_palette),
    !!!cluster_names_dict[[ct]][[actual_day]]
  )
  
  # and recode the sample names
  prop_long$Sample <- dplyr::recode(
    as.character(prop_long$Sample),
    !!!sample_names_dict
  )
  # also keep order
  prop_long$Sample <- factor(prop_long$Sample, levels = unique(prop_long$Sample))
  
  # Plot proportions as a single stacked barplot
  p_prop <- ggplot(prop_long, aes(x = Sample, y = Percentage, fill = Cluster)) +
    geom_bar(stat = "identity", width = 0.7) +
    scale_fill_manual(values = my_palette) +
    theme_bw() +
    labs(y = "Percentage of cells (%)", x = "") +
    theme(
      axis.text.x = element_text(angle = 30, hjust = 1, size = 11),
      #axis.text.x = element_text(size = 14),
      axis.text.y = element_text(size = 14),
      panel.grid.major.x = element_blank(),
      legend.position = "right", # Move legend to the side
      legend.text = element_text(size = 12),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
    )
  
  print(p_prop)
  # Save the proportion plot
  ggsave(file.path(Figures, paste("Cluster_percentages_stacked_", ct, "_selres_", selres, "_day_", day,
                        "_nHVG", nHVG_str, ".pdf", sep="")), p_prop, width = 8,   height = 5)
}

day_options <- c("D42_hTH")
for (day in day_options){
      seu.day <- subset_seu_by_day(seu.integrated, day)
      
  # drop unused levels - do it everywhere!!!
    cluster_sizes <- table(seu.day$cluster_type)
    keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
    seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
    if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
        seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
    }
    seu.day$cluster_type <- droplevels(seu.day$cluster_type)
  
  # Additionally, plot from previous annotation
  ##### first all the renaming:
  new_cluster_ids <- dplyr::recode(
    as.character(Idents(seu.day)),
    !!!cluster_names_dict[[ct]][[actual_day]]
  )
  # Preserve cluster order
  renamed_cluster_order <- unique(dplyr::recode(
    bio_order,
    !!!cluster_names_dict[[ct]][[actual_day]]
  ))
  
  new_cluster_ids <- factor(
    new_cluster_ids,
    levels = unique(renamed_cluster_order)
  )
  
  # Store back into Seurat object
  seu.day$cluster_type_renamed <- new_cluster_ids
  Idents(seu.day) <- "cluster_type_renamed"
  ##### end of renaming
    
  p_qc <- VlnPlot(seu.day, features = c("percent.mito", "StressScore1"),
      cols = my_palette)
  p_qc[[1]] <- p_qc[[1]] + labs(title = "Mitochondrial %")+
    scale_x_discrete(labels = function(x) abbrev_dict[x]) 
  p_qc[[2]] <- p_qc[[2]] + labs(title = "Stress Score")+
    scale_x_discrete(labels = function(x) abbrev_dict[x])
  
  # Standard Color/Theme adjustments
  p_qc <- p_qc & theme(
      axis.title.x = element_blank(),
      axis.text.x = element_text(size = 16, angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = 16),
      plot.title = element_text(size = 20),
      plot.margin = margin(l = 30, r = 10, t = 10, b = 10)
  ) 
  #& scale_x_discrete(limits = rev) # reverse order 
  plot(p_qc)
  ggsave(file.path(Figures, paste("clusters_percent.mito_", ct, "_res_", selres,
                                  "_day_", day, "_nHVG", nHVG_str, ".pdf", sep="")),
         p_qc, width = 15, height = 10)
  
  
  # add cell cycle as well
  
  # --- 1. Cell Cycle Distribution (Proportion Barplot) ---
  cell_cycle_df <- as.data.frame(seu.day@meta.data[, c("cluster_type_renamed", "Phase")])
  cell_cycle_df$cluster_type_renamed <- factor(as.character(cell_cycle_df$cluster_type_renamed),
                                         levels = renamed_cluster_order)
  
  # Color/Standard Version
  p_cc <- ggplot(cell_cycle_df, aes(x = cluster_type_renamed, fill = Phase)) +
      geom_bar(position = "fill") +
      theme_minimal() +
      labs(y = "Proportion", x = NULL) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
            axis.text.y = element_text(size = 14),
            legend.text = element_text(size = 12))
  plot(p_cc)
  ggsave(file.path(Figures, paste("clusters_cellcycle_", ct, "_res_", selres, 
                                  "_day_", day, "_nHVG", nHVG_str, ".pdf", sep="")),
         p_cc, width = 8, height = 6)
  
  
  # plot additional violinplots
  p_vln <- VlnPlot(seu.day, features = violinplot_array, 
                   # split.by = "sample_names_renamed", split.plot = T,
      cols = my_palette)
  
  # Standard Color/Theme adjustments
  p_vln <- p_vln & theme(
      axis.title.x = element_blank(),
      plot.margin = margin(l = 30, r = 10, t = 10, b = 10)
  ) 
  #& scale_x_discrete(limits = rev) # reverse order 
  plot(p_vln)
  ggsave(file.path(Figures, paste("violin_", ct, "_res_", selres,
                                  "_day_", day, "_nHVG", nHVG_str, ".pdf", sep="")),
         p_vln, width = 20, height = 15)
  
 }

```


<!-- # Van-Haver Marker Module Scores -->
<!-- In this section, the expression of each of the Van-Haver marker sets is visulaized as an aggregate score.  -->
<!-- ```{r EpiC-Marker-Plots, fig.width=15, fig.height=10} -->

<!-- day_options <- c( "D26_WT", "D26_STOP_MYCN", "D42_novirus", "D42_hTH") -->
<!-- for (day in day_options){ -->
<!--       seu.day <- subset_seu_by_day(seu.integrated, day) -->

<!--   # drop unused levels - do it everywhere!!! -->
<!--     cluster_sizes <- table(seu.day$cluster_type) -->
<!--     keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters]) -->
<!--     seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters) -->
<!--     if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){ -->
<!--       seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH) -->
<!--     } -->
<!--     seu.day$cluster_type <- droplevels(seu.day$cluster_type) -->


<!--   # ------------------------------------------------------- -->
<!--   # FeaturePlots -->
<!--   # ------------------------------------------------------- -->

<!--   Idents(seu.day) <- paste("num_id", day, sep="_") -->

<!--   pp <- list() -->

<!--   for (marker_name in marker_names) { -->

<!--     pp[[marker_name]] <- FeaturePlot( -->
<!--       seu.day, -->
<!--       label = TRUE, -->
<!--       label.size = 8, -->
<!--       repel = TRUE, -->
<!--       features = marker_name, -->
<!--       reduction = "umap", -->
<!--       min.cutoff = "q05", -->
<!--       max.cutoff = "q95" -->
<!--     ) + -->

<!--       scale_colour_gradient( -->
<!--         low = "lightgrey", -->
<!--         high = "purple", -->
<!--         limits = c(min_val, max_val), -->
<!--         oob = scales::squish -->
<!--       ) + -->

<!--       theme( -->
<!--         plot.title = element_text( -->
<!--           hjust = 0.5, -->
<!--           face = "bold", -->
<!--           size = 16 -->
<!--         ) -->
<!--       ) -->
<!--   } -->

<!--     # ========================================== -->
<!--     # 1. CREATE A DUAL-KEY PALETTE FOR TODAY -->
<!--     # ========================================== -->
<!--     legend_name <- paste("legend_label", day, sep="_") -->
<!--     day_mapping <- unique(as.data.frame(seu.day@meta.data[,  -->
<!--                             c("cluster_type", legend_name)])) -->

<!--     # Map colors to the old cluster names (what the hulls are looking for) -->
<!--     palette_old <- my_palette_orig[as.character(day_mapping$cluster_type)] -->
<!--     names(palette_old) <- as.character(day_mapping$cluster_type) -->

<!--     # Map colors to the new legend labels (what DimPlot is looking for) -->
<!--     palette_new <- my_palette_orig[as.character(day_mapping$cluster_type)] -->
<!--     names(palette_new) <- as.character(day_mapping[[legend_name]]) -->

<!--     # Merge them together so the palette can answer to BOTH names correctly -->
<!--     day_universal_palette <- c(palette_old, palette_new) -->
<!--     day_universal_palette <- day_universal_palette[!duplicated(names(day_universal_palette))] -->

<!--   # ------------------------------------------------------- -->
<!--   # Legend -->
<!--   # ------------------------------------------------------- -->
<!--   legend_order <- sort(unique(seu.day[[legend_name]][,1])) -->
<!--   dummy_data <- data.frame(umap_1 = seq_along(legend_order), umap_2 = 0,  -->
<!--                            #legend_label = unique(seu.day$legend_label) -->
<!--                            legend_label = factor(legend_order, levels = legend_order) -->
<!--                            ) -->
<!--   p_legend <- ggplot(dummy_data, aes(x = umap_1, y = umap_2, color = legend_label)) + -->
<!--   geom_point(alpha = 0) +  -->
<!--   theme_void() + -->
<!--   scale_color_manual(values = day_universal_palette) + -->
<!--   labs(color = "Clusters") +  -->
<!--   theme( -->
<!--     legend.position = "left",  -->
<!--     legend.justification = "left", -->
<!--     legend.text = element_text(size = 12),  -->
<!--     legend.title = element_text(size = 14, face = "bold") -->
<!--   ) + -->
<!--   guides(color = guide_legend(override.aes = list(alpha = 0, size = 0))) -->

<!--     # ------------------------------------------------------- -->
<!--   # Combine -->
<!--   # ------------------------------------------------------- -->

<!--   p_final <- wrap_plots( -->
<!--     A = pp[[marker_names[1]]], -->
<!--     B = pp[[marker_names[2]]], -->
<!--     C = pp[[marker_names[3]]], -->
<!--     D = pp[[marker_names[4]]], -->
<!--     E = pp[[marker_names[5]]], -->
<!--     F = pp[[marker_names[6]]], -->
<!--     G = p_legend, -->
<!--     design = "ABCG\nDEFG", -->
<!--     widths = c(1, 1, 1, 0.4) -->
<!--   ) -->

<!--   # print(p_final) -->
<!--   # Save result -->
<!--   outfig <- file.path(Figures, paste("clusters_samples_newnames_vanHaver_markers_", ct, -->
<!--                                      "_res_", selres, "_day_", day,  "_nHVG", nHVG_str, ".pdf", sep="")) -->
<!--   ggsave(outfig, plot=p_final, width=15, height=10) -->

<!-- } -->

<!-- ``` -->





<!-- # Marker Gene DotPlots -->

<!-- Visualizing top markers and Van-Haver markers on dotplots. The name of the top markers is legible on the saved plot, since that is wider. -->
<!-- ```{r EpiC-Marker-DotPlots, fig.width=50, fig.height=6} -->

<!-- # --- 3. Van-Haver Markers DotPlot (Color) --- -->
<!-- gene_all <- c() -->
<!-- marker_list_flat <- marker_list -->
<!-- for (mname in names(marker_list_flat)){ -->
<!--   marker_list_flat[[mname]] <- setdiff(marker_list_flat[[mname]], gene_all) -->
<!--   gene_all <- union(gene_all, marker_list_flat[[mname]]) -->
<!-- } -->

<!-- Idents(seu.integrated) <- "cluster_type" -->

<!-- p_dot_vh <- DotPlot( -->
<!--       seu.integrated, -->
<!--       features = marker_list_flat,  -->
<!--       cols = c("lightgrey", "red"), -->
<!--       col.min = 0, -->
<!--       dot.scale = 6, -->
<!--       assay = "SCT" -->
<!--     ) + -->
<!--     theme_bw() + -->
<!--     RotatedAxis() +  -->
<!--     theme( -->
<!--       axis.text.x = element_text(size = 12), -->
<!--       axis.title.x = element_blank(),  -->
<!--       axis.title.y = element_blank(),  -->
<!--       strip.clip = "off" -->
<!--     )  -->
<!-- p_dot_vh$data <- p_dot_vh$data[p_dot_vh$data$pct.exp > 0, ] # remove zero percent dots! -->

<!-- # Display and save Van-Haver color version -->
<!-- plot(p_dot_vh)   -->
<!-- ggsave(file.path(Figures, paste("VanHaver_Markers_DotPlot_newnames_", ct, "_selres_", selres, -->
<!--                                 "_all_nHVG", nHVG_str, ".pdf", sep="")),  -->
<!--        p_dot_vh, width = 16, height = 6) -->

<!-- ``` -->




# Marker Gene DotPlots for different days
Visualizing top markers and Van-Haver markers on dotplots. The name of the top markers is legible on the saved plot, since that is wider. A grayscale figure is also generated and saved.

```{r EpiC-Marker-DotPlots for days, fig.width=50, fig.height=6}

gene_all <- c()
marker_list_flat <- marker_list
for (mname in names(marker_list_flat)){
  marker_list_flat[[mname]] <- setdiff(marker_list_flat[[mname]], gene_all)
  gene_all <- union(gene_all, marker_list_flat[[mname]])
}


day_options <- c("D26_WT", "D26_STOP_MYCN", "D42_novirus", "D42_hTH")
for (day in day_options){
      seu.day <- subset_seu_by_day(seu.integrated, day)
      
  # # drop unused levels - do it everywhere!!!
  #   cluster_sizes <- table(seu.day$cluster_type)
  #   keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
  #   seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
  #   if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
  #     seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
  #   }
  #   seu.day$cluster_type <- droplevels(seu.day$cluster_type)
    
    
        # drop unused levels - do it everywhere!!!
      cluster_sizes <- table(seu.day$cluster_type)
      keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
      seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
      if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
        seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
      }
      seu.day$cluster_type <- droplevels(seu.day$cluster_type)
      
      seu.day$num_id <- seu.day[[selres]]
      seu.day$num_id <- droplevels(seu.day$num_id)
      
      # add num_id order; TO DO later: num_id_final - take into account that some clusters appear in earlier days first
      actual_numbers <- as.numeric(as.character(unique(seu.day$num_id)))
      sorted_numeric_levels <- as.character(sort(actual_numbers))
      seu.day$num_id <- factor(seu.day$num_id, levels = sorted_numeric_levels)
        
      # print(day)
      # print(table(seu.day$sample.names))
          
      # use new names
      actual_day <- limma::strsplit2(day, "_")[[1]]
      
      # Rename clusters from previous annotation
      if (actual_day %in% names(cluster_names_dict[[ct]])){
          new_cluster_ids <- dplyr::recode(
          as.character(seu.day$cluster_type),
          !!!cluster_names_dict[[ct]][[actual_day]]
        )
        # Preserve cluster order
        renamed_cluster_order <- unique(dplyr::recode(
          bio_order,
          !!!cluster_names_dict[[ct]][[actual_day]]
        ))
        
        new_cluster_ids <- factor(
          new_cluster_ids,
          levels = unique(renamed_cluster_order)
        )
        
        # Store back into Seurat object and construct final legend text
        seu.day$cluster_type_renamed <- new_cluster_ids
        Idents(seu.day) <- "cluster_type_renamed"
        seu.day$legend_label <- paste0(seu.day$num_id, ": ", seu.day$cluster_type_renamed)
      } else {
          new_cluster_ids <- seu.day$cluster_type
          renamed_cluster_order <- bio_order
          # Store back into Seurat object and construct final legend text
          seu.day$cluster_type_renamed <- new_cluster_ids
          Idents(seu.day) <- "cluster_type_renamed"
          seu.day$legend_label <- paste0(seu.day$num_id, ": ", seu.day$cluster_type_renamed)
      }
  

  # -------------------------------------------------------
  # FeaturePlots
  # -------------------------------------------------------

  #Idents(seu.day) <- paste("cluster_type_renamed", day, sep="_")
  Idents(seu.day) <- "cluster_type_renamed"
  
  # --- 3. Van-Haver Markers DotPlot (Color) ---
  p_dot_vh <- DotPlot(
        seu.day,
        features = marker_list_flat, 
        cols = c("lightgrey", "red"),
        #col.min = 0,
        dot.scale = 6,
        assay = "SCT"
        # assay = "RNA"
      ) +
      theme_bw() +
      scale_y_discrete(labels = function(x) abbrev_dict[x]) +
      RotatedAxis() + 
      theme(
        axis.text.x = element_text(size = 12, angle = 45, hjust = 1, vjust = 1),
        #axis.text.y = element_text(size = 14, angle = -45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 16),
        strip.text.x = element_text(size = 16, angle = 30, hjust = 0.5),
        #axis.text.x = element_text(size = 12),
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        strip.clip = "off"
      ) 

  p_dot_vh$data <- p_dot_vh$data[p_dot_vh$data$pct.exp > 0, ] # remove zero percent dots!

  # # Display and save Van-Haver color version
  # plot(p_dot_vh)  
  # ggsave(file.path(Figures, paste("VanHaver_Markers_DotPlot_newnames_", ct, "_selres_", selres,
  #                                 "_day_", day, "_nHVG", nHVG_str, ".pdf", sep="")), 
  #        p_dot_vh, width = 16, height = 6)
  
  # # problem: different x tick label -> different actual dotplot sizes
  g <- ggplotGrob(p_dot_vh)
  g$layout[grepl("axis-l", g$layout$name), c("name", "t", "l", "b", "r")] # run once to find name
  axis_l_row <- which(g$layout$name == "axis-l-1")
  axis_l_col <- g$layout$l[axis_l_row]
  
  g$widths[axis_l_col] <- unit(2, "in")   # <- SAME fixed value every iteration
  
  ggsave(file.path(Figures, paste("VanHaver_Markers_DotPlot_newnames_", ct, "_selres_", selres,
                                  "_day_", day, "_nHVG", nHVG_str, ".pdf", sep="")), 
         g, width = 16, height = 6)
  
    }

```


# HOX and Structural Gene Sets for different days
HOX and structural gene sets for different days  - but it takes a long time to plot the dimplots
change toraster to TRUE for faster but not so nice figure generation
```{r HOX and structural gene sets for different days}
# ==============================================================================
# STEP 1: PRE-COMPUTE ALL GENE STRUCTURES (Run ONCE outside the loop!)
# ==============================================================================
toraster <- FALSE

existing_genes <- rownames(seu.integrated[["SCT"]])

hox_list_flat       <- clean_and_flatten_list(hox_list, existing_genes)
structure_list_flat <- clean_and_flatten_list(structure_list, existing_genes)
dimplot_list_flat   <- clean_and_flatten_list(dimplot_list, existing_genes)
CD44_flat <- clean_and_flatten_list(CD44_dimplots, existing_genes)
CD44_mod <- clean_and_flatten_list(CD44_dimplots_MES, existing_genes)

# ==============================================================================
# STEP 2: THE HIGH-SPEED LOOP
# ==============================================================================
# day_options <- c("D26_both", "D42_both", "D26_WT", "D26_STOP_MYCN",
#                   "D42_novirus", "D42_hTH", "all")
# day_options <- c("D26_both", "D42_both", "D26_WT", "D26_STOP_MYCN",
#                   "D42_novirus", "D42_hTH")
# # temporary
day_options <- c("D42_hTH")
for (day in day_options){
  message("Processing day condition: ", day)
  
  # Subset and clean identity slots
  seu.day <- subset_seu_by_day(seu.integrated, day)
  cluster_sizes <- table(seu.day$cluster_type)
  keep_clusters <- names(cluster_sizes[cluster_sizes >= drop_clusters])
  seu.day <- subset(seu.day, subset = cluster_type %in% keep_clusters)
  if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
      seu.day <- subset(seu.day, subset = !cluster_type %in% drop_in_hTH)
  }
  seu.day$cluster_type <- droplevels(seu.day$cluster_type)
  
  Idents(seu.day) <- paste("cluster_type_renamed", day, sep="_")
  
  # ----------------------------------------------------------------------------
  # FUNCTION TO PROCESS A GENE LIST IN A SINGLE PASS
  # ----------------------------------------------------------------------------
  process_gene_plots <- function(seu.in, flat_gene_list, prefix_name, 
                                 dot_height, dot_width, grid_height, is_main_marker = FALSE) {
    features_v <- unlist(flat_gene_list)
    if(length(features_v) == 0) return(NULL)
    #Idents(seu.in) <- paste("cluster_type_renamed", day, sep="_")
    
    # 1. DotPlot Generation
    p_dot <- DotPlot(seu.in, features = flat_gene_list, cols = c("lightgrey", "red"),
                     col.min = 0, dot.scale = 6, assay = "SCT") +
      theme_bw() + RotatedAxis() + scale_y_discrete(labels = function(x) abbrev_dict[x]) +
      theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1, vjust = 1), 
            axis.text.y = element_text(size = 16), 
            strip.text.x = element_text(size = 16),
            axis.title.x = element_blank(), axis.title.y = element_blank(), strip.clip = "off")
    # Apply extra formatting block if it's the main marker block
    if(is_main_marker) {
      p_dot <- p_dot + theme(strip.text.x = element_text(size = 12, face = "bold", angle = 90, hjust = 0, vjust = 0.5, lineheight = 0.8))
    }
    p_dot$data <- p_dot$data[p_dot$data$pct.exp > 0, ] # remove zero percent dots!
    
    ggsave(file.path(Figures, paste0(prefix_name, "_Markers_DotPlot_newnames_", ct, "_selres_", selres, "_day_", day, "_nHVG", nHVG_str, ".pdf")), 
           p_dot, width = dot_width, height = dot_height)
    
    # 2. FeaturePlot Generation - RUN EXACTLY ONCE 
    p_list <- FeaturePlot(seu.in, features = features_v, combine = FALSE, raster = toraster)
    names(p_list) <- features_v
    
    # Apply the standard custom coloring to our base list
    p_list_colored <- lapply(p_list, function(p) {
      p + scale_colour_gradient(low = "lightgrey", high = "purple", limits = c(min_val, max_val), oob = scales::squish)
    })
    
    # 3. Save Wrapped Master Matrix Plot
    p_dim_wrapped <- patchwork::wrap_plots(p_list_colored)
    ggsave(file.path(Figures, paste0(prefix_name, "_Markers_DimPlot_newnames_", ct, "_selres_", selres, "_day_", day, "_nHVG", nHVG_str, ".pdf")),
           p_dim_wrapped, width = if(is_main_marker) 32 else 16, height = grid_height)
    
    # 4. Apply Canvas Axes to individual copies and save separately
    for (pname in names(p_list_colored)) {
      pf_canvas <- add_corner_axes(p_list_colored[[pname]])
      ggsave(file.path(Figures_dimplots, paste0(prefix_name, "_Markers_DimPlot_newnames_", ct, "_selres_", selres, "_day_", day, "_nHVG", nHVG_str, "_", pname, ".pdf")), 
             pf_canvas, width = 4, height = 4)
    }
  }
  
  # ----------------------------------------------------------------------------
  # EXECUTE BACK-TO-BACK PROCESSSING
  # ----------------------------------------------------------------------------
  process_gene_plots(seu.day, hox_list_flat, "HOX", dot_height = 6, dot_width=16, grid_height = 10)
  process_gene_plots(seu.day, structure_list_flat, "Structure", dot_height = 6, dot_width=16, grid_height = 12)
  process_gene_plots(seu.day, dimplot_list_flat, "Main", dot_height = 6, dot_width=16, grid_height = 18, is_main_marker = TRUE)
  process_gene_plots(seu.day, CD44_flat, "CD44", dot_height = 10, dot_width=10, grid_height = 10) # note: adjust name handling if needed
  if (ct=="kolf2" & (day=="D42_hTH" | day=="D42_both")){
    seu.subset <- subset(seu.day, subset = cluster_type %in% c("type_EGFP_MES", "type_EGFP_tumor"))
    seu.subset$cluster_type <- droplevels(seu.subset$cluster_type)
    process_gene_plots(seu.subset, CD44_mod, "CD44_mod", dot_height = 3, dot_width=10, grid_height = 10) 
  }
}
```

